---
title: "SIG13 GLM Validation "
author: "Ian Zumpano"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
editor_options:
  markdown:
    wrap: 72
---

# SIG07 vs SIG13 GLM Validation

Goal: to get pearson correlation of GLM LFC terms between SIG07 and SIG13

## Load packages:
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
# Repo-relative paths: inputs are read from imports_stable/, outputs are written to analysis_outs/
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find the repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "02_combinatorial_screen_signalseq_SIG13")
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
dir.create(file.path(outs_dir, "plots/qc"), recursive = TRUE, showWarnings = FALSE)
theme_set(theme_Publication())
```


## Interaction LFC Correlation

```{r}
sig13_data <- read.csv(file.path(imports_dir, "SIG13/analysis_outs/glmGamPoi/interactions_scored_v3_glmGamPoi_0.05filter_sig.csv"))
sig07_data <- read.csv(file.path(imports_dir, "SIG07/analysis_outs/glmGamPoi/glmGamPoi_interaction/glmGamPoi_interaction_lfc_0.05filter.csv"))
```


## Counts of significant interactions per group

```{r}
sig07_data_interact <- sig07_data %>% filter(!is.na(lfc))
sig13_data_interact <- sig13_data %>% filter(!is.na(lfc_interaction))
```

```{r}
sig13_data_interact <- sig13_data_interact %>% mutate(
    parts = str_split(interaction, "_"),
    # alphabetically sort each pair
    interaction_sorted = sapply(parts, function(x) paste(sort(x), collapse = "_"))
  )

sig07_data_interact <- sig07_data_interact %>% mutate(
    parts = str_split(interaction, "_"),
    # alphabetically sort each pair
    interaction_sorted = sapply(parts, function(x) paste(sort(x), collapse = "_"))
  )

sig07_data_interact <- sig07_data_interact %>% mutate(lfc_interaction = lfc)
```

```{r}
nsig07 <- sig07_data_interact %>% group_by(interaction_sorted) %>% summarize(nsig = sum(adj_pval < 0.05, na.rm = TRUE)) %>% arrange(nsig) %>% as_data_frame()

nsig13 <- sig13_data_interact %>% group_by(interaction_sorted) %>% summarize(nsig = sum(adj_pval_interaction < 0.05, na.rm = TRUE)) 

nsig_interactions <- inner_join(nsig13, nsig07, by = 'interaction_sorted', suffix = c(".sig13", ".sig07"))
nsig_interactions
```
Based on this, threshold of 100 significant interactions per group in both SIG07 and SIG13 seems reasonable

```{r}
interactions_to_keep = nsig_interactions %>% filter(nsig.sig13 >= 100 & nsig.sig07 >= 100) %>% pull(interaction_sorted)
```

### Take mean of duplicate interaction terms

```{r}
sig13_to_corr <- sig13_data_interact %>%
  group_by(interaction_sorted, name) %>%
  summarise(mean_interaction_lfc = mean(lfc_interaction, na.rm = TRUE))

sig07_to_corr <- sig07_data_interact %>%
  group_by(interaction_sorted, name) %>%
  summarise(mean_interaction_lfc = mean(lfc_interaction, na.rm = TRUE))

sig13_to_corr <- as_data_frame(sig13_to_corr)
sig07_to_corr <- as_data_frame(sig07_to_corr)
```

```{r}
joint_data_interact <- inner_join(sig13_to_corr, sig07_to_corr, by = c("interaction_sorted", "name"), suffix = c(".sig13", ".sig07")) 
```


```{r}
joint_data_interact_filtered <- joint_data_interact %>% filter(interaction_sorted %in% interactions_to_keep)
```


### Get correlation

```{r}
correlation_results <- joint_data_interact_filtered %>%
  group_by(interaction_sorted) %>%
  summarise(
    pearson_corr = cor(mean_interaction_lfc.sig13, mean_interaction_lfc.sig07, method = "pearson")
  )

correlation_results %>% arrange(desc(pearson_corr))
```

### scatterplots

```{r}
library(ggplot2)

p <- ggplot(joint_data_interact_filtered, aes(x = mean_interaction_lfc.sig13, y = mean_interaction_lfc.sig07)) +
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "lm", color = "red", se = FALSE) + 
  
  # Set ncol = 3 to force 3 columns
  facet_wrap(~interaction_sorted, scales = "free", ncol = 3) +
  
  theme_bw()

# WHEN SAVING:
# 1. 'limitsize = FALSE' lets you save very tall images.
# 2. Increase 'width' to stretch the plots horizontally.
# 3. Increase 'height' to accommodate all the rows (since 3 cols = many rows).
ggsave(file.path(outs_dir, "plots/qc/interaction_scatterplots.pdf"), plot = p, width = 15, height = 40, limitsize = FALSE)
```

### Plot interaction correlation for combinations with >= 100 significant interactions

```{r}
ggplot(correlation_results, aes(x = pearson_corr)) +
  geom_histogram(binwidth = 0.05) +
  labs(
    title = "Distribution of Pearson Correlations", 
    subtitle = "SIG07 vs SIG13 Interaction Term LFC",
    x = "Pearson Correlation",
    y = "Count"
  ) +
  theme(aspect.ratio = 1) +
  xlim(0,1) 
```

## Single-Term LFC Correlation

### Wrangle data

```{r}
sig07_data_singles <- read_csv(file.path(imports_dir, "SIG07/analysis_outs/glmGamPoi/glmGamPoi_interaction/glmGamPoi_singles_lfc_0.05filter.csv"))
sig13_data_singles <- read.csv(file.path(imports_dir, "SIG13/analysis_outs/glmGamPoi/glmGamPoi_singleTerm_lfc_0.05filter.csv"))
```

```{r}
sig13_data_singles_filtered <- sig13_data_singles %>%
  filter(condition %in% sig07_data_singles$interaction) %>%
  rename(interaction = condition)
joint_singles <- inner_join(sig07_data_singles, sig13_data_singles_filtered, by = c("name", "interaction"), suffix = c(".sig07", ".sig13"))
```

### Get correlation grouped by interaction and ligand 

```{r}
ligand_correlations_singles <- joint_singles %>%
  group_by(interaction) %>% 
  summarise(
    pearson_corr = cor(lfc.sig13, lfc.sig07, method = "pearson"),
    # It is often helpful to count how many genes are driving this correlation
    n_genes = n(), 
    .groups = "drop"
  ) %>%
  arrange(desc(pearson_corr))

# View the results
print(ligand_correlations_singles)
```

```{r}
ggplot(ligand_correlations_singles, aes(x = pearson_corr)) +
  geom_histogram(binwidth = 0.05) +
  labs(
    title = "Distribution of Pearson Correlations", 
    subtitle = "SIG07 vs SIG13 Single Term LFC",
    x = "Pearson Correlation",
    y = "Count"
  ) +
  theme(aspect.ratio = 1) +
  xlim(0,1) 
ggsave(file.path(outs_dir, "plots/qc/SIG07_SIG13_pearson_corr_single_term.pdf"))
```

