---
title: "SIG03 oBC Direct Amp Barcode Analysis (CD4)"
author: "Eric Y. Wang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

Circular barcode analysis on the HPC scanpy outputs
(`SIG03_oBCDirect_full.h5mu`, the same object loaded in `03_linear_vs_circular_counts_SIG02_SIG03.ipynb`).
Only CD4 T cells are analyzed; the splenocyte and Jurkat sections are dropped.

```{r setup, include=FALSE}
library(tidyverse)
library(Matrix)
library(rhdf5)
library(ggplot2)
library(cowplot)
library(yardstick)
library(ggpointdensity)
library(RColorBrewer)
library(patchwork)
knitr::opts_chunk$set(echo = TRUE)

# Repo-relative paths: inputs from imports_stable/, outputs to analysis_outs/
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "01_barcode_comparison_signalseq_SIG02_SIG03")
```

```{r}
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
source(file.path(repo_dir, "functions/r_custom/scRNA_seq_analysis_functions.R"))
theme_set(theme_Publication())
```

### [Import Data]{.underline}

The `.h5mu` is read directly with `rhdf5` (reticulate + muon fails on the cluster
because of an OpenSSL library conflict between R and the `scanpy_standard2` env).

```{r}
OUT_DIR <- file.path(imports_dir, "SIG03/scanpy_outs")
FULL_H5MU <- file.path(OUT_DIR, "SIG03_oBCDirect_full.h5mu")

ANALYSIS_OUTS <- file.path(outs_dir, "01_barcode_analysis_oBCDirect")
dir.create(ANALYSIS_OUTS, recursive = TRUE, showWarnings = FALSE)

# read an AnnData obs column (categoricals are stored as 0-based codes + categories)
read_obs_col <- function(file, mod, col) {
  path <- paste0("/mod/", mod, "/obs/", col)
  if (h5readAttributes(file, path)[["encoding-type"]] == "categorical") {
    codes <- as.integer(h5read(file, paste0(path, "/codes")))
    cats <- as.character(h5read(file, paste0(path, "/categories")))
    codes[codes < 0] <- NA  # -1 = missing; must be NA (not 0) so indexing stays aligned
    return(cats[codes + 1])
  }
  as.vector(h5read(file, path))
}

# read a csr_matrix (cells x features) as a features x cells dgCMatrix (Seurat orientation)
read_csr <- function(file, path, features, cells) {
  sparseMatrix(
    i = as.integer(h5read(file, paste0(path, "/indices"))),
    p = as.integer(h5read(file, paste0(path, "/indptr"))),
    x = as.numeric(h5read(file, paste0(path, "/data"))),
    dims = c(length(features), length(cells)),
    dimnames = list(features, cells),
    index1 = FALSE
  )
}

cells <- as.character(h5read(FULL_H5MU, "/mod/bc/obs/_index"))
bc_features <- as.character(h5read(FULL_H5MU, "/mod/bc/var/_index"))
bcCounts <- read_csr(FULL_H5MU, "/mod/bc/layers/counts", bc_features, cells)

stopifnot(identical(cells, as.character(h5read(FULL_H5MU, "/mod/rna/obs/_index"))))
metadataAll <- tibble(
  cell_BC = cells,
  celltype = read_obs_col(FULL_H5MU, "rna", "celltype"),
  timepoint = read_obs_col(FULL_H5MU, "rna", "timepoint"),
  titer = read_obs_col(FULL_H5MU, "rna", "titer"),
  nCount_RNA = read_obs_col(FULL_H5MU, "rna", "total_counts"),
  nCount_BC = colSums(bcCounts)
)
h5closeAll()
```

Subset to CD4 cells and rebuild the legacy `hash.ID` labels (`CD4-{timepoint}-{titer}`).

```{r}
TITER_ORDER <- c("2e11", "6e10", "2e10", "6e9", "0")
TIMEPOINTS <- c("6h", "22h")
HASH_ORDER <- as.vector(outer(TITER_ORDER, TIMEPOINTS, function(ti, tp) paste0("CD4-", tp, "-", ti)))

metadata <- metadataAll %>%
  filter(celltype == "CD4", timepoint %in% TIMEPOINTS, titer %in% TITER_ORDER) %>%
  mutate(hash.ID = factor(paste0("CD4-", timepoint, "-", titer), HASH_ORDER))

countsCD4 <- bcCounts[, metadata$cell_BC]

metadata %>% count(hash.ID)
```

### [Visualize Barcode Distributions]{.underline}

#### CD4 Barcodes

```{r}
countData <- tibble(log1p_total_count = log1p(metadata$nCount_RNA),
                    log1p_oBC_count = log1p(metadata$nCount_BC))

countData %>%
  ggplot(aes(x = log1p_oBC_count, y = log1p_total_count)) +
    geom_pointdensity() +
    scale_color_viridis_c() +
    theme(aspect.ratio = 1)
```

```{r}
BCdata <- countsCD4 %>%
  as.matrix() %>%
  as_tibble(rownames = "BC") %>%
  pivot_longer(-BC, names_to = "cell_BC", values_to = "counts") %>%
  mutate(log10counts = log10(counts + 1)) %>%
  left_join(metadata, by = "cell_BC")

# overlaid per-dose histograms for one barcode at one timepoint
plot_bc_overlay <- function(bc, tp) {
  hashes <- paste0("CD4-", tp, "-", TITER_ORDER)
  BCdata %>%
    filter(BC == bc, hash.ID %in% hashes) %>%
    ggplot(aes(x = log10counts, fill = hash.ID)) +
    geom_histogram(aes(y = after_stat(density), group = hash.ID),
                   bins = 100, alpha = 0.75, position = "identity") +
    geom_density(aes(y = after_stat(density), group = hash.ID),
                 alpha = 0.25, color = "black", linewidth = 1) +
    scale_fill_manual(values = setNames(c(brewer.pal(8, "Dark2")[1:4], "grey50"), hashes)) +
    theme(legend.position = "none", aspect.ratio = 0.5) +
    ggtitle(paste(bc, tp))
}
```

```{r, fig.height=10, fig.width=15}
plot_bc_overlay("p139-BC4", "6h") + plot_bc_overlay("p139-BC5", "6h") +
  plot_bc_overlay("p139-BC4", "22h") + plot_bc_overlay("p139-BC5", "22h") +
  plot_layout(ncol = 2)
ggsave(file.path(ANALYSIS_OUTS, "barcode_histogram_CD4_overlay.pdf"))
```

```{r}
# get mean counts for barcodes
BCdata %>%
  filter(BC %in% c("p139-BC4")) %>%
  filter(hash.ID %in% c("CD4-6h-2e11","CD4-6h-6e10","CD4-6h-2e10","CD4-6h-6e9","CD4-6h-0")) %>%
  group_by(hash.ID) %>%
  summarise(mean = mean(counts))
```

```{r}
expressionData <- tibble(cell_bc = metadata$cell_BC,
                         hash.ID = metadata$hash.ID,
                         BC4 = countsCD4["p139-BC4",],
                         BC5 = countsCD4["p139-BC5",])

# BC4 vs BC5 density scatter for one hash.ID
plot_bc4_bc5 <- function(hash, title) {
  expressionData %>%
    filter(hash.ID %in% hash) %>%
    ggplot(aes(x = log(BC4 + 1), y = log(BC5 + 1))) +
      geom_pointdensity() +
      scale_color_viridis_c() +
      theme(aspect.ratio = 1) +
      xlim(0,10) +
      ylim(0,10) +
      ggtitle(title)
}
```

```{r, fig.width=10, fig.height=10}
plot_grid(plot_bc4_bc5("CD4-6h-2e11", "2e11 LP/mL 6h"),
          plot_bc4_bc5("CD4-6h-6e10", "6e10 LP/mL 6h"),
          plot_bc4_bc5("CD4-6h-2e10", "2e10 LP/mL 6h"),
          plot_bc4_bc5("CD4-6h-6e9",  "6e9 LP/mL 6h"),
          ncol = 2)
ggsave(file.path(ANALYSIS_OUTS, "bc4_bc5_6h_density_plot.pdf"))
```

```{r, fig.width=10, fig.height=10}
plot_grid(plot_bc4_bc5("CD4-22h-2e11", "2e11 LP/mL 22h"),
          plot_bc4_bc5("CD4-22h-6e10", "6e10 LP/mL 22h"),
          plot_bc4_bc5("CD4-22h-2e10", "2e10 LP/mL 22h"),
          plot_bc4_bc5("CD4-22h-6e9",  "6e9 LP/mL 22h"),
          ncol = 2)
ggsave(file.path(ANALYSIS_OUTS, "bc4_bc5_22h_density_plot.pdf"))
```

Get frequency of assignments at a cutoff of ln(BC)
```{r}
expressionData %>%
  mutate(pos = ifelse(log(BC5+1) > 4 & log(BC4+1) > 4,T,F)) %>%
  group_by(hash.ID) %>%
  summarise(freq = sum(pos)/n())
```

### [Barcode Classification]{.underline}

For each barcode x timepoint, each transduced dose is scored against the matched
`0`-titer control using raw barcode UMIs as the predictor.

```{r}
# one test dataset per dose: that dose (positive) vs the matched 0 control (negative)
make_test_data <- function(bc, tp) {
  doses <- paste0("CD4-", tp, "-", setdiff(TITER_ORDER, "0"))
  neg <- paste0("CD4-", tp, "-0")
  testDataList <- map(setNames(doses, doses), function(dose) {
    tibble(hash.ID = metadata$hash.ID,
           umi_counts = countsCD4[bc,]) %>%
      filter(hash.ID %in% c(dose, neg)) %>%
      mutate(label = case_when(hash.ID == neg ~ "negative",
                               hash.ID %in% doses ~ "positive")) %>%
      mutate(label = factor(label, c("positive","negative")))
  })
  testDataList
}

# AUC table (written to csv) + ROC and PR curves
run_classification <- function(bc, tp) {
  testDataList <- make_test_data(bc, tp)
  bc_short <- str_remove(bc, "p139-")

  results <- imap_dfr(testDataList, function(data, name) {
    data %>%
      roc_auc(truth = label, umi_counts) %>%
      mutate(group = name)
  })
  write_csv(results, file.path(ANALYSIS_OUTS, paste0("umi_cutoff_auc_metrics_", bc_short, "_", tp, ".csv")))

  curve_data <- imap_dfr(testDataList, function(data, name) {
    roc <- roc_curve(data, truth = label, umi_counts) %>%
      mutate(type = "ROC")
    pr <- pr_curve(data, truth = label, umi_counts) %>%
      mutate(type = "PR")
    bind_rows(roc, pr) %>%
      mutate(group = name)
  }) %>%
    mutate(group = factor(group, names(testDataList)))

  roc <- curve_data %>%
    filter(type == "ROC") %>%
    ggplot(aes(x = 1 - specificity, y = sensitivity, color = group)) +
    geom_path(linewidth = 1) +
    scale_color_brewer(palette = "Dark2") +
    labs(title = paste("ROC", tp, bc_short), x = "False Positive Rate", y = "True Positive Rate") +
    theme(aspect.ratio = 1) +
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1))

  pr <- curve_data %>%
    filter(type == "PR") %>%
    ggplot(aes(x = recall, y = precision, color = group)) +
    geom_path(linewidth = 1) +
    scale_color_brewer(palette = "Dark2") +
    labs(title = paste("PR Curve", tp, bc_short), x = "Recall", y = "Precision") +
    theme(aspect.ratio = 1) +
    coord_cartesian(xlim = c(0.5, 1), ylim = c(0.5, 1))

  list(results = results, roc = roc, pr = pr)
}
```

#### Barcode 4 Recall

```{r}
bc4_6h <- run_classification("p139-BC4", "6h")
bc4_22h <- run_classification("p139-BC4", "22h")
bind_rows(bc4_6h$results, bc4_22h$results)
```

#### Barcode 5 Recall

```{r}
bc5_6h <- run_classification("p139-BC5", "6h")
bc5_22h <- run_classification("p139-BC5", "22h")
bind_rows(bc5_6h$results, bc5_22h$results)
```

#### Visualize

```{r, fig.height=12, fig.width=12}
plot_grid(bc4_6h$pr, bc4_22h$pr, bc5_6h$pr, bc5_22h$pr)
ggsave(file.path(ANALYSIS_OUTS, "pr_curves.pdf"))
plot_grid(bc4_6h$roc, bc4_22h$roc, bc5_6h$roc, bc5_22h$roc)
ggsave(file.path(ANALYSIS_OUTS, "roc_curves.pdf"))
```

```{r}
metadata %>%
  group_by(hash.ID) %>%
  summarise(count = n())
```

QC: library size distribution per sample.

```{r, fig.height=5, fig.width=12}
metadata %>%
  ggplot(aes(color=hash.ID, x=nCount_RNA, fill=hash.ID)) +
    geom_density() +
    scale_x_log10() +
    facet_wrap(~hash.ID, ncol = 5)
```
