---
title: "Th receptor cross-regulation"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
---

```{r setup}
library(tidyverse)
library(pheatmap)
library(RColorBrewer)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Repo-relative paths: inputs from imports_stable/, outputs to analysis_outs/<module>/
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "07_crispr_arrayed_validation_RNAseq_SIG30")
dir.create(file.path(outs_dir, "program_crossreg"), recursive = TRUE, showWarnings = FALSE)
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
theme_set(theme_Publication())
```

# Import

```{r}
deg_SIG14 <- read_csv(file.path(imports_dir, "SIG14/analysis_outs/deg_updated/res_conditions_SIG14.csv"))
deg_SIG18 <- read_csv(file.path(imports_dir, "SIG18/analysis_outs/deg/res_conditions_SIG18.csv"))
deg_SIG30_targets <- read_csv(file.path(imports_dir, "SIG30/analysis_outs/res_targets_SIG30.csv"))
int_SIG18 <- read_csv(file.path(imports_dir, "SIG18/analysis_outs/deg/res_interaction_scored_SIG18.csv"))
deg_SIG29 <- read_csv(file.path(imports_dir, "SIG29/analysis_outs/res_conditions_SIG29.csv"))
int_SIG29 <- read_csv(file.path(imports_dir, "SIG29/analysis_outs/res_interaction_scored_SIG29.csv"))
```

```{r}
deg_SIG18 <- deg_SIG18 %>%
  mutate(condition = case_match(condition,
                                 "Th2"  ~ "IL4",
                                 "Th1" ~ "IFNg",
                                 "TGFb" ~ "TGFb",
                                 "TNF" ~ "TNF",
                                 "Th2-TNF" ~ "IL4-TNF",
                                 "Th1-TNF" ~ "IFNg-TNF",
                                 "TGFb-TNF" ~ "TGFb-TNF",
                                 "iTreg" ~ "TGFb-IL2",
                                 "iTreg-TNF" ~ "TGFb-TNF-IL2",
                                  .default = condition)) %>%
  mutate(timepoint = "96hours")

deg_SIG14 <- deg_SIG14  %>%
  mutate(timepoint = "6hours")

# Single-ligand conditions only; combination conditions (e.g. TGFb-TNF) are
# handled separately below alongside the SIG30 CRISPR perturbations.
deg_joint <- bind_rows(deg_SIG18, deg_SIG14) %>%
  filter(condition %in% c("IL4","IFNg","TGFb","TNF"))
```

SIG29 is the ligand experiment run in the same sequencing batch as SIG30
(TGFb x {TNF, TL1A, OX40L, DTA1}). Its conditions are coded `ligand1_ligand2`
with `none` placeholders; recode them to the `TGFb-TNF` style used above.

```{r}
clean_SIG29_condition <- function(x) {
  x %>%
    str_remove("^none_") %>%
    str_remove("_none$") %>%
    str_replace("_", "-")
}

deg_SIG29 <- deg_SIG29 %>%
  mutate(condition = clean_SIG29_condition(condition))

int_SIG29 <- int_SIG29 %>%
  mutate(condition = clean_SIG29_condition(condition))
```

# Helper functions

Shared helpers for pivoting long DEG tables into gene x condition LFC/padj
matrices and rendering them as annotated heatmaps.

```{r}
# --- Pivot a long (gene, condition, log2FoldChange, padj) df into matrices ---
build_lfc_padj_mats <- function(df, genes, conditions) {
  df <- df %>% filter(gene %in% genes, condition %in% conditions)

  pivot_mat <- function(value_col) {
    mat <- df %>%
      pivot_wider(id_cols = "gene", names_from = "condition",
                  values_from = all_of(value_col), values_fill = NaN) %>%
      right_join(tibble(gene = genes), by = "gene") %>%
      column_to_rownames("gene") %>%
      as.matrix()

    for (cn in setdiff(conditions, colnames(mat))) mat[, cn] <- NA
    mat[genes, conditions, drop = FALSE]
  }

  list(lfc = pivot_mat("log2FoldChange"), padj = pivot_mat("padj"))
}

# --- Convert a padj matrix into a matrix of significance stars ---
stars_from_padj <- function(p) {
  p_vec <- as.vector(p)
  stars_vec <- case_when(
    is.na(p_vec)  ~ "",
    p_vec < 0.001 ~ "***",
    p_vec < 0.01  ~ "**",
    p_vec < 0.05  ~ "*",
    TRUE          ~ ""
  )
  matrix(stars_vec, nrow = nrow(p), ncol = ncol(p), dimnames = dimnames(p))
}

lfc_color_scale <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100)
```

# Map Receptor Induction

For each receptor panel, keep genes significant (`padj < 0.1`) in at least
one ligand/timepoint, then plot LFC across all ligand conditions at both
6h and 96h (using the unfiltered LFC so nonsignificant values are retained,
with significance stars overlaid).

```{r, fig.height=3.5, fig.width=4}
ligands <- c("IFNg","IL4","TGFb","TNF","IL6","IL21","IL12","IL2")
receptors <- c("Ifngr1","Il12rb1","Il2ra","Il2rb","Il4ra","Il6ra","Il21r",
               "Tnfrsf1b","Tgfbr2","Tnfrsf25","Tnfrsf4","Tnfrsf18")

deg_df <- deg_SIG14 %>%
  filter(condition %in% ligands)

genes <- deg_df %>%
  filter(gene %in% receptors, padj < 0.1) %>%
  distinct(gene) %>%
  pull(gene) %>%
  sort()

conditions <- deg_df %>%
  filter(gene %in% genes) %>%
  distinct(condition) %>%
  pull(condition) %>%
  sort()

m_6h  <- build_lfc_padj_mats(filter(deg_df, timepoint == "6hours"), genes, conditions)

colnames(m_6h$lfc)  <- colnames(m_6h$padj)  <- paste0(conditions, "_6h")

mat   <- cbind(m_6h$lfc)
stars <- cbind(stars_from_padj(m_6h$padj))

col_anno <- data.frame(timepoint = rep(c("6hours"), each = length(conditions)),
                       row.names = colnames(mat))

hm <- pheatmap(
  mat,
  cluster_cols = T,
  cluster_rows = F,
  annotation_col = col_anno,
  labels_col = sub("_6h$|_96h$", "", colnames(mat)),
  color = lfc_color_scale,
  breaks = seq(-2, 2, length.out = 101),
  na_col = "grey90",
  display_numbers = stars,
  fontsize_number = 12,
  number_color = "black"
)

save_pheatmap_pdf(hm, file.path(outs_dir, "program_crossreg/crossreg_6h_all.pdf"), height = 3.5, width = 4)
```

```{r, fig.height=2.5, fig.width=4.5}
ligands <- c("IFNg","IL12","IL2","IL4","IL6","IL21","TNF","TGFb")
receptors <- c("Ifngr1","Il12rb1","Il2ra","Il2rb","Il4ra","Il6ra","Il21r",
               "Tnfrsf1b","Tgfbr2")

deg_df <- deg_SIG14 %>%
  filter(condition %in% ligands)

genes <- deg_df %>%
  filter(gene %in% receptors, padj < 0.1) %>%
  distinct(gene) %>%
  pull(gene) %>%
  sort()

conditions <- deg_df %>%
  filter(gene %in% genes) %>%
  distinct(condition) %>%
  pull(condition) %>%
  sort()

m_6h  <- build_lfc_padj_mats(filter(deg_df, timepoint == "6hours"), genes, conditions)

colnames(m_6h$lfc)  <- colnames(m_6h$padj)

mat   <- cbind(m_6h$lfc)
mat   <- mat[receptors ,ligands]
stars <- cbind(stars_from_padj(m_6h$padj))
stars <- stars[receptors ,ligands]

col_anno <- data.frame(timepoint = rep(c("6hours"), each = length(conditions)),
                       row.names = colnames(mat))

hm <- pheatmap(
  mat,
  cluster_cols = F,
  cluster_rows = F,
  annotation_col = col_anno,
  labels_col = sub("_6h$|_96h$", "", colnames(mat)),
  color = lfc_color_scale,
  breaks = seq(-2, 2, length.out = 101),
  na_col = "grey90",
  display_numbers = stars,
  fontsize_number = 12,
  angle_col = 90,
  number_color = "black"
)

save_pheatmap_pdf(hm, file.path(outs_dir, "program_crossreg/crossreg_6h_subset.pdf"), height = 2.5, width = 4.5)
```

```{r}
plot_receptor_heatmap_joint <- function(deg_df, receptors, sig_thresh = 0.1, lfc_limit = 3) {
  genes <- deg_df %>%
    filter(gene %in% receptors, padj < sig_thresh) %>%
    distinct(gene) %>%
    pull(gene) %>%
    sort()

  conditions <- deg_df %>%
    filter(gene %in% genes) %>%
    distinct(condition) %>%
    pull(condition) %>%
    sort()

  m_6h  <- build_lfc_padj_mats(filter(deg_df, timepoint == "6hours"),  genes, conditions)
  m_96h <- build_lfc_padj_mats(filter(deg_df, timepoint == "96hours"), genes, conditions)

  colnames(m_6h$lfc)  <- colnames(m_6h$padj)  <- paste0(conditions, "_6h")
  colnames(m_96h$lfc) <- colnames(m_96h$padj) <- paste0(conditions, "_96h")

  mat   <- cbind(m_6h$lfc, m_96h$lfc)
  stars <- cbind(stars_from_padj(m_6h$padj), stars_from_padj(m_96h$padj))

  col_anno <- data.frame(
    timepoint = rep(c("6hours", "96hours"), each = length(conditions)),
    row.names = colnames(mat)
  )

  pheatmap(mat,
           cluster_cols = FALSE,
           cluster_rows = FALSE,
           gaps_col = length(conditions),
           annotation_col = col_anno,
           labels_col = sub("_6h$|_96h$", "", colnames(mat)),
           color = lfc_color_scale,
           breaks = seq(-lfc_limit, lfc_limit, length.out = 101),
           na_col = "grey90",
           display_numbers = stars,
           fontsize_number = 12,
           number_color = "black")
}
```


```{r, fig.height=3, fig.width=4.5}
plot_receptor_heatmap_joint(deg_joint, c("Il12rb1","Il4ra","Il2ra","Il2rb","Il6ra","Ifngr1",
                                    "Tnfrsf1b","Tnfrsf25","Tgfbr2","Tnfrsf4","Tnfrsf18"))
```

```{r, fig.height=2.5, fig.width=4}
plot_receptor_heatmap_joint(deg_joint, c("Il12rb1","Il4ra","Il2ra","Il2rb","Il6ra","Tnfrsf1b","Tgfbr2"))
```

# Ligand induction vs. CRISPR perturbation

`Tgfbr2`, `Tnfrsf1b`, `Nr4a3`, and `Ptpn1` are both ligand-regulated
receptors/signaling genes (in the SIG29 ligand panel) and direct CRISPR
KO targets in this screen (SIG30). Ligand-induced changes are taken from
SIG29, the TGFb x costimulatory-ligand experiment run in the same sequencing
batch as SIG30. This compares each gene's TGFb x TNF *interaction* term
(deviation of the combo from the additive expectation, `lfc_interaction` /
`padj_interaction` from SIG29) against its LFC under CRISPR knockout of each
of the 9 SIG30 targets (vs. control), so cross-regulation between targets
(not just self-knockout) is visible.

```{r, fig.height=3, fig.width=7}
crispr_genes <- c("Stat5b", "Stat5a", "Il2ra","Il2rb","Tgfbr1", "Tgfbr2", "Ptpn1","Nr4a3","Tnfrsf1b")
order <- c("TNF","TGFb","TGFb-TNF","TGFb:TNF","Stat5b", "Stat5a", "Il2ra", "Tgfbr2","Tnfrsf1b", "Ptpn1","Nr4a3")

# TGFb+TNF shown as the interaction term, not the overall combo estimate
int_lfc <- int_SIG29 %>%
  filter(gene %in% crispr_genes, condition == "TGFb-TNF") %>%
  transmute(gene,
            condition = "TGFb:TNF",
            log2FoldChange = lfc_interaction,
            padj = padj_interaction)

deg_lfc <- deg_SIG29 %>%
  filter(gene %in% crispr_genes, condition %in% c("TNF","TGFb","TGFb-TNF"))%>%
  transmute(gene,
            condition = condition,
            log2FoldChange = log2FoldChange,
            padj = padj)

crispr_targets <- sort(unique(deg_SIG30_targets$target))

crispr_lfc <- deg_SIG30_targets %>%
  filter(gene %in% crispr_genes) %>%
  rename(condition = target)

conditions <- c(deg_lfc$condition %>% unique(), "TGFb:TNF", crispr_targets)

mats <- build_lfc_padj_mats(bind_rows(deg_lfc, int_lfc, crispr_lfc), crispr_genes, conditions)
mats$lfc <- mats$lfc[,order]
mats$padj <- mats$padj[,order]

col_anno <- data.frame(
  perturbation = c(rep("SIG29", length(deg_lfc$condition %>% unique())),
                   "TGFb x TNF interaction",
                   rep("CRISPR KO", length(crispr_targets))),
  row.names = conditions
)

hm <- pheatmap(mats$lfc,
         cluster_cols = FALSE,
         cluster_rows = T,
         gaps_col = c(3,4),
         annotation_col = col_anno,
         color = lfc_color_scale,
         breaks = seq(-2, 2, length.out = 101),
         na_col = "grey90",
         display_numbers = stars_from_padj(mats$padj),
         fontsize_number = 12,
         number_color = "black",
         angle_col = 90)
save_pheatmap_pdf(hm, file.path(outs_dir, "program_crossreg/crossreg_hm_crispr_perturbations_all_SIG29.pdf"), width=7, height = 3)
```

```{r, fig.height=3, fig.width=8}
crispr_genes <- c("Stat5b", "Stat5a", "Il2ra","Il2rb","Tgfbr1", "Tgfbr2", "Ptpn1","Nr4a3","Tnfrsf1b")
ligand_conditions <- c("TGFb","TNF","TGFb-TNF","TL1A","TGFb-TL1A",
                       "OX40L","TGFb-OX40L","DTA1","TGFb-DTA1")
order <- c(ligand_conditions,
           "Stat5b", "Stat5a", "Il2ra", "Tgfbr2","Tnfrsf1b", "Ptpn1","Nr4a3")

deg_lfc <- deg_SIG29 %>%
  filter(gene %in% crispr_genes, condition %in% ligand_conditions)%>%
  transmute(gene,
            condition = condition,
            log2FoldChange = log2FoldChange,
            padj = padj)

crispr_targets <- sort(unique(deg_SIG30_targets$target))

crispr_lfc <- deg_SIG30_targets %>%
  filter(gene %in% crispr_genes) %>%
  rename(condition = target)

conditions <- c(deg_lfc$condition %>% unique(), crispr_targets)

mats <- build_lfc_padj_mats(bind_rows(deg_lfc, crispr_lfc), crispr_genes, conditions)
mats$lfc <- mats$lfc[,order]
mats$padj <- mats$padj[,order]

col_anno <- data.frame(
  cat = c(rep("SIG29 LFC", length(deg_lfc$condition %>% unique())),
                   rep("CRISPR KO LFC", length(crispr_targets))),
  row.names = conditions
)

hm <- pheatmap(mats$lfc,
         cluster_cols = FALSE,
         cluster_rows = T,
         gaps_col = length(ligand_conditions),
         annotation_col = col_anno,
         color = lfc_color_scale,
         breaks = seq(-2, 2, length.out = 101),
         na_col = "grey90",
         display_numbers = stars_from_padj(mats$padj),
         fontsize_number = 12,
         number_color = "black",
         angle_col = 90)
save_pheatmap_pdf(hm, file.path(outs_dir, "program_crossreg/crossreg_hm_crispr_perturbations_SIG29_ligand_panel.pdf"), width=8, height = 3)
```

```{r, fig.height=3, fig.width=6}
crispr_genes <- c("Stat5b", "Stat5a", "Il2ra","Il2rb","Tgfbr1", "Tgfbr2", "Ptpn1","Nr4a3","Tnfrsf1b")
order <- c("IL2","TNF","TGFb","IL2-TNF","Stat5b", "Stat5a", "Il2ra", "Tgfbr2","Tnfrsf1b", "Ptpn1","Nr4a3")

deg_lfc <- deg_SIG14 %>%
  filter(gene %in% crispr_genes, condition %in% c("TNF","IL2","TGFb","IL2-TNF"))%>%
  transmute(gene,
            condition = condition,
            log2FoldChange = log2FoldChange,
            padj = padj)

crispr_targets <- sort(unique(deg_SIG30_targets$target))

crispr_lfc <- deg_SIG30_targets %>%
  filter(gene %in% crispr_genes) %>%
  rename(condition = target)

conditions <- c(deg_lfc$condition %>% unique(), crispr_targets)

mats <- build_lfc_padj_mats(bind_rows(deg_lfc, crispr_lfc), crispr_genes, conditions)
mats$lfc <- mats$lfc[,order]
mats$padj <- mats$padj[,order]

col_anno <- data.frame(
  cat = c(rep("SIG14 LFC", length(deg_lfc$condition %>% unique())),
                   rep("CRISPR KO LFC", length(crispr_targets))),
  row.names = conditions
)

hm <- pheatmap(mats$lfc,
         cluster_cols = FALSE,
         cluster_rows = F,
         gaps_col = 3,
         annotation_col = col_anno,
         color = lfc_color_scale,
         breaks = seq(-2, 2, length.out = 101),
         na_col = "grey90",
         display_numbers = stars_from_padj(mats$padj),
         fontsize_number = 12,
         number_color = "black",
         angle_col = 90)
save_pheatmap_pdf(hm, file.path(outs_dir, "program_crossreg/crossreg_hm_crispr_perturbations_SIG14_Tgfb_Tnf_single.pdf"), width=6, height = 3)
```

```{r, fig.height=2.5, fig.width=8}
crispr_genes <- int_SIG29$gene[grepl("^Ptpn|^Nr4a", int_SIG29$gene)] %>% unique() %>% sort()
order <- c("TNF","TGFb","TGFb-TNF","TGFb:TNF","Stat5b", "Stat5a", "Il2ra", "Tgfbr2","Tnfrsf1b", "Ptpn1","Nr4a3")
# TGFb+TNF shown as the interaction term, not the overall combo estimate
int_lfc <- int_SIG29 %>%
  filter(gene %in% crispr_genes, condition == "TGFb-TNF") %>%
  transmute(gene,
            condition = "TGFb:TNF",
            log2FoldChange = lfc_interaction,
            padj = padj_interaction)
deg_lfc <- deg_SIG29 %>%
  filter(gene %in% crispr_genes, condition %in% c("TNF","TGFb","TGFb-TNF"))%>%
  transmute(gene,
            condition = condition,
            log2FoldChange = log2FoldChange,
            padj = padj)
crispr_targets <- sort(unique(deg_SIG30_targets$target))
crispr_lfc <- deg_SIG30_targets %>%
  filter(gene %in% crispr_genes) %>%
  rename(condition = target)
conditions <- c(deg_lfc$condition %>% unique(), "TGFb:TNF", crispr_targets)
mats <- build_lfc_padj_mats(bind_rows(deg_lfc, int_lfc, crispr_lfc), crispr_genes, conditions)
mats$lfc <- mats$lfc[,order]
mats$padj <- mats$padj[,order]

# Transpose so conditions become rows and genes become columns
mats$lfc <- t(mats$lfc)
mats$padj <- t(mats$padj)

row_anno <- data.frame(
  cat = c(rep("SIG29 overall LFC", length(deg_lfc$condition %>% unique())),
                   "TGFb x TNF interaction coeff",
                   rep("CRISPR KO LFC", length(crispr_targets))),
  row.names = conditions
)
hm <- pheatmap(mats$lfc,
         cluster_cols = FALSE,
         cluster_rows = F,
         gaps_row = c(3,4),
         annotation_row = row_anno,
         color = lfc_color_scale,
         breaks = seq(-2, 2, length.out = 101),
         na_col = "grey90",
         display_numbers = stars_from_padj(mats$padj),
         fontsize_number = 12,
         number_color = "black",
         angle_col = 90)
save_pheatmap_pdf(hm, file.path(outs_dir, "program_crossreg/crossreg_hm_nr4a_ptpn_family_SIG29.pdf"), width=8, height = 2.5)
```





