---
title: "DEG Analysis SIG30"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(DESeq2)
library(IHW)
library(RColorBrewer)
library(pheatmap)
library(ggthemes)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Repo-relative paths: inputs from imports_stable/, outputs to analysis_outs/<module>/
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "07_crispr_arrayed_validation_RNAseq_SIG30")
dir.create(outs_dir, recursive = TRUE, showWarnings = FALSE)
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
theme_set(theme_Publication())

```

### [Import Data]{.underline}

```{r}
counts <- read.csv(file.path(imports_dir, "SIG30/processing_outs/count_matrix_umiDeDup_SIG30.csv"), row.names = "X")
colnames(counts) <- gsub("\\.", "-", colnames(counts))

metadata <- read_csv(file.path(imports_dir, "SIG30/processing_outs/processed_metadata_SIG30.csv"))

featureNames <- read_csv(file.path(imports_dir, "SIG30/processing_outs/featureNames_SIG30.csv")) %>%
  dplyr::select(-category)
```

```{r}
# set metadata factors (control = reference level)
metaDF <- metadata %>%
  mutate(
    replicate = factor(replicate),
    target    = factor(target, levels = c("control", sort(setdiff(unique(target), "control"))))
  ) %>%
  as.data.frame()
rownames(metaDF) <- metaDF$sample_ID

counts <- counts[, rownames(metaDF)]
```

### [Helper Functions]{.underline}

```{r}
# Build DDS object with counts filtering
build_dds <- function(counts_mat, meta, design, min_count = 10, min_samples = 3) {
  dds <- DESeqDataSetFromMatrix(counts_mat, colData = meta, design = design)
  dds <- dds[rowSums(counts(dds) >= min_count) >= min_samples, ]
  DESeq(dds, parallel = TRUE)
}

# Extract DEG results for a vector of coefficient names and combine into one tibble
extract_deg_results <- function(dds, terms, label_col = "term", clean_fn = identity) {
  map_dfr(terms, function(term) {
    results(dds, name = term,
            independentFiltering = FALSE, filterFun = ihw, parallel = TRUE) %>%
      as_tibble(rownames = "ensembl_ID") %>%
      left_join(featureNames, by = "ensembl_ID") %>%
      mutate(!!label_col := clean_fn(term)) %>%
      arrange(padj)
  })
}
```

### [QC Plots]{.underline}

```{r, message=FALSE}
# Target-level model on all samples; reused for both QC and the DEG analysis below
dds_target <- build_dds(counts, metaDF, ~ target + replicate)
vsd        <- vst(dds_target, blind = TRUE)
```

#### Sample Distance Heatmap

```{r, fig.height=10, fig.width=11}
sampleDists <- dist(t(assay(vsd)))
distMatrix  <- as.matrix(sampleDists)
rownames(distMatrix) <- colnames(dds_target)
colnames(distMatrix) <- colnames(dds_target)

ann_df <- as.data.frame(colData(vsd)[, c("target", "replicate")])
mk_pal <- function(n) colorRampPalette(tableau_color_pal("Tableau 10")(10))(n)
ann_colors <- list(
  target    = setNames(mk_pal(length(unique(ann_df$target))), sort(unique(ann_df$target))),
  replicate = setNames(mk_pal(length(unique(ann_df$replicate))), sort(unique(ann_df$replicate)))
)

pheatmap(distMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = viridisLite::viridis(100),
         annotation_row = ann_df,
         annotation_colors = ann_colors,
         show_rownames = FALSE,
         show_colnames = FALSE)
```

#### Per-Replicate Distance Heatmaps

```{r, fig.height=8, fig.width=9}
for (rep in levels(metaDF$replicate)) {
  keep    <- colnames(vsd)[colData(vsd)$replicate == rep]
  vsd_rep <- vsd[, keep]

  dists_rep <- dist(t(assay(vsd_rep)))
  mat_rep   <- as.matrix(dists_rep)
  rownames(mat_rep) <- colnames(vsd_rep)
  colnames(mat_rep) <- colnames(vsd_rep)

  ann_rep <- as.data.frame(colData(vsd_rep)[, "target", drop = FALSE])
  ann_colors_rep <- list(
    target = setNames(mk_pal(length(unique(ann_rep$target))), sort(unique(ann_rep$target)))
  )

  pheatmap(mat_rep,
           clustering_distance_rows = dists_rep,
           clustering_distance_cols = dists_rep,
           col = viridisLite::viridis(100),
           annotation_row = ann_rep,
           annotation_colors = ann_colors_rep,
           show_rownames = FALSE,
           show_colnames = FALSE,
           main = paste("Replicate:", rep))
}
```

#### PCA

```{r, fig.height=5, fig.width=10}
p1 <- plotPCA(vsd, intgroup = "target",    ntop = 2000) + theme(aspect.ratio = 1) + scale_color_tableau()
p2 <- plotPCA(vsd, intgroup = "replicate", ntop = 2000) + theme(aspect.ratio = 1)
p1 + p2 + plot_layout(ncol = 2)
```

### [Analysis: Target (KO) Effects]{.underline}

Effect of every CRISPR-KO target vs. `control`, using the target-level model.
SIG30 has only single-gene KO perturbations (no combination conditions), so —
unlike the SIG26 deg analysis this was ported from — there is no
interaction/synergy scoring step here.

```{r}
target_terms <- grep("^target_", resultsNames(dds_target), value = TRUE)

res_targets <- extract_deg_results(
  dds_target, target_terms,
  label_col = "target",
  # DESeq2 may encode hyphens as dots in term names; restore hyphens for readability
  clean_fn  = function(x) gsub("\\.", "-", gsub("^target_|_vs_control$", "", x))
)
```

```{r}
write_csv(res_targets, file.path(outs_dir, "res_targets_SIG30.csv"))
```

```{r, fig.width=5, fig.height=5}
res_targets %>%
  filter(padj < 0.1) %>%
  group_by(target) %>%
  summarise(n_degs = n()) %>%
  ggplot(aes(x = target, y = n_degs)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```
