---
title: "DEG Analysis SIG29"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(DESeq2)
library(IHW)
library(RColorBrewer)
library(pheatmap)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Repo-relative paths: inputs from imports_stable/, outputs to analysis_outs/<module>/
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "08_tnf_family_tgfb_interaction_RNAseq_SIG29")
dir.create(outs_dir, recursive = TRUE, showWarnings = FALSE)
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
theme_set(theme_Publication())
```

### [Import Data]{.underline}

```{r}
counts <- read.csv(file.path(imports_dir, "SIG29/processing_outs/count_matrix_umiDeDup_SIG29.csv"), row.names = "X")
colnames(counts) <- gsub("\\.", "-", colnames(counts))

metadata <- read_csv(file.path(imports_dir, "SIG29/processing_outs/processed_metadata_SIG29.csv"))

featureNames <- read_csv(file.path(imports_dir, "SIG29/processing_outs/featureNames_SIG29.csv")) %>%
  dplyr::select(-category)
```

```{r}
# set metadata factors
metaDF <- metadata %>%
  mutate(
    replicate = factor(replicate),
    ligand1   = factor(ligand1, levels = c("none", sort(setdiff(unique(ligand1), "none")))),
    ligand2   = factor(ligand2, levels = c("none", sort(setdiff(unique(ligand2), "none")))),
    condition = factor(condition, levels = c("none_none", "TGFb_none",
                                             "none_TNF","TGFb_TNF",
                                             "none_TL1A","TGFb_TL1A",
                                             "none_OX40L","TGFb_OX40L",
                                             "none_DTA1","TGFb_DTA1"))
  ) %>%
  as.data.frame()
rownames(metaDF) <- metaDF$sample_ID

counts <- counts[, rownames(metaDF)]
```

### [Helper Functions]{.underline}

```{r}
# Build DDS object with counts filtering
build_dds <- function(counts_mat, meta, design, min_count = 10, min_samples = 4) {
  dds <- DESeqDataSetFromMatrix(counts_mat, colData = meta, design = design)
  dds <- dds[rowSums(counts(dds) >= min_count) >= min_samples, ]
  DESeq(dds, parallel = TRUE)
}

# Extract DEG results for a vector of coefficient names and combine into one tibble
extract_deg_results <- function(dds, terms, label_col = "term", clean_fn = identity) {
  map_dfr(terms, function(term) {
    results(dds, name = term,
            independentFiltering = FALSE, filterFun = ihw, parallel = TRUE) %>%
      as_tibble(rownames = "ensembl_ID") %>%
      left_join(featureNames, by = "ensembl_ID") %>%
      mutate(!!label_col := clean_fn(term)) %>%
      arrange(padj)
  })
}

# Look up the exact resultsNames() term for a condition value.
# Handles DESeq2 encoding hyphens as either "-" or "." in model matrix names.
cond_term <- function(res_names, cond) {
  pat  <- paste0("^condition_", gsub("-", "[-.]", cond, fixed = FALSE), "_vs_none.none$")
  hits <- res_names[grepl(pat, res_names)]
  if (length(hits) != 1) stop("Expected 1 match for '", cond, "', found: ", length(hits))
  hits
}
```

### [QC Plots]{.underline}

```{r, message=FALSE}
# Condition-level model on all samples; reused for both QC and all DEG analyses below
dds_cond <- build_dds(counts, metaDF, ~ condition + replicate)
vsd      <- vst(dds_cond, blind = TRUE)
```

#### Sample Distance Heatmap

```{r, fig.height=10, fig.width=11}
sampleDists <- dist(t(assay(vsd)))
distMatrix  <- as.matrix(sampleDists)
rownames(distMatrix) <- colnames(dds_cond)
colnames(distMatrix) <- colnames(dds_cond)

ann_df <- as.data.frame(colData(vsd)[, c("ligand1", "ligand2","replicate")])
mk_pal <- function(n) colorRampPalette(brewer.pal(8, "Dark2"))(n)
ann_colors <- list(
  ligand1 = setNames(mk_pal(length(unique(ann_df$ligand1))), sort(unique(ann_df$ligand1))),
  ligand2 = setNames(mk_pal(length(unique(ann_df$ligand2))), sort(unique(ann_df$ligand2))),
  replicate = setNames(mk_pal(length(unique(ann_df$replicate))), sort(unique(ann_df$replicate)))
)

pheatmap(distMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = viridisLite::viridis(100),
         annotation_row = ann_df,
         annotation_colors = ann_colors,
         show_rownames = FALSE,
         show_colnames = FALSE)
```

#### Per-Replicate Distance Heatmaps

```{r, fig.height=8, fig.width=9}
for (rep in levels(metaDF$replicate)) {
  keep    <- colnames(vsd)[colData(vsd)$replicate == rep]
  vsd_rep <- vsd[, keep]

  dists_rep  <- dist(t(assay(vsd_rep)))
  mat_rep    <- as.matrix(dists_rep)
  rownames(mat_rep) <- colnames(vsd_rep)
  colnames(mat_rep) <- colnames(vsd_rep)

  ann_rep <- as.data.frame(colData(vsd_rep)[, c("ligand1", "ligand2")])
  ann_colors_rep <- list(
    ligand1 = setNames(mk_pal(length(unique(ann_rep$ligand1))), sort(unique(ann_rep$ligand1))),
    ligand2 = setNames(mk_pal(length(unique(ann_rep$ligand2))), sort(unique(ann_rep$ligand2)))
  )

  pheatmap(mat_rep,
           clustering_distance_rows = dists_rep,
           clustering_distance_cols = dists_rep,
           col = viridisLite::viridis(100),
           annotation_row = ann_rep,
           annotation_colors = ann_colors_rep,
           show_rownames = FALSE,
           show_colnames = FALSE,
           main = paste("Replicate:", rep))
}
```

#### PCA

```{r, fig.height=10, fig.width=10}
p1 <- plotPCA(vsd, intgroup = "ligand1",   ntop = 2000) + theme(aspect.ratio = 1)
p2 <- plotPCA(vsd, intgroup = "ligand2",   ntop = 2000) + theme(aspect.ratio = 1)
p3 <- plotPCA(vsd, intgroup = "replicate", ntop = 2000) + theme(aspect.ratio = 1)
p4 <- plotPCA(vsd, intgroup = "condition", ntop = 2000) + theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Paired")
p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
ggsave(file.path(outs_dir, "SIG29_PCA_plots.pdf"))
```

### [Analysis 1: Condition Effects]{.underline}

Effect of every condition vs. `none_none`, using the condition-level model.
Captures both single-ligand (`TGFb_none`, `none_TNF`, ...) and combination
(`TGFb_TNF`, `TGFb_TL1A`, ...) conditions in a single unified analysis.

```{r}
cond_terms <- grep("^condition_", resultsNames(dds_cond), value = TRUE)

res_conditions <- extract_deg_results(
  dds_cond, cond_terms,
  label_col = "condition",
  # DESeq2 may encode hyphens as dots in term names; restore hyphens for readability
  clean_fn  = function(x) gsub("\\.", "-", gsub("^condition_|_vs_none.none$", "", x))
)
```

```{r}
write_csv(res_conditions, file.path(outs_dir, "res_conditions_SIG29.csv"))
```

```{r}
res_conditions %>%
  filter(padj < 0.01) %>%
  group_by(condition) %>%
  summarise(n_degs = n()) %>%
  ggplot(aes(x = condition, y = n_degs)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### [Analysis 2: Interaction Scoring]{.underline}

For each gene in each combination condition (TGFb + a TNF-family member),
quantifies whether the combined response is greater than (synergy), less
than (buffering), or equal to (additive) the sum of the individual
responses. Since every SIG29 combo has `ligand1 = TGFb`, this directly
yields one interaction profile per TNF-family member (`TNF`, `TL1A`,
`OX40L`, `DTA1`), which is what the deg-visualization FC-FC plots compare.

**Definitions** (following SIG18 scoring):

All LFC values are pulled directly from the DESeq2 model.

- `lfc_lig1`: logFC of ligand1 (`TGFb`) alone vs. `none_none`
- `lfc_lig2`: logFC of ligand2 (TNF-family member) alone vs. `none_none`
- `lfc_interaction`: `lfc_combo − lfc_lig1 − lfc_lig2` (deviation from additivity)
- `lfc_total`: `lfc_lig1 + lfc_lig2 + lfc_interaction` = `lfc_combo`
- `interaction_score`: `lfc_interaction / lfc_total` — fraction of total effect attributable to non-additivity (set to 0 when `padj_interaction > 0.1`)
- `padj_interaction`: adjusted p-value from the synergy contrast `lfc_combo − lfc_lig1 − lfc_lig2`

**Classification**:
- `synergy positive`: interaction significant, total effect positive, score > 0
- `synergy negative`: interaction significant, total effect negative, score > 0
- `buffering`: interaction significant, score < 0
- `none`: interaction not significant

```{r}
# Define combo conditions and respective ligands
combo_map <- metaDF %>%
  filter(as.character(ligand1) != "none", as.character(ligand2) != "none") %>%
  dplyr::select(condition, ligand1, ligand2) %>%
  mutate(across(everything(), as.character)) %>%
  distinct()

combo_map
```

#### Synergy contrast p-values

Here we are calculating interaction effects using a contrast design.
e.g. TGFb-TNF_vs_none_none - (TGFb-none_vs_none_none + none-TNF_vs_none_none)

```{r}
rn_cond <- resultsNames(dds_cond)

res_synergy_pvals <- map_dfr(seq_len(nrow(combo_map)), function(i) {
  row <- combo_map[i, ]
  results(dds_cond,
          contrast = list(
            c(cond_term(rn_cond, row$condition)),
            c(cond_term(rn_cond, paste0(row$ligand1, "_none")),
              cond_term(rn_cond, paste0("none_", row$ligand2)))
          ),
          independentFiltering = FALSE, filterFun = ihw, parallel = TRUE) %>%
    as_tibble(rownames = "ensembl_ID") %>%
    dplyr::select(ensembl_ID, lfc_interaction = log2FoldChange, padj_interaction = padj) %>%
    mutate(condition = row$condition)
})
```

#### Compute interaction scores

```{r}
lfc_lookup <- res_conditions %>%
  dplyr::select(ensembl_ID, gene, condition, log2FoldChange) %>%
  dplyr::rename(lfc = log2FoldChange)

interaction_raw <- map_dfr(seq_len(nrow(combo_map)), function(i) {
  row <- combo_map[i, ]
  lig1_cond <- paste0(row$ligand1, "_none")
  lig2_cond <- paste0("none_", row$ligand2)
  lfc_lookup %>%
    filter(condition == row$condition) %>%
    dplyr::select(ensembl_ID, gene, lfc_combo = lfc) %>%
    left_join(
      lfc_lookup %>% filter(condition == lig1_cond) %>%
        dplyr::select(ensembl_ID, lfc_lig1 = lfc),
      by = "ensembl_ID"
    ) %>%
    left_join(
      lfc_lookup %>% filter(condition == lig2_cond) %>%
        dplyr::select(ensembl_ID, lfc_lig2 = lfc),
      by = "ensembl_ID"
    ) %>%
    left_join(
      res_synergy_pvals %>% filter(condition == row$condition) %>%
        dplyr::select(ensembl_ID, lfc_interaction, padj_interaction),
      by = "ensembl_ID"
    ) %>%
    mutate(condition = row$condition, ligand1 = row$ligand1, ligand2 = row$ligand2)
})
```

#### Classify interaction effects

```{r}
interaction_scored <- interaction_raw %>%
  mutate(
    lfc_total       = lfc_combo,
    interaction_score = case_when(
      is.na(padj_interaction) | padj_interaction > 0.1 ~ 0,
      TRUE ~ lfc_interaction / lfc_total
    ),
    interaction_class = case_when(
      padj_interaction <= 0.1 & lfc_total > 0 & interaction_score > 0 ~ "synergy positive",
      padj_interaction <= 0.1 & lfc_total < 0 & interaction_score > 0 ~ "synergy negative",
      padj_interaction <= 0.1 & interaction_score < 0                  ~ "buffering",
      TRUE ~ "none"
    )
  )
```

```{r}
interaction_scored %>%
  filter(interaction_class != "none") %>%
  group_by(condition, interaction_class) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(condition, interaction_class)
```

```{r}
write_csv(interaction_scored, file.path(outs_dir, "res_interaction_scored_SIG29.csv"))
```
