---
title: "DEG Visualization SIG29: TNF-family interaction FC-FC"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
---

Visualizes whether the TGFb:TNF interaction (synergy/buffering) effect from
`01_deg_analysis_SIG29.Rmd` (stable copy read from
`imports_stable/SIG29/analysis_outs/res_interaction_scored_SIG29.csv`)
is **conserved across other TNF-family members** (`TL1A`, `OX40L`, `DTA1`)
stimulated alongside TGFb in the same experiment, via per-gene FC-FC plots:
x = the gene's `TGFb:TNF` interaction LFC, y = the same gene's `TGFb:<member>`
interaction LFC.

```{r setup}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggrepel)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Repo-relative paths: inputs from imports_stable/, outputs to analysis_outs/<module>/
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "08_tnf_family_tgfb_interaction_RNAseq_SIG29")
dir.create(outs_dir, recursive = TRUE, showWarnings = FALSE)
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
theme_set(theme_Publication())
```

### [Import Data]{.underline}

```{r}
interaction_scored <- read_csv(file.path(imports_dir, "SIG29/analysis_outs/res_interaction_scored_SIG29.csv"))

# TNF is the reference TNF-family member; TL1A/OX40L/DTA1 are compared against it
REF_MEMBER <- "TNF"
OTHER_MEMBERS <- setdiff(unique(interaction_scored$ligand2), REF_MEMBER)
OTHER_MEMBERS
```

### [Build FC-FC table]{.underline}

One row per gene per non-reference member, with both the reference (`TNF`)
and comparison member's interaction LFC/significance/class joined side by
side.

```{r}
ref_int <- interaction_scored %>%
  filter(ligand2 == REF_MEMBER) %>%
  dplyr::select(ensembl_ID, gene,
                lfc_interaction_ref = lfc_interaction,
                padj_interaction_ref = padj_interaction,
                interaction_class_ref = interaction_class)

fc_fc <- interaction_scored %>%
  filter(ligand2 %in% OTHER_MEMBERS) %>%
  dplyr::select(ensembl_ID, gene, member = ligand2,
                lfc_interaction_member = lfc_interaction,
                padj_interaction_member = padj_interaction,
                interaction_class_member = interaction_class) %>%
  inner_join(ref_int, by = c("ensembl_ID", "gene")) %>%
  mutate(
    sig_ref    = !is.na(padj_interaction_ref)    & padj_interaction_ref    < 0.1,
    sig_member = !is.na(padj_interaction_member) & padj_interaction_member < 0.1,
    concordance = case_when(
      sig_ref & sig_member &
        sign(lfc_interaction_ref) == sign(lfc_interaction_member) ~ "concordant (both sig)",
      sig_ref & sig_member                                        ~ "discordant (both sig)",
      sig_ref | sig_member                                        ~ "sig in one",
      TRUE                                                        ~ "n.s. in both"
    )
  )

write_csv(fc_fc, file.path(outs_dir, "res_interaction_fcfc_SIG29.csv"))
```

### [FC-FC plots: TGFb:member interaction vs TGFb:TNF interaction]{.underline}

Per-facet Pearson correlation (computed over all genes, not just significant
ones) summarizes how conserved the interaction program is between TNF and
each family member; genes significant in both are labelled.

```{r}
rho <- fc_fc %>%
  group_by(member) %>%
  summarise(r = cor(lfc_interaction_ref, lfc_interaction_member, use = "complete.obs"),
            n_concordant = sum(concordance == "concordant (both sig)"),
            n_discordant = sum(concordance == "discordant (both sig)"),
            .groups = "drop")
rho
```

```{r, fig.width=12, fig.height=4}
fc_fc %>%
  left_join(rho, by = "member") %>%
  mutate(facet_label = paste0("TGFb:", member, " vs TGFb:", REF_MEMBER,
                              " (r = ", round(r, 2), ")")) %>%
  ggplot(aes(x = lfc_interaction_ref, y = lfc_interaction_member)) +
  geom_hline(yintercept = 0, lty = 2, color = "grey70") +
  geom_vline(xintercept = 0, lty = 2, color = "grey70") +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(size = 0.8, alpha = 0.7) +
  facet_wrap(~ facet_label, nrow = 1) +
  theme(aspect.ratio = 1) +
  labs(x = paste0("TGFb:", REF_MEMBER, " interaction LFC"),
       y = "TGFb:member interaction LFC")

ggsave(file.path(outs_dir, "res_interaction_fcfc_simple_SIG29.pdf"))
```

```{r, fig.width=12, fig.height=4}
concordance_palette <- c(
  "concordant (both sig)"  = "#D55E00",
  "discordant (both sig)"  = "#0072B2",
  "sig in one"             = "#009E73",
  "n.s. in both"           = "grey80"
)

fc_fc %>%
  left_join(rho, by = "member") %>%
  mutate(facet_label = paste0("TGFb:", member, " vs TGFb:", REF_MEMBER,
                              " (r = ", round(r, 2), ")")) %>%
  ggplot(aes(x = lfc_interaction_ref, y = lfc_interaction_member)) +
  geom_hline(yintercept = 0, lty = 2, color = "grey70") +
  geom_vline(xintercept = 0, lty = 2, color = "grey70") +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(aes(color = concordance), size = 0.8, alpha = 0.7) +
  geom_text_repel(aes(label = ifelse(concordance == "concordant (both sig)" |
                                     concordance == "discordant (both sig)", gene, "")),
                  size = 2, max.overlaps = 15) +
  scale_color_manual(values = concordance_palette) +
  facet_wrap(~ facet_label, nrow = 1) +
  theme(aspect.ratio = 1) +
  labs(x = paste0("TGFb:", REF_MEMBER, " interaction LFC"),
       y = "TGFb:member interaction LFC",
       color = "concordance\n(padj < 0.1)")

ggsave(file.path(outs_dir, "res_interaction_fcfc_SIG29.pdf"))
```



### [Concordance summary]{.underline}

Fraction of genes significant in the reference (`TNF`) interaction that are
concordant (same sign, also significant) vs. discordant vs. not significant
in each other member.

```{r, fig.width=8, fig.height=5}
fc_fc %>%
  filter(sig_ref) %>%
  count(member, concordance) %>%
  group_by(member) %>%
  mutate(frac = n / sum(n)) %>%
  ggplot(aes(x = member, y = frac, fill = concordance)) +
  geom_col() +
  scale_fill_manual(values = concordance_palette) +
  labs(x = "TNF-family member", y = "fraction of TGFb:TNF-significant genes",
       title = "Conservation of TGFb:TNF-significant interaction genes in other members")
```
