---
title: "Ligand Activity Regressions: AMP 2023 (rheumatoid arthritis)"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
source("regression_helpers.R")
```

# Goal

Test which SIG13 ligand activities differ in RA synovial CD4 T cells (AMP Phase 2),
using the activity scores from `02_score_disease_datasets.py`.

Two questions, each asked across all CD4 T cells and per celltype:

1. **Disease** -- RA vs osteoarthritis control
2. **CTAP** -- do the AMP cell-type-abundance phenotypes (the synovial tissue
   "neighbourhoods") differ in the combinatorial signals CD4 T cells experience?

**Covariates:** sex, scaled age, and joint in every model. Age is complete for all
77 biopsies, so unlike the atlas there is no missingness argument for binning it.
Joint matters most: the osteoarthritis controls are drawn from different joints than
the RA biopsies (RA is mostly knee and wrist), so it is adjusted for throughout. The
earlier analysis omitted age from the celltype models but not the whole-biopsy ones.

# Import

```{r}
AMP_KEYS <- c("subject_id", "biopsy_id", "joint", "sex", "treatment", "CTAP",
              "das28_crp3_binary")

# `Repeat` biopsies are repeat sampling of the same joint, excluded here; they are
# also exactly the samples with no CTAP assignment.
prepare_amp <- function(df) {
  df %>%
    filter(treatment != "Repeat") %>%
    mutate(disease_binary = factor(ifelse(treatment == "osteoarthritis control",
                                          "control", "RA"), c("RA", "control")),
           das28_crp3_binary = factor(das28_crp3_binary, c("high", "low", "control")),
           across(c(sex, joint, CTAP), factor),
           age = as.vector(scale(age)))
}

amp <- load_activity_scores("amp_2023", AMP_KEYS) %>% prepare_amp()
amp_celltype <- load_activity_scores("amp_2023_celltype", c(AMP_KEYS, "cluster_name")) %>%
  rename(celltype = cluster_name) %>%
  prepare_amp() %>%
  mutate(celltype = fct_relevel(factor(celltype), "T-4: CD4+ naive"))

activities <- activity_names()
cat(nrow(amp), "biopsies,", length(activities), "ligand activities\n")
count(amp, disease_binary, CTAP)
```

# Covariate QC

```{r, fig.height=6, fig.width=8}
amp %>%
  pivot_longer(c(age, CDAI, das28_crp3, das28_esr3, krenn_inflammation, krenn_lining),
               names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = treatment, y = value, fill = treatment)) +
  geom_boxplot(show.legend = FALSE, na.rm = TRUE) +
  facet_wrap(~variable, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        aspect.ratio = 1) +
  labs(title = "Continuous covariates by treatment group", x = "", y = "")
```

```{r, fig.height=5, fig.width=12}
map(c("joint", "sex", "CTAP"), function(variable) {
  ggplot(amp, aes(x = treatment, fill = !!sym(variable))) +
    geom_bar(position = "fill") +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          aspect.ratio = 1) +
    labs(title = variable, x = "", y = "proportion")
}) %>% wrap_plots(nrow = 1)
```

Joint is the one notable imbalance -- the osteoarthritis controls are dominated by
one joint -- so it is adjusted for in every model below.

# Disease associations

```{r}
res_disease <- fit_and_save(
  amp,
  rhs = "disease_binary + sex + age + joint",
  emm = ~ disease_binary,
  contrast_args = list(method = "pairwise"),
  name = "amp_2023_disease_binary"
)
```

```{r, fig.height=3.5, fig.width=3.5}
res_disease %>%
  ggplot(aes(x = estimate, y = -log10(padj))) +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dotted", color = "red") +
  theme(aspect.ratio = 1) +
  labs(title = "RA vs OA control")
```

# CTAP associations

Each CTAP against the osteoarthritis controls. Disease severity is not adjusted for
here because it is undefined for the controls.

```{r}
res_ctap <- fit_and_save(
  amp,
  rhs = "CTAP + sex + age + joint",
  emm = ~ CTAP,
  contrast_args = list(method = "trt.vs.ctrl", ref = "control"),
  name = "amp_2023_ctap_control_ref"
)
```

# Celltype associations

Which activities distinguish CD4 subsets from each other, irrespective of disease.
`del.eff` contrasts each celltype against the mean of the others; HC3 robust
covariance is used because the per-celltype sample sizes are uneven.

```{r}
res_celltype <- fit_and_save(
  amp_celltype,
  rhs = "celltype + sex + age + joint",
  emm = ~ celltype,
  contrast_args = list(method = "del.eff"),
  robust = TRUE,
  name = "amp_2023_celltype"
)
```

## Disease x celltype

```{r}
res_disease_celltype <- fit_and_save(
  amp_celltype,
  rhs = "disease_binary * celltype + sex + age + joint",
  emm = ~ disease_binary | celltype,
  contrast_args = list(method = "pairwise"),
  padj_by = c("contrast", "celltype"),
  name = "amp_2023_disease_celltype"
)
```

## CTAP x celltype

HC3 is unstable for this model (many CTAP x celltype cells are small), so
model-based standard errors are used.

```{r}
res_ctap_celltype <- fit_and_save(
  amp_celltype,
  rhs = "CTAP * celltype + sex + age + joint",
  emm = ~ CTAP | celltype,
  contrast_args = list(method = "trt.vs.ctrl", ref = "control"),
  padj_by = c("contrast", "celltype"),
  name = "amp_2023_ctap_celltype_control_ref"
)
```

```{r}
res_ctap_celltype %>%
  filter(padj < 0.1) %>%
  count(celltype, name = "significant_activity_contrasts") %>%
  arrange(desc(significant_activity_contrasts))
```

Results are written to `analysis_outs/03_activity_inference_model/inference_model_disease_bulk/regressions/`
and visualized in `09_viz_amp_2023.Rmd`.
