---
title: "Ligand Activity Regressions: Thomas IBD"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
source("regression_helpers.R")
```

# Goal

Test which SIG13 ligand activities differ in colonic CD4 T cells in UC and CD
(Thomas et al. 2024), using the activity scores from `02_score_disease_datasets.py`.

Model families, each fit across all conventional CD4 T cells and per celltype:

1. **Disease** (pre-treatment only) -- UC / CD / Healthy, pairwise
2. **Disease x remission** (pre-treatment) -- does baseline activity predict who
   later goes into remission?
3. **Vs healthy** -- every disease x remission x timepoint group against healthy
4. **Post vs pre treatment** -- within each disease x remission group
5. **Celltype** -- which activities distinguish CD4 subsets

All models adjust for sex and biopsy site and use HC3 robust covariance (patient
numbers per group are uneven). Age is deliberately *not* a covariate: the healthy
donors are much older than every patient group, so age is collinear with disease.

# Import

```{r}
IBD_KEYS <- c("Disease", "Patient", "Gender", "Remission_status", "Site",
              "Inflammation", "Treatment")

# The healthy donors carry placeholder metadata: no remission status and no
# timepoint (stored as the string "nan"). Renaming the timepoint to "None" gives
# the clean reference group label "Healthy_None_None" -- the old analysis used
# "Healthy_None _nan", the stray space coming from a trailing space in
# Remission_status that step 02 now strips at the source.
prepare_ibd <- function(df) {
  df %>%
    mutate(Treatment = ifelse(Treatment == "nan", "None", Treatment),
           disease_remission = paste(Disease, Remission_status, sep = "_"),
           disease_remission_treatment = paste(Disease, Remission_status, Treatment, sep = "_"),
           patient_site = paste(Patient, Site, sep = "_"),
           sex = factor(Gender),
           site = factor(Site),
           Disease = factor(Disease, c("UC", "CD", "Healthy")),
           Treatment = factor(Treatment, c("Pre", "Post", "None")))
}

ibd <- load_activity_scores("thomas_ibd", IBD_KEYS) %>% prepare_ibd()
ibd_celltype <- load_activity_scores("thomas_ibd_celltype", c(IBD_KEYS, "final_analysis")) %>%
  rename(celltype = final_analysis) %>%
  prepare_ibd() %>%
  mutate(celltype = fct_relevel(factor(celltype), "CD4 naive T"))

activities <- activity_names()
cat(nrow(ibd), "samples,", length(activities), "ligand activities\n")
count(ibd, Disease, Remission_status, Treatment)
```

The Tregs are excluded from the whole-sample scores upstream (step 02), so `ibd`
is conventional CD4 T cells only; `ibd_celltype` retains every subset.

# Covariate QC

```{r, fig.height=4, fig.width=8}
ibd %>%
  pivot_longer(c(Age, Inflammation_score), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = Disease, y = value, fill = Disease)) +
  geom_boxplot(show.legend = FALSE, na.rm = TRUE) +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(~variable, scales = "free_y") +
  theme(aspect.ratio = 1) +
  labs(title = "Age and molecular inflammation score by disease", x = "", y = "")
```

Age is strongly confounded with disease (healthy donors are older than all patient
groups) and `Batch` partly so (all healthy donors are Batch 3) -- neither is used
as a covariate.

```{r, fig.height=4, fig.width=5}
ibd %>%
  ggplot(aes(x = Inflammation, y = Inflammation_score, fill = Inflammation)) +
  geom_boxplot() +
  theme(aspect.ratio = 1) +
  labs(title = "Molecular inflammation score\nvs histological call", x = "", y = "score")
```

The paper filters IBD samples to molecular inflamed (`Inflammation_score > 6.5`,
just above the healthy maximum) rather than trusting the histological call. That
filter is available below but, is **not** applied -- every model uses all samples.

```{r}
baseline_uninflamed <- ibd %>%
  filter(Treatment == "Pre", Disease %in% c("UC", "CD"), Inflammation_score < 6.5) %>%
  pull(patient_site)
cat(length(baseline_uninflamed), "pre-treatment IBD samples are molecularly uninflamed",
    "(not excluded)\n")
```

# Disease associations

## Pre-treatment disease

```{r}
ibd_pre <- filter(ibd, Treatment %in% c("Pre", "None"))

res_disease <- fit_and_save(
  ibd_pre,
  rhs = "Disease + sex + site",
  emm = ~ Disease,
  contrast_args = list(method = "pairwise"),
  robust = TRUE,
  name = "thomas_ibd_disease_pre"
)
```

```{r, fig.height=4, fig.width=8}
res_disease %>%
  ggplot(aes(x = estimate, y = -log10(padj))) +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dotted", color = "red") +
  facet_wrap(~contrast) +
  theme(aspect.ratio = 1) +
  labs(title = "Pre-treatment disease contrasts")
```

## Pre-treatment disease x remission

```{r}
res_disease_remission <- fit_and_save(
  ibd_pre %>%
    mutate(disease_remission = factor(disease_remission,
                                      c("UC_Remission", "UC_Non_Remission",
                                        "CD_Remission", "CD_Non_Remission",
                                        "Healthy_None"))),
  rhs = "disease_remission + sex + site",
  emm = ~ disease_remission,
  contrast_args = list(method = "pairwise"),
  robust = TRUE,
  name = "thomas_ibd_disease_remission_pre"
)
```

## Every group vs healthy

```{r}
# healthy donors have no remission status or timepoint, so they are the single
# reference group; UC patients with unknown remission status are dropped
ibd_vs_healthy <- ibd %>%
  filter(Treatment %in% c("Pre", "Post") | Disease == "Healthy") %>%
  filter(Remission_status != "Not_avail" | Disease == "Healthy") %>%
  mutate(disease_remission_treatment = factor(disease_remission_treatment))

res_vs_healthy <- fit_and_save(
  ibd_vs_healthy,
  rhs = "disease_remission_treatment + sex + site",
  emm = ~ disease_remission_treatment,
  contrast_args = list(method = "trt.vs.ctrl", ref = "Healthy_None_None"),
  robust = TRUE,
  name = "thomas_ibd_disease_treatment_vsHealthy"
)
```

## Post vs pre treatment

```{r}
ibd_paired <- ibd %>%
  filter(Treatment %in% c("Pre", "Post"), Remission_status != "Not_avail") %>%
  mutate(disease_remission = factor(disease_remission),
         Treatment = factor(Treatment, c("Pre", "Post")))

res_treatment <- fit_and_save(
  ibd_paired,
  rhs = "disease_remission * Treatment + sex + site",
  emm = ~ Treatment | disease_remission,
  contrast_args = list(method = "trt.vs.ctrl", ref = "Pre"),
  robust = TRUE,
  name = "thomas_ibd_disease_treatment"
)
```

# Celltype associations

## Celltype identity (pre-treatment)

```{r}
ibd_celltype_pre <- filter(ibd_celltype, Treatment %in% c("Pre", "None"))

res_celltype <- fit_and_save(
  ibd_celltype_pre,
  rhs = "celltype + sex + site",
  emm = ~ celltype,
  contrast_args = list(method = "del.eff"),
  robust = TRUE,
  name = "thomas_ibd_celltype_pre"
)
```

## Disease x celltype (pre-treatment)

```{r}
res_disease_celltype <- fit_and_save(
  ibd_celltype_pre,
  rhs = "Disease * celltype + sex + site",
  emm = ~ Disease | celltype,
  contrast_args = list(method = "pairwise"),
  robust = TRUE,
  padj_by = c("contrast", "celltype"),
  name = "thomas_ibd_disease_celltype_pre"
)
```

## Disease x remission x celltype (pre-treatment)

```{r}
res_disease_celltype_remission <- fit_and_save(
  ibd_celltype_pre %>%
    mutate(disease_remission = factor(disease_remission,
                                      c("UC_Remission", "UC_Non_Remission",
                                        "CD_Remission", "CD_Non_Remission",
                                        "Healthy_None"))),
  rhs = "disease_remission * celltype + sex + site",
  emm = ~ disease_remission | celltype,
  contrast_args = list(method = "pairwise"),
  robust = TRUE,
  padj_by = c("contrast", "celltype"),
  name = "thomas_ibd_disease_celltype_remission_pre"
)
```

## Post vs pre treatment x celltype

```{r}
res_treatment_celltype <- fit_and_save(
  ibd_celltype %>%
    filter(Treatment %in% c("Pre", "Post"), Remission_status != "Not_avail") %>%
    mutate(disease_remission_celltype = factor(paste(disease_remission, celltype, sep = "_")),
           Treatment = factor(Treatment, c("Pre", "Post"))),
  rhs = "disease_remission_celltype * Treatment + sex + site",
  emm = ~ Treatment | disease_remission_celltype,
  contrast_args = list(method = "trt.vs.ctrl", ref = "Pre"),
  robust = TRUE,
  padj_by = c("contrast", "disease_remission_celltype"),
  name = "thomas_ibd_disease_celltype_treatment"
)
```

## Every group vs healthy x celltype

```{r}
res_vs_healthy_celltype <- fit_and_save(
  ibd_celltype %>%
    filter(Treatment %in% c("Pre", "Post") | Disease == "Healthy") %>%
    filter(Remission_status != "Not_avail" | Disease == "Healthy") %>%
    mutate(disease_remission_treatment = factor(disease_remission_treatment)),
  rhs = "disease_remission_treatment * celltype + sex + site",
  emm = ~ disease_remission_treatment | celltype,
  contrast_args = list(method = "trt.vs.ctrl", ref = "Healthy_None_None"),
  robust = TRUE,
  padj_by = c("contrast", "celltype"),
  name = "thomas_ibd_disease_treatment_vsHealthy_celltype"
)
```

```{r}
res_disease_celltype %>%
  filter(padj < 0.1, contrast != "CD - UC") %>%
  count(contrast, celltype, name = "n") %>%
  arrange(desc(n)) %>%
  head(15)
```

Results are written to `analysis_outs/03_activity_inference_model/inference_model_disease_bulk/regressions/`
and visualized in `10_viz_thomas_ibd.Rmd`.
