---
title: "Ligand Activity Regressions: SIG19 (Treg depletion + cytokine blockade)"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
source("regression_helpers.R")
```

# Goal

Test which SIG13 ligand activities respond to in vivo antibody treatment in SIG19
inguinal lymph node CD4 T cells, using the activity scores from
`02_score_disease_datasets.py`.

Treatment groups pair a Treg-perturbing antibody with a cytokine blockade:

| group       | stimulation  |
|-------------|--------------|
| `isotype`   | control      |
| `DTA1`      | TNF.         |
| `GK1.5-IL4` | IL4.         |
| `GK1.5-BGo` | TGF-beta     |
| `DTA1-IL4`  | TNF + IL4    |
| `DTA1-BGo`  | TNF + TGF-b  |

This 2x2 layout supports a **synergy** test per combination: is the double
treatment more than the sum of its single arms,
`(DTA1-IL4 - GK1.5-IL4) - (DTA1 - isotype)`?

**Covariate: `sex`.** Cage is the other candidate -- it is the shared-housing unit,
and it is not aliased with treatment (each cage holds 3-4 treatment groups, each
treatment spans 3-5 cages) -- but with only 23 mice it costs 4 more residual df
(12 vs 16) for very little change in the estimates themselves (r = 0.97 between the
two models' estimates, SEs 1.09x).

**Multiple testing: emmeans' own adjustment**, i.e. across the contrast family within
each activity -- Dunnett for the vs-isotype comparisons, Tukey for the pairwise ones.

# Import

```{r}
SIG19_KEYS <- c("treatment", "mouse", "cage", "sex")
TREATMENTS <- c("isotype", "DTA1", "GK1.5-IL4", "GK1.5-BGo", "DTA1-IL4", "DTA1-BGo")

sig19 <- load_activity_scores("sig19_iln", SIG19_KEYS) %>%
  mutate(treatment = factor(treatment, TREATMENTS), cage = factor(cage), sex = factor(sex))

sig19_celltype <- load_activity_scores("sig19_iln_celltype", c(SIG19_KEYS, "leiden_0.5")) %>%
  rename(leiden = `leiden_0.5`) %>%
  mutate(treatment = factor(treatment, TREATMENTS),
         cage = factor(cage), sex = factor(sex),
         celltype = factor(paste0("cluster_", leiden)))

activities <- activity_names()
cat(nrow(sig19), "mice,", length(activities), "ligand activities\n")
count(sig19, treatment, sex)
```

```{r}
# synergy contrasts, in the order emmeans reports the treatment levels
levs <- levels(sig19$treatment)
synergy_contrasts <- list(
  "DTA1-IL4-synergy" = synergy_contrast(levs, both = "DTA1-IL4",
                                        single_a = "GK1.5-IL4", single_b = "DTA1",
                                        control = "isotype"),
  "DTA1-BGo-synergy" = synergy_contrast(levs, both = "DTA1-BGo",
                                        single_a = "GK1.5-BGo", single_b = "DTA1",
                                        control = "isotype")
)
```

# Treatment associations

## Each treatment vs isotype

```{r}
res_treatment <- fit_and_save(
  sig19,
  rhs = "treatment + sex",
  emm = ~ treatment,
  contrast_args = list(method = "trt.vs.ctrl", ref = "isotype"),
  name = "sig19_iln_treatment",
  padj_source = "emmeans"
)
```

```{r, fig.height=4, fig.width=12}
res_treatment %>%
  mutate(contrast = str_remove(contrast, " - isotype")) %>%
  ggplot(aes(x = estimate, y = -log10(padj))) +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dotted", color = "red") +
  facet_wrap(~contrast, nrow = 1) +
  theme(aspect.ratio = 1) +
  labs(title = "Activity change vs isotype")
```

## All pairwise treatment comparisons

```{r}
res_pairs <- fit_and_save(
  sig19,
  rhs = "treatment + sex",
  emm = ~ treatment,
  contrast_args = list(method = "pairwise"),
  name = "sig19_iln_treatment_pairs",
  padj_source = "emmeans"
)
```

## Combination synergy

The synergy contrasts are user-specified weights, for which emmeans applies no
adjustment at all -- there is no Dunnett or Tukey family to correct within. These two
models therefore keep the folder's default, BH across the 38 activities within each
contrast, so the synergy p-values stay corrected for something. (The original
analysis corrected these with `p.adjust()`'s default, which is holm, not BH.)

```{r}
res_synergy <- fit_and_save(
  sig19,
  rhs = "treatment + sex",
  emm = ~ treatment,
  contrast_args = list(method = synergy_contrasts),
  name = "sig19_iln_treatment_synergy"
)

res_synergy %>%
  filter(padj < 0.1) %>%
  select(contrast, component, estimate, padj) %>%
  arrange(padj)
```

# Celltype associations

## Celltype identity

Which activities distinguish the leiden clusters from one another

```{r}
res_celltype <- fit_and_save(
  sig19_celltype,
  rhs = "celltype + sex",
  emm = ~ celltype,
  contrast_args = list(method = "del.eff"),
  name = "sig19_iln_celltype",
  padj_source = "emmeans"
)
```

```{r}
res_celltype_pairs <- fit_and_save(
  sig19_celltype,
  rhs = "celltype + sex",
  emm = ~ celltype,
  contrast_args = list(method = "pairwise"),
  name = "sig19_iln_celltype_pairs",
  padj_source = "emmeans"
)
```


## Treatment x celltype

```{r}
res_treatment_celltype <- fit_and_save(
  sig19_celltype,
  rhs = "treatment * celltype + sex",
  emm = ~ treatment | celltype,
  contrast_args = list(method = "trt.vs.ctrl", ref = "isotype"),
  padj_by = c("contrast", "celltype"),
  name = "sig19_iln_celltype_treatment",
)
```

## Combination synergy x celltype

```{r}
res_synergy_celltype <- fit_and_save(
  sig19_celltype,
  rhs = "treatment * celltype + sex",
  emm = ~ treatment | celltype,
  contrast_args = list(method = synergy_contrasts),
  padj_by = c("contrast", "celltype"),
  name = "sig19_iln_celltype_treatment_synergy"
)

res_synergy_celltype %>%
  filter(padj < 0.1) %>%
  count(contrast, celltype, name = "n_activities") %>%
  arrange(desc(n_activities))
```

Results are written to `analysis_outs/03_activity_inference_model/inference_model_disease_bulk/regressions/`
and visualized in `11_viz_sig19.Rmd`.
