---
title: "Ligand Activity Figures: Thomas IBD"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
source("viz_helpers.R")
```

# Import

Regression results from `06_regressions_thomas_ibd.Rmd`.

```{r}
res_disease <- read_csv(file.path(RES_IN_DIR, "thomas_ibd_disease_pre.csv"),
                        show_col_types = FALSE)
res_disease_remission <- read_csv(file.path(RES_IN_DIR, "thomas_ibd_disease_remission_pre.csv"),
                                  show_col_types = FALSE)
res_celltype <- read_csv(file.path(RES_IN_DIR, "thomas_ibd_celltype_pre.csv"),
                         show_col_types = FALSE)
res_disease_celltype <- read_csv(file.path(RES_IN_DIR, "thomas_ibd_disease_celltype_pre.csv"),
                                 show_col_types = FALSE)
res_disease_celltype_remission <- read_csv(
  file.path(RES_IN_DIR, "thomas_ibd_disease_celltype_remission_pre.csv"), show_col_types = FALSE)

# "<disease>_<remission>_<timepoint> - Healthy_None_None" -> three columns.
# Non_Remission is hyphenated first so the split on "_" gives three fields.
split_vs_healthy <- function(res) {
  res %>%
    mutate(contrast = str_remove(contrast, " - Healthy_None_None"),
           contrast = str_replace(contrast, "Non_Remission", "Non-Remission")) %>%
    separate_wider_delim(contrast, delim = "_",
                         names = c("Disease", "Remission", "Timepoint"))
}
res_vs_healthy <- read_csv(file.path(RES_IN_DIR, "thomas_ibd_disease_treatment_vsHealthy.csv"),
                           show_col_types = FALSE) %>% split_vs_healthy()
res_vs_healthy_celltype <- read_csv(
  file.path(RES_IN_DIR, "thomas_ibd_disease_treatment_vsHealthy_celltype.csv"),
  show_col_types = FALSE) %>% split_vs_healthy()
```

# Model fit

```{r, fig.height=9, fig.width=9}
(plot_model_fit("thomas_ibd", "whole sample") + plot_model_fit("thomas_ibd_celltype", "per celltype")) /
  (plot_alpha_fit("thomas_ibd", "whole sample") + plot_alpha_fit("thomas_ibd_celltype", "per celltype"))
```

# Disease associations

```{r, fig.height=8, fig.width=12}
plot_volcano(res_disease, facet = "contrast", label = FALSE,
             title = "Pre-treatment disease contrasts") /
  plot_volcano(res_disease, facet = "contrast",
               title = "Pre-treatment disease contrasts (labelled)")

save_plot("ibd_disease_volcano")
```

## By celltype

```{r, fig.height=6, fig.width=5}
res_disease_celltype %>%
  filter(contrast != "CD - UC") %>%
  plot_significance_grid(title = "Activities differing from healthy")

save_plot("ibd_disease_celltype_significance_grid")
```

```{r}
# per-celltype activity heatmap, showing the n most significant activities in that
# celltype together with their constituent single-ligand activities
plot_celltype_activities <- function(res, target, n_top = 10) {
  activities <- top_activities(res, n = n_top, within = tibble(celltype = target)) %>%
    with_component_singles()

  res %>%
    filter(celltype == target) %>%
    mutate(contrast = str_replace_all(contrast, " - ", "_")) %>%
    plot_activity_tiles(activities, title = paste(target, "activity contrasts"))
}
```

```{r, fig.height=20, fig.width=20}
responding_celltypes <- res_disease_celltype %>%
  filter(contrast != "CD - UC", padj < SIG_CUTOFF) %>%
  distinct(celltype) %>%
  pull(celltype)

map(responding_celltypes, ~ plot_celltype_activities(res_disease_celltype, .x)) %>%
  wrap_plots(ncol = 3)
```

```{r, fig.height=12, fig.width=6}
plot_celltype_activities(res_disease_celltype, "CD4 HSPhi CD70pos Treg", n_top = 5) /
  plot_celltype_activities(res_disease_celltype, "Th1", n_top = 5) /
  plot_celltype_activities(res_disease_celltype, "Th1 17 GZMAhi", n_top = 6)

save_plot("ibd_Th1_Th17_pretreatment_disease")
```

# Remission

Baseline (pre-treatment) differences between patients who later did or did not
enter remission. Estimates are inverted so positive means *higher in
non-remission*.

```{r, fig.height=5, fig.width=8}
non_vs_remission <- function(res) {
  res %>%
    filter(contrast %in% c("UC_Remission - UC_Non_Remission",
                           "CD_Remission - CD_Non_Remission")) %>%
    mutate(contrast = str_extract(contrast, "^(UC|CD)"),
           estimate = -estimate)
}

plot_volcano(non_vs_remission(res_disease_remission), facet = "contrast", ymax = 3,
             title = "Non-remission vs remission, pre-treatment")

save_plot("ibd_pretreatment_non_vs_remission_volcano")
```

```{r, fig.height=3.5, fig.width=4}
non_vs_remission(res_disease_celltype_remission) %>%
  filter(estimate > 0) %>%
  plot_significance_grid(title = "Activities raised pre-treatment\nin non-remission")

save_plot("ibd_celltype_non_vs_remission_grid", width = 4, height = 3.5)
```

```{r, fig.height=4, fig.width=16}
celltype_remission <- non_vs_remission(res_disease_celltype_remission)

pmap(list(disease = c("UC", "UC", "UC", "CD"),
          target = c("Th1 17 GZMApos", "CD4 KLF2int T", "CD4 TNFSF13Bhi T","Tph Tfh"),
          ymax = c(3, 3, 3, 3)),
     function(disease, target, ymax) {
       celltype_remission %>%
         filter(contrast == disease, celltype == target) %>%
         plot_volcano(ymax = ymax, title = paste(disease, target))
     }) %>%
  wrap_plots(ncol = 4)

save_plot("ibd_celltype_non_vs_remission")
```

# Treatment

Every disease x remission x timepoint group against healthy controls.

```{r, fig.height=8, fig.width=10}
GROUP_LEVELS <- c("CD_Remission_Pre", "CD_Remission_Post",
                  "CD_Non-Remission_Pre", "CD_Non-Remission_Post",
                  "UC_Remission_Pre", "UC_Remission_Post",
                  "UC_Non-Remission_Pre", "UC_Non-Remission_Post")

label_groups <- function(res) {
  mutate(res, group = factor(paste(Disease, Remission, Timepoint, sep = "_"),
                             GROUP_LEVELS))
}

res_vs_healthy %>%
  label_groups() %>%
  filter(Timepoint == "Pre") %>%
  plot_volcano(facet = "group", title = "Pre-treatment vs healthy")

save_plot("ibd_pretreatment_vs_healthy_volcano")
```

Activities elevated pre-treatment, tracked across treatment: if an activity drops
after treatment in the patients who remit but not in those who do not, it is a
candidate readout of response.

```{r, fig.height=6, fig.width=18}
plot_elevated_activities <- function(target_disease, n_top = 100) {
  df <- res_vs_healthy %>% label_groups() %>% filter(Disease == target_disease)

  elevated <- df %>%
    filter(estimate > 0, padj < SIG_CUTOFF) %>%
    group_by(component) %>%
    slice_max(estimate, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    slice_max(estimate, n = n_top) %>%
    pull(component) %>%
    with_component_singles()

  # order activities by their pre-treatment estimate in the remission arm
  order_by <- df %>%
    filter(group == paste0(target_disease, "_Remission_Pre")) %>%
    annotate_results() %>%
    arrange(desc(estimate)) %>%
    pull(name)

  plot_activity_ranges(df, elevated, group = "group",
                       order_by = order_by,
                       title = paste(target_disease, "activities elevated pre-treatment")) +
    scale_fill_manual(values = c("#33A02C", "#B2DF8A", "#E31A1C", "#FB9A99")) +
    scale_color_manual(values = c("#33A02C", "#B2DF8A", "#E31A1C", "#FB9A99"))
}

(plot_elevated_activities("CD") + theme(legend.position = "none")) +
  plot_elevated_activities("UC") +
  plot_layout(widths = c(2, 1.5))

save_plot("ibd_pretreatment_elevated_ranges", width = 18, height = 6)
```

# Celltype identity

```{r, fig.height=10, fig.width=12}
plot_activity_tiles(res_celltype,
                    activities = top_activities(res_celltype, n = 20),
                    title = "Celltype-distinguishing activities (deleted-effect contrasts)")

save_plot("ibd_celltype_tiles")
```
