---
title: "Ligand Activity Figures: SIG19 (Treg depletion + cytokine blockade)"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
source("viz_helpers.R")
```

# Import

Activity scores and regression results for the SIG19 iLN experiment.

```{r}
SIG19_KEYS <- c("treatment", "mouse", "cage", "sex")
TREATMENTS <- c("isotype", "DTA1", "GK1.5-BGo", "DTA1-BGo", "GK1.5-IL4", "DTA1-IL4")

sig19 <- load_activity_scores("sig19_iln", SIG19_KEYS) %>%
  mutate(treatment = factor(treatment, TREATMENTS))

sig19_celltype <- load_activity_scores("sig19_iln_celltype", c(SIG19_KEYS, "leiden_0.5")) %>%
  rename(leiden = "leiden_0.5") %>%
  mutate(treatment = factor(treatment, TREATMENTS),
         celltype = factor(paste0("cluster_",leiden)))

# contrast labels come back as "<treatment> - isotype", sometimes parenthesized
clean_contrast <- function(res) {
  mutate(res, contrast = str_remove(contrast, " - isotype") %>% str_remove_all("[()]"))
}

res_treatment <- read_csv(file.path(RES_IN_DIR, "sig19_iln_treatment.csv"),
                          show_col_types = FALSE) %>% clean_contrast()
res_synergy <- read_csv(file.path(RES_IN_DIR, "sig19_iln_treatment_synergy.csv"),
                        show_col_types = FALSE)
res_celltype <- read_csv(file.path(RES_IN_DIR, "sig19_iln_celltype.csv"),
                         show_col_types = FALSE) %>%
  mutate(contrast = str_remove(contrast, " effect") %>% str_remove_all("[()]"))
res_celltype_pairs <- read_csv(file.path(RES_IN_DIR, "sig19_iln_celltype_pairs.csv"),
                         show_col_types = FALSE) %>%
  mutate(contrast = str_remove(contrast, " effect") %>% str_remove_all("[()]"))
res_treatment_celltype <- read_csv(file.path(RES_IN_DIR, "sig19_iln_celltype_treatment.csv"),
                                   show_col_types = FALSE) %>% clean_contrast()
res_synergy_celltype <- read_csv(file.path(RES_IN_DIR, "sig19_iln_celltype_treatment_synergy.csv"),
                                 show_col_types = FALSE)

# clusters 8 and 12 are Tregs; excluded from the treatment panels below because
# the DTA1 arm depletes them, so their activity shifts are not comparable
TREG_CLUSTERS <- c("cluster_8", "cluster_12")
TCONV_ORDER <- paste0("cluster_", c(0:7, 9:11))
```

# Model fit

```{r, fig.height=10, fig.width=12}
(plot_model_fit("sig19_iln", "whole iLN") + plot_model_fit("sig19_iln_celltype", "per cluster")) /
  (plot_alpha_fit("sig19_iln", "whole iLN") + plot_alpha_fit("sig19_iln_celltype", "per cluster"))
```

# Treatment associations

```{r, fig.height=4, fig.width=8}
plot_volcano(res_treatment %>% filter(contrast %in% c("DTA1","GK1.5-BGo","DTA1-BGo")), facet = "contrast", title = "Activity change vs isotype")

save_plot("sig19_treatment_bulk_BGo_volcano")
```

```{r, fig.height=4, fig.width=8}
plot_volcano(res_treatment %>% filter(contrast %in% c("DTA1","GK1.5-IL4","DTA1-IL4")), facet = "contrast", title = "Activity change vs isotype")

save_plot("sig19_treatment_bulk_IL4_volcano")
```

```{r, fig.height=4, fig.width=12}
plot_volcano(res_treatment %>% mutate(contrast = factor(contrast, c("isotype","DTA1","GK1.5-IL4","DTA1-IL4","GK1.5-BGo","DTA1-BGo"))), facet = "contrast", title = "Activity change vs isotype", ncol = 5)

save_plot("sig19_treatment_bulk_volcano")
```

Per-mouse activity scores for the activities the antibodies were expected to move.
`IL4_c` and `TNFSF18_c` are this model's names for the single IL4 and TNF-family
consensus activities.

```{r, fig.width=12, fig.height=6}
plot_activity_by_treatment <- function(activity) {
  label <- load_annotations() %>% filter(annotation == activity) %>% pull(name)
  sig19 %>%
    ggplot(aes(x = treatment, y = .data[[activity]], color = treatment)) +
    geom_point(alpha = 0.7, shape = 21, fill = "white") +
    stat_summary(fun = mean, fun.min = mean, fun.max = mean,
                 geom = "crossbar", width = 0.75) +
    scale_color_brewer(palette = "Dark2") +
    #ylim(-3, 4) +
    theme(aspect.ratio = 1, axis.text.x = element_blank()) +
    labs(title = label, x = "", y = "ligand activity")
}

map(c("TGFB1", "IL4_c", "TNFSF18_c", "TGFB1_TNFSF18_c", "IL4_TNFSF18_c"),
    plot_activity_by_treatment) %>%
  wrap_plots(ncol = 3)

save_plot("sig19_treatment_activity_points")
```

# Celltype associations

```{r, fig.height=5, fig.width=6}
res_celltype %>%
  filter(padj < SIG_CUTOFF, !contrast %in% TREG_CLUSTERS) %>%
  filter(component %in% c("TNFSF18_c", "TGFB1", "TGFB1_TNFSF18_c")) %>%
  annotate_results() %>%
  ggplot(aes(x = contrast, y = name)) +
  geom_point(aes(size = -log10(padj), fill = estimate), shape = 21) +
  scale_fill_distiller(palette = "RdBu", limits=c(-2.2,2.2)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  coord_equal() +
  labs(title = "Baseline activity by cluster", x = "", y = "")

save_plot("sig19_tgfb_tnf_celltype_dotplot")
```

```{r, fig.height=5, fig.width=6}
res_celltype %>%
  filter(padj < SIG_CUTOFF, !contrast %in% TREG_CLUSTERS) %>%
  filter(component %in% c("TNFSF18_c", "IL4_c", "IL4_TNFSF18_c")) %>%
  annotate_results() %>%
  ggplot(aes(x = contrast, y = name)) +
  geom_point(aes(size = -log10(padj), fill = estimate), shape = 21) +
  scale_fill_distiller(palette = "RdBu", limits=c(-2.2,2.2)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  coord_equal() +
  labs(title = "Baseline activity by cluster", x = "", y = "")

save_plot("sig19_IL4_tnf_celltype_dotplot")
```

```{r}
res_celltype_pairs %>%
  filter(contrast == "cluster_1 - cluster_3") %>%
  mutate(component_label = ifelse(component %in% c("TGFB1","TGFB1_TNFSF18_c"), component,"")) %>%
  mutate(estimate = -estimate) %>% # flip sign
  ggplot(aes(x = estimate, y = -log10(padj))) +
  geom_point() +
  geom_text_repel(aes(label = component_label)) +
  geom_hline(yintercept = 1, color = "red", linetype = "dashed") +
  xlim(-2.5, 2.5) +
  theme(aspect.ratio = 1) +
  xlab("Estimate (Cluster 3 vs Cluster 1)")

```


```{r}
# treatment effect per cluster, for one antibody arm and its related activities
plot_arm_by_cluster <- function(treatments, activities, limits) {
  res_treatment_celltype %>%
    filter(contrast %in% treatments, !celltype %in% TREG_CLUSTERS,
           component %in% activities) %>%
    annotate_results() %>%
    mutate(cluster = factor(str_remove(celltype, "cluster_"),
                            str_remove(TCONV_ORDER, "cluster_"))) %>%
    ggplot(aes(x = cluster, y = contrast)) +
    geom_tile(aes(fill = estimate)) +
    geom_text(aes(label = sig_stars(padj)), vjust = 0.8, size = 2.5) +
    scale_fill_distiller(palette = "RdBu", limits = limits) +
    facet_wrap(~name, ncol = 1) +
    coord_equal()+
    labs(x = "leiden cluster", y = "")
}
```

```{r, fig.height=4, fig.width=5}
plot_arm_by_cluster(c("GK1.5-BGo", "DTA1-BGo", "DTA1"),
                    c("TGFB1", "TGFB1_TNFSF18_c", "TNFSF18_c"),
                    limits = c(-2, 2))

save_plot("sig19_BGo_arm_by_cluster")
```

```{r, fig.height=4, fig.width=5}
plot_arm_by_cluster(c("GK1.5-IL4", "DTA1-IL4", "DTA1"),
                    c("IL4_c", "IL4_TNFSF18_c", "TNFSF18_c"),
                    limits = c(-5, 5))

save_plot("sig19_IL4_arm_by_cluster")
```

# Activity scores by treatment within a cluster

Per-mouse `TGFB1_TNFSF18_c` activity scores restricted to `cluster_2`, across
treatment groups.

```{r, fig.width=4, fig.height=5}
label <- load_annotations() %>% filter(annotation == "TGFB1_TNFSF18_c") %>% pull(name)

sig19_celltype %>%
  filter(treatment %in% c("isotype","DTA1-BGo")) %>%
  filter(leiden %in% c("0","2","5","1","3")) %>%
  mutate(leiden = factor(leiden, c("0","2","5","1","3"))) %>%
  ggplot(aes(x = leiden, y = TGFB1_TNFSF18_c, color = treatment)) +
  geom_boxplot(outlier.shape = NA) +
  #geom_jitter(width = 0.15, alpha = 0.7) +
  scale_color_brewer(palette = "Dark2") +
  theme(aspect.ratio = 1.5) +
  labs(title = paste(label, "\nacross clusters"), x = "", y = "ligand activity")

save_plot("sig19_TGFB1_TNFSF18_boxplot")
```






































