---
title: "glm_coefficient_concordance_cutoffs"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup}
library(tidyverse)
library(patchwork)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Repo-relative paths: inputs are read from imports_stable/, outputs are written to analysis_outs/
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find the repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "02_combinatorial_screen_signalseq_SIG13")
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
theme_set(theme_Publication())

plot_dir <- file.path(outs_dir, "plots/qc_barcode_cutoff")
out_dir <- file.path(outs_dir, "qc_barcode_cutoff")
# cutoff dataset summaries written by 03_generate_cutoff_datasets.ipynb
in_dir <- file.path(imports_dir, "SIG13/analysis_outs/qc_barcode_cutoff")
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

cutoffs <- 1:10
ref_cutoff <- 7
```

# Import

DSB7 is the cutoff fit used in the manuscript
(`05_interaction_scoring/glmGamPoi_interaction_slurm.r`); the
other cutoffs come from `04_glmGamPoi_interaction_cutoff_slurm.r`.

```{r}
glm_dir <- function(k) {
  if (k == ref_cutoff) file.path(imports_dir, "SIG13/analysis_outs/glmGamPoi")
  else file.path(imports_dir, "SIG13/analysis_outs/glmGamPoi", paste0("glmGamPoi_interaction_DSB", k))
}

load_cutoff <- function(k) {
  singles <- read_csv(file.path(glm_dir(k), "glmGamPoi_singles_lfc_0.1filter.csv"),
                      show_col_types = FALSE) %>%
    mutate(coef_type = if_else(str_ends(single_ligand, "_round1"), "ligand1", "ligand2"))
  inter <- read_csv(file.path(glm_dir(k), "glmGamPoi_interaction_lfc_0.1filter.csv"),
                    show_col_types = FALSE) %>%
    mutate(coef_type = "interaction")

  bind_rows(singles %>% select(interaction, name, coef_type, lfc, adj_pval),
            inter %>% select(interaction, name, coef_type, lfc, adj_pval)) %>%
    mutate(cutoff = k)
}

lfc <- map(cutoffs, load_cutoff) %>% list_rbind()
lfc %>% count(cutoff, coef_type)
```

```{r}
cell_counts <- read_csv(file.path(in_dir, "cutoff_dataset_summary.csv"), show_col_types = FALSE)
cell_counts
```

# Define Consistent Ligands

We first filter for consistent ligands as defined in the manuscript.

```{r}
repCor <- read_csv(file.path(imports_dir, "SIG13/analysis_outs/replicate_corr/replicate_correlation_interactionLfc_0.2filter.csv"),
                   show_col_types = FALSE)

consistent_ligands <- repCor %>%
  filter(pearson_corr > 0.25 & num_shared_genes > 1) %>%
  pull(interaction)

length(consistent_ligands)
```

Conditions are identical across DSB1-9, but DSB10 loses a couple to the
10-cell filter, so combos are also restricted to those present at every
cutoff (written by notebook 3).

```{r}
common_conditions <- read_csv(file.path(in_dir, "cutoff_common_conditions.csv"),
                              show_col_types = FALSE) %>% pull(ligand_call)

lfc <- lfc %>%
  filter(interaction %in% consistent_ligands, interaction %in% common_conditions)

n_distinct(lfc$interaction)
```

# Calculate Concordance

For each ligand combo and coefficient type, correlate per-gene LFCs
between two cutoffs, keeping genes significant in either. The two
single-ligand terms are correlated separately and reported together as
`single`, so each combo contributes two `single` points and one
`interaction` point.

```{r}
concordance <- function(k_a, k_b) {
  inner_join(
    lfc %>% filter(cutoff == k_a) %>% select(interaction, name, coef_type, lfc_a = lfc, padj_a = adj_pval),
    lfc %>% filter(cutoff == k_b) %>% select(interaction, name, coef_type, lfc_b = lfc, padj_b = adj_pval),
    by = c("interaction", "name", "coef_type")
  ) %>%
    drop_na(lfc_a, lfc_b) %>%
    filter(padj_a < 0.1 | padj_b < 0.1) %>%
    group_by(interaction, coef_type) %>%
    filter(n() > 2) %>%
    summarise(
      pearson_corr = cor(lfc_a, lfc_b, method = "pearson"),
      n_genes = n(),
      .groups = "drop"
    ) %>%
    # ligand1 and ligand2 are correlated separately - pooling their genes would mix two
    # coefficients into one Pearson - then reported together as "single"
    mutate(coef_type = if_else(coef_type == "interaction", "interaction", "single"),
           cutoff_a = k_a, cutoff_b = k_b)
}
```

```{r}
vs_ref <- map(setdiff(cutoffs, ref_cutoff), ~ concordance(.x, ref_cutoff)) %>%
  list_rbind() %>%
  mutate(cutoff = factor(cutoff_a, levels = cutoffs))

write_csv(vs_ref, file.path(out_dir, "glm_coefficient_concordance_cutoffs.csv"))
vs_ref
```

```{r}
pairs <- combn(cutoffs, 2, simplify = FALSE)
pairwise <- map(pairs, ~ concordance(.x[1], .x[2])) %>% list_rbind()
write_csv(pairwise, file.path(out_dir, "glm_coefficient_concordance_pairwise.csv"))
```

# Plots

```{r, fig.width=10, fig.height=4}
p_box <- vs_ref %>%
  ggplot(aes(x = cutoff, y = pearson_corr, color = cutoff)) +
  geom_boxplot(width = 0.6, outlier.size = 0.5) +
  ylim(0, 1) +
  facet_wrap(~coef_type) +
  theme(aspect.ratio = 1) +
  scale_color_viridis_d(option = "turbo", end = 0.9) +
  labs(x = "DSB cutoff", y = "Pearson r\n(vs. DSB7)",
       title = sprintf("Reproducible combos only (n=%d)", n_distinct(vs_ref$interaction)))

p_box
ggsave(file.path(plot_dir, "glm_coefficient_concordance_cutoffs.pdf"), p_box)
```

```{r, fig.width=16, fig.height=6}
heat_df <- pairwise %>%
  group_by(coef_type, cutoff_a, cutoff_b) %>%
  summarise(median_corr = median(pearson_corr, na.rm = TRUE), .groups = "drop")

heat_df <- bind_rows(
  heat_df,
  heat_df %>% rename(cutoff_a = cutoff_b, cutoff_b = cutoff_a),
  expand_grid(coef_type = unique(heat_df$coef_type), cutoff_a = cutoffs) %>%
    mutate(cutoff_b = cutoff_a, median_corr = 1)
)

p_heat <- heat_df %>%
  ggplot(aes(x = factor(cutoff_a), y = factor(cutoff_b), fill = median_corr)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", median_corr)), size = 2.5) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  facet_wrap(~coef_type) +
  theme(aspect.ratio = 1) +
  labs(x = "DSB cutoff", y = "DSB cutoff", fill = "median r",
       title = "Median per-combo LFC correlation between cutoffs")

p_heat
ggsave(file.path(plot_dir, "glm_coefficient_concordance_pairwise_heatmap.pdf"), p_heat)
```

Cutoffs differ in cell number as well as cell identity, so concordance
is also plotted against the size of each cutoff's dataset.

```{r, fig.width=10, fig.height=4}
p_n <- vs_ref %>%
  left_join(cell_counts %>% select(cutoff_a = cutoff, n_cells), by = "cutoff_a") %>%
  group_by(cutoff, coef_type, n_cells) %>%
  summarise(median_corr = median(pearson_corr, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = n_cells, y = median_corr, color = cutoff)) +
  geom_point(size = 2) +
  facet_wrap(~coef_type) +
  theme(aspect.ratio = 1) +
  scale_color_viridis_d(option = "turbo", end = 0.9) +
  labs(x = "n cells in cutoff dataset", y = "median Pearson r (vs. DSB7)")

p_n
ggsave(file.path(plot_dir, "glm_concordance_vs_dataset_size.pdf"), p_n)
```

# Summary

```{r}
vs_ref %>%
  group_by(coef_type, cutoff) %>%
  summarise(median_corr = median(pearson_corr, na.rm = TRUE),
            mean_corr = mean(pearson_corr, na.rm = TRUE),
            n_combos = n(),
            .groups = "drop") %>%
  arrange(coef_type, cutoff)
```

```{r}
model <- aov(pearson_corr ~ cutoff, data = vs_ref %>% filter(coef_type == "interaction"))
TukeyHSD(model)
```
