---
title: "SIG13 sPCA program annotation"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
---

```{r setup}
library(tidyverse)
library(patchwork)
library(viridisLite)
library(ggrepel)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(ggpointdensity)
library(scales)
library(viridisLite)
knitr::opts_chunk$set(echo = TRUE)
set.seed(314)
```

```{r}
# Repo-relative paths (inputs: imports_stable/, outputs: analysis_outs/)
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "02_combinatorial_screen_signalseq_SIG13")
dir.create(file.path(outs_dir, "plots"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(outs_dir, "spca"), recursive = TRUE, showWarnings = FALSE)
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
theme_set(theme_Publication())
interaction_palette <- c("#808080",brewer.pal(5,"Dark2"))
names(interaction_palette) <- c("none", "additive over saturation", "buffering", 
                      "buffering complete", "synergy negative", "synergy positive")
```

# Goal

-   remove components correlated with technical factors
-   annotate sparse PCA programs based on ligands
-   annotate interaction effects in sparse PCA programs

# Import

```{r}
# import genes in components
features <- readr::read_csv(file.path(imports_dir, "SIG13/analysis_outs/spca/degs_zscore_allLigands/zscore_degs_allLigands_0.1_alpha1.0_sPCA_components.csv"))
colnames(features)[-1] <- paste0("comp_",colnames(features)[-1])

# import GEP scores
codes <- readr::read_csv(file.path(imports_dir, "SIG13/analysis_outs/spca/degs_zscore_allLigands/zscore_degs_allLigands_0.1_alpha1.0_waggr_score.csv")) 
codes <- codes %>%
  dplyr::mutate(replicate = str_extract(interaction, "rep[12]"))

# import gene-level interactions
interScored <- readr::read_csv(file.path(imports_dir, "SIG13/analysis_outs/glmGamPoi/interactions_scored_v3_glmGamPoi_0.05filter_sig.csv"))
```

I've done testing with a few different alpha values. Ultimately I settled on a value of 1 empirically. At lower alpha values, the number of genes per program is increased and it dilutes out genes that have true interaction effects. At higher numbers some programs disappear (e.g. CCL21A driven program)

# Component Gene Numbers

```{r, fig.height=4, fig.width=12}
compNum <- features[,-1]
compNum <- tibble(components = names(compNum),
                  counts = colSums(compNum != 0))

ggplot(compNum, aes(x = components, y = counts)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

# LM for sPCA components

To connect sparse PCA programs with specific ligands and technical factors, I will perform linear regression on scaled sparse PCA component scores with the following structure: scores = ligand1 + ligand2 + ligand1:ligand2 + replicate. Linker only will be used as the reference level (aka intercept).

P-value adjustment is performed across each globally.

I will perform lm with a robust covariance model given the lm assumption violations.

```{r}
library(performance)
library(broom)
library(sandwich)   # robust vcov
library(MASS)       # rlm
library(quantreg)   # rq
```

```{r}
# center/scale and clean codes
mat <- codes %>%
  dplyr::mutate(across(where(is.numeric), ~ as.numeric(scale(.x)))) %>%
  column_to_rownames("interaction") %>%
  dplyr::select(where(is.numeric)) %>%
  as.matrix()

# Extract metadata from rownames
meta <- tibble(interaction = rownames(mat)) %>%
  separate(interaction, into = c("ligand1", "ligand2", "replicate"), sep = "_", remove = FALSE) %>%
  mutate(
    replicate = factor(replicate),
    ligand1   = fct_relevel(ligand1, "linker"),
    ligand2   = fct_relevel(ligand2, "linker")
  )

# Long format
df <- as.data.frame(mat) %>%
  rownames_to_column("interaction") %>%
  left_join(meta, by = "interaction") %>%
  mutate(interaction = gsub("_rep[12]","",interaction)) %>%
  mutate(interaction = fct_relevel(interaction, "linker_linker")) %>%
  pivot_longer(cols = starts_with("comp"), names_to = "component", values_to = "loading")
```

## Interaction Regression with Robust SE

Due to the OLS assumption violations, I will redo the regression using robust standard errors.

```{r}
# Ensure factors
df <- df %>%
  mutate(
    ligand1    = factor(ligand1),
    ligand2    = factor(ligand2),
    interaction= factor(interaction),
    replicate  = factor(replicate)
  )

# ---- Robust covariance helpers ----
# Robust SE
.robust_vcov <- function(mod) sandwich::vcovHC(mod, type = "HC3")
.tidy_ols_hc <- function(mod) broom::tidy(mod, conf.int = TRUE, vcov = .robust_vcov(mod))

# Robust estimation (Huber) — coefficients may shift if outliers matter
.fit_rlm <- function(d, fml) MASS::rlm(fml, data = d, psi = psi.huber)
.tidy_rlm <- function(mod) broom::tidy(mod)  # treat as sensitivity (SEs are large-sample)

# =========================
# Model 1: loading ~ ligand1 * ligand2 + replicate
# =========================
interaction_models <- df %>%
  group_by(component) %>%
  nest() %>%
  mutate(
    model_ols   = map(data, ~ lm(loading ~ ligand1 * ligand2 + replicate, data = .x)),
    tidy_ols_hc = map(model_ols, .tidy_ols_hc),
    glance_ols  = map(model_ols, glance),
    augment_ols = map(model_ols, augment),

    model_rlm   = map(data, ~ .fit_rlm(.x, loading ~ ligand1 * ligand2 + replicate)),
    tidy_rlm    = map(model_rlm, .tidy_rlm)
  )

# OLS+HC coefficient table with BH adjust globally
lm_interaction <- interaction_models %>%
  dplyr::select(component, tidy_ols_hc) %>%
  unnest(tidy_ols_hc) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p.value, method = "BH"))

rlm_interaction <- interaction_models %>%
  dplyr::select(component, tidy_rlm) %>%
  unnest(tidy_rlm)
```

### Compare RLM to LM

```{r}
joint_interaction <- left_join(lm_interaction, rlm_interaction,
                               by = c("component","term"),
                               suffix = c("_lm","_rlm"))
```

```{r}
joint_interaction %>%
  ggplot(aes(x = estimate_lm, y = estimate_rlm)) +
    geom_point()
```

There's a near perfect correlation of robust LM estimates and OLS estimates, suggesting that outliers are not strongly driving any particular component

## Condition Regression with Robust SE

Due to the OLS assumption violations, I will redo the regression using robust standard errors.

```{r}
# Ensure factors
df <- df %>%
  mutate(
    ligand1    = factor(ligand1),
    ligand2    = factor(ligand2),
    interaction= factor(interaction),
    replicate  = factor(replicate)
  )

# ---- Robust covariance helpers ----
# Robust SE
.robust_vcov <- function(mod) sandwich::vcovHC(mod, type = "HC3")
.tidy_ols_hc <- function(mod) broom::tidy(mod, conf.int = TRUE, vcov = .robust_vcov(mod))

# =========================
# Model 2: loading ~ interaction + replicate
# =========================
condition_models <- df %>%
  group_by(component) %>%
  nest() %>%
  mutate(
    model_ols   = map(data, ~ lm(loading ~ interaction + replicate, data = .x)),
    tidy_ols_hc = map(model_ols, .tidy_ols_hc),
    glance_ols  = map(model_ols, glance),
    augment_ols = map(model_ols, augment)
  )

# OLS+HC coefficient table with BH adjust globally
lm_condition <- condition_models %>%
  dplyr::select(component, tidy_ols_hc) %>%
  unnest(tidy_ols_hc) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p.value, method = "BH"))
```

## Identify replicate-driven components

Here, I want to identify and remove components that are primarily driven by technical factors. This will specifically manifest as those that are highly associated with technical replicates

### Visualize replicate associations

```{r, fig.width=14, fig.height=5}
lm_interaction %>%
  filter(p_adj < 0.01) %>%
  mutate(is_replicate_association = grepl("replicate", term)) %>%
  ggplot(aes(x = component, y = -log10(p_adj), color = is_replicate_association, size = abs(estimate))) +
  geom_point() +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))
```

```{r, fig.width=16, fig.height=4}
lm_interaction %>%
  filter(p_adj < 0.01) %>%
  mutate(is_replicate_association = grepl("replicate", term)) %>%
  ggplot(aes(x = component, y = -log10(p_adj), color = is_replicate_association)) +
  geom_point() +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust=0.5))

ggsave(file.path(outs_dir, "plots/spca_replicate_association_dotplot.pdf"))
```

### Filter replicate-driven componets

I will take a conservative approach and remove any component where the most significant association is replicate

```{r}
bad_rep_comps <- lm_interaction %>%
  group_by(component) %>%
  slice_min(p_adj, with_ties = FALSE) %>%
  filter(grepl("replicate", term)) %>%
  pull(component)

bad_rep_comps
```

```{r}
# make filtered lm tibble
lm_interaction_filtered <- lm_interaction %>%
  filter(!(component %in% bad_rep_comps))

lm_condition_filtered <- lm_condition %>%
  filter(!(component %in% bad_rep_comps))
```

```{r}
# count number of components remaining
lm_interaction_filtered %>%
  pull(component) %>%
  unique() %>%
  length()
```

### Visualize Filtered Correlation

```{r, fig.width=8, fig.height=5}
repDiff <- codes  %>%
  dplyr::mutate(interaction = str_remove(interaction, "_rep[12]")) %>%
  # scale components
  pivot_longer(-c(interaction, replicate), names_to = "component", values_to = "codes_scaled") %>%
  pivot_wider(names_from = replicate, values_from = codes_scaled) %>%
  dplyr::mutate(rep_diff = abs(rep1 - rep2))

repDiffFiltered <- codes  %>%
  dplyr::select(-any_of(bad_rep_comps)) %>%
  dplyr::mutate(interaction = str_remove(interaction, "_rep[12]")) %>%
  # scale components
  pivot_longer(-c(interaction, replicate), names_to = "component", values_to = "codes_scaled") %>%
  pivot_wider(names_from = replicate, values_from = codes_scaled) %>%
  dplyr::mutate(rep_diff = abs(rep1 - rep2))

p1 <- ggplot(repDiff, aes(x = rep1, y = rep2)) +
  geom_point(alpha=0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme(aspect.ratio = 1) +
  xlab("replicate 1") +
  ylab("replicate 2") +
  ggtitle("Inter-replicate correlation\nsPCA component\nscores")

p2 <- ggplot(repDiffFiltered, aes(x = rep1, y = rep2)) +
  geom_point(alpha=0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme(aspect.ratio = 1) +
  xlab("replicate 1") +
  ylab("replicate 2") +
  ggtitle("Inter-replicate correlation\nFiltered sPCA component\nscores")

p1+p2
ggsave(file.path(outs_dir, "plots/spca_replicate_corr.pdf"))

cor(repDiff$rep1, repDiff$rep2)
cor(repDiffFiltered$rep1, repDiffFiltered$rep2)
```

## Classify interaction effects

Here, I will take a similar approach as in the gene-level analysis and identify synergy, buffering and purely additive effects

```{r}
# make formated tibble
l1 <- lm_interaction_filtered %>% 
  filter(grepl("^ligand1", term)) %>% 
  transmute(component,
            ligand1 = sub("^ligand1", "", term),
            estimate_ligand1 = estimate,
            p_adj_ligand1 = p_adj)
l2 <- lm_interaction_filtered %>% 
  filter(grepl("^ligand2", term)) %>% 
  transmute(component,
            ligand2 = sub("^ligand2", "", term),
            estimate_ligand2 = estimate,
            p_adj_ligand2 = p_adj)
int <- lm_interaction_filtered %>% 
  filter(grepl("^ligand1.+:ligand2", term)) %>%
  separate(term, c("l1","l2"), sep = ":", remove = FALSE) %>%
  transmute(component,
            ligand1 = sub("^ligand1", "", l1),
            ligand2 = sub("^ligand2", "", l2),
            interaction = paste0(ligand1, "_", ligand2),
            estimate_interaction = estimate,
            p_adj_interaction = p_adj)

# assemble
out <- int %>%
  left_join(l1, by = c("component","ligand1"), relationship = "many-to-many") %>%
  left_join(l2, by = c("component","ligand2"), relationship = "many-to-many") %>%
  arrange(component, ligand1, ligand2)
```

```{r}
lm_scored <- out %>%
  mutate(estimate_total_interaction_model = (estimate_ligand1+estimate_ligand2+estimate_interaction),
         interaction_score = case_when(p_adj_interaction > 0.1 ~ 0,
                                       p_adj_interaction <= 0.1 ~ (estimate_interaction/(estimate_total_interaction_model)))) %>%
    mutate(interaction_class = case_when(
        # conditions are evaluated sequentially
        p_adj_interaction <= 0.1 & estimate_total_interaction_model > 0 & interaction_score > 0 ~ "synergy positive",
        p_adj_interaction <= 0.1 & estimate_total_interaction_model < 0 & interaction_score > 0 ~ "synergy negative",
        p_adj_interaction <= 0.1 & interaction_score < 0 ~ "buffering",
        p_adj_interaction > 0.1 ~ "none"
        ))
```

### Merge with gene-level analysis

Here I will merge the program features with the gene-level analysis of interaction effects which is more reliable

```{r}
spca_genes <- features %>%
  pivot_longer(-gene, names_to = "spca_component", values_to = "loading") %>%
  filter(loading != 0) %>%
  arrange(spca_component,desc(loading))

interScored_minimal <- interScored %>%
  dplyr::select(name, interaction, interaction_class) %>%
  mutate(interaction = gsub("\\.","_", interaction)) %>%
  dplyr::rename(interaction_class_gene = interaction_class)

# create annotated lm_scored with proportion of synergy and buffering
# genes out of total for every component and interaction
lm_scored_anno <- lm_scored %>%
  left_join(spca_genes, by = c("component" = "spca_component"), relationship = "many-to-many") %>%
  left_join(interScored_minimal, by = c("gene" = "name","interaction")) %>%
  group_by(across(-c(gene, loading, interaction_class_gene))) %>%
  count(interaction_class_gene, name = "n") %>%
  mutate(prop = n / sum(n))  %>%
  ungroup() %>%
  filter(interaction_class_gene != "none") %>%
  pivot_wider(id_cols = -n,
              names_from = interaction_class_gene,
              names_prefix = "freq_",
              values_from = prop,
              values_fill = 0)
```

# Export

```{r}
# export loadings of components
features %>%
  pivot_longer(-gene, names_to = "spca_component", values_to = "loading") %>%
  filter(loading != 0) %>%
  arrange(spca_component,desc(loading)) %>%
  write_csv(file.path(outs_dir, "spca/zscore_degs_allLigands_0.1_alpha1.0_sPCA_loadings.csv"))
```

```{r}
compNum %>%
  write_csv(file.path(outs_dir, "spca/zscore_degs_allLigands_0.1_alpha1.0_sPCA_component_num.csv"))
```

```{r}
write_csv(lm_interaction, file.path(outs_dir, "spca/lm_fit_zscore_degs_allLigands_0.1_alpha1.0_sPCA.csv"))
write_csv(lm_interaction_filtered, file.path(outs_dir, "spca/lm_fit_zscore_degs_allLigands_0.1_alpha1.0_sPCA_clean.csv"))
write_csv(lm_scored_anno, file.path(outs_dir, "spca/lm_scored_zscore_degs_allLigands_0.1_alpha1.0_sPCA_clean.csv"))
```

```{r}
write_csv(lm_condition, file.path(outs_dir, "spca/lm_fit_condition_zscore_degs_allLigands_0.1_alpha1.0_sPCA.csv"))
write_csv(lm_condition_filtered, file.path(outs_dir, "spca/lm_fit_condition_zscore_degs_allLigands_0.1_alpha1.0_sPCA_clean.csv"))
```






