---
title: "sPCA specific program visualization"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
---

```{r setup}
library(tidyverse)
library(patchwork)
library(viridisLite)
library(ggrepel)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(ggpointdensity)
library(scales)
library(viridisLite)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Repo-relative paths (inputs: imports_stable/, outputs: analysis_outs/)
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "02_combinatorial_screen_signalseq_SIG13")
dir.create(file.path(outs_dir, "plots"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(outs_dir, "spca"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(outs_dir, "plots/spca_gene_hm"), recursive = TRUE, showWarnings = FALSE)
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
theme_set(theme_Publication())
interaction_palette <- c("#808080",brewer.pal(5,"Dark2"))
names(interaction_palette) <- c("none", "additive over saturation", "buffering", 
                      "buffering complete", "synergy negative", "synergy positive")
```

# Goal

Visualize select annotated sparse PCA components for SIG13

# Import

```{r}
# import feature scores
features <- readr::read_csv(file.path(imports_dir, "SIG13/analysis_outs/spca/degs_zscore_allLigands/zscore_degs_allLigands_0.1_alpha1.0_sPCA_components.csv"))
colnames(features)[-1] <- paste0("comp_",colnames(features)[-1])

# import ligand component values
codes <- readr::read_csv(file.path(imports_dir, "SIG13/analysis_outs/spca/degs_zscore_allLigands/zscore_degs_allLigands_0.1_alpha1.0_waggr_score.csv")) 
codes <- codes %>%
  dplyr::mutate(replicate = str_extract(interaction, "rep[12]"))

# import lm results
lm_scored <- read_csv(file.path(imports_dir, "SIG13/analysis_outs/spca/lm_scored_zscore_degs_allLigands_0.1_alpha1.0_sPCA_clean.csv"))
lm_interaction <- read_csv(file.path(imports_dir, "SIG13/analysis_outs/spca/lm_fit_zscore_degs_allLigands_0.1_alpha1.0_sPCA.csv"))

# define replicate-driven comps
bad_rep_comps <- lm_interaction$component[!(lm_interaction$component %in% lm_scored$component)] %>% unique()
```

# Component Gene Numbers

```{r, fig.height=4, fig.width=12}
compNum <- features[,-1]
compNum <- tibble(components = names(compNum),
                  counts = colSums(compNum != 0))
ggplot(compNum, aes(x = components, y = counts)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

# Visualize interaction effects

```{r, fig.height=10, fig.width=8}
lm_scored %>%
  filter(p_adj_interaction < 0.1 | p_adj_ligand1 < 0.1 | p_adj_ligand2 < 0.1) %>%
  mutate(interaction_class = factor(interaction_class, levels = names(interaction_palette))) %>%
  count(component, interaction_class, name = "n") %>%
  group_by(component) %>%
  mutate(freq = n / sum(n)) %>%
  ungroup() %>%
  filter(interaction_class != "none") %>%
  ggplot(aes(x = n, y = component, fill = interaction_class)) +
  geom_col() +
  scale_fill_manual(values = interaction_palette, breaks = names(interaction_palette))
```

# Heatmap Subsets

```{r}
lm_interaction_res <- lm_interaction %>%
  filter(component %in% lm_scored$component) %>% # remove replicate associated comps
  filter(grepl("^ligand1.*:ligand2.*$", term)) %>%
  mutate(term = gsub("ligand[12]","",term)) %>%
  mutate(term = gsub(":","_",term)) %>%
  mutate(p_score = -log10(p_adj)) %>%
  rename(interaction = term)

lm_single_res <- lm_interaction %>%
  filter(component %in% lm_scored$component) %>% # remove replicate associated comps
  filter(!grepl("^ligand1.*:ligand2.*$", term)) %>%
  filter(grepl("ligand",term)) %>%
  mutate(p_score = -log10(p_adj))

codesMean <- codes %>%
      dplyr::mutate(interaction = gsub("_rep[12]","",interaction)) %>%
      group_by(interaction) %>%
      summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))
```

# Combination Single

```{r, fig.width=12, fig.height=4}
df <- codes %>%
  dplyr::select(-any_of(bad_rep_comps)) %>%
  mutate(interaction = gsub("_rep[12]", "", interaction)) %>%
  group_by(interaction) %>%
  summarise(across(where(is.numeric), ~mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  mutate(across(where(is.numeric), ~as.numeric(scale(.x)))) %>%
  pivot_longer(-interaction, names_to = "component", values_to = "scaled_score") %>%
  mutate(is_single = ifelse(grepl("linker", interaction),"yes","no")) %>%
  left_join(lm_scored, by = c("interaction", "component")) %>%
  mutate(filter_col = case_when(is_single == "no" & interaction_class == "none" ~ "remove",
                                TRUE ~ "keep")) %>%
  #filter(filter_col != "remove") %>%
  group_by(component, is_single) %>%
  slice_max(order_by = scaled_score, n=3) %>%
  # Calculate difference (yes - no) and reorder component factor
  group_by(component) %>%
  mutate(diff_val = mean(scaled_score[is_single == "yes"]) - mean(scaled_score[is_single == "no"])) %>%
  ungroup() %>%
  mutate(component = reorder(component, diff_val))

single_v_comb_plot <- df %>%
  ggplot(aes(x = component, y = scaled_score, color = is_single)) +
  geom_point(alpha=0.5) +
  scale_color_manual(values = c("red","black")) +
  theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5)) +
  ggtitle("Distribution scaled sPCA score for top 3 conditions\nsingle vs combinatorial")
single_v_comb_plot
```

```{r}
ggsave(file.path(outs_dir, "plots/spca_alpha1.0_single_v_comb_plot.pdf"), single_v_comb_plot, width = 12, height = 4)
write_csv(df, file.path(outs_dir, "spca/zscore_degs_allLigands_0.1_alpha1.0_sPCA_single_comb_top3.csv"))
```

# sPCA HM Visualization

## Full Heatmap

```{r, fig.height=10, fig.width=10}
# scale and clean codes
mat <- codes %>%
  dplyr::select(-any_of(bad_rep_comps))  %>%
  dplyr::mutate(interaction = gsub("_rep[12]","",interaction)) %>%
  group_by(interaction) %>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) %>%
  dplyr::mutate(across(where(is.numeric), ~ as.numeric(scale(.x, center=F)))) %>%
  column_to_rownames("interaction") %>%
  as.matrix()

# extract ligand1 from rownames (before transposing!)
interaction_names <- rownames(mat)
ligand1_vec <- sapply(strsplit(interaction_names, "_"), `[`, 1)

# build column annotation (applies to columns *after* transpose)
col_anno <- HeatmapAnnotation(
  ligand1 = ligand1_vec,
  col = list(ligand1 = setNames(brewer.pal(length(unique(ligand1_vec)), "Set3"), unique(ligand1_vec)))
)

# transpose
mat <- t(mat)

col_fun <- colorRamp2(
  seq(-6, 6, length.out = 11),
  rev(brewer.pal(n = 11, name = "RdBu"))
)

# Plot
hm_full <- Heatmap(
  mat,
  name = "scaled sPCA loadings",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = FALSE,
  top_annotation = col_anno,
  column_title = "sPCA components across interactions\nmean across replicates",
  heatmap_legend_param = list(
    title = "scaled sPCA\nloadings",
    at = seq(-5, 5, by = 3),
    color_bar = "continuous"
  )
)
draw(hm_full)
```

## Single Ligands + Interaction Bar Plots

```{r, fig.height=12, fig.width=9}
class_levels <- setdiff(names(interaction_palette), "none")

# pull out significant ligands
sig_ligands <- {
  x <- lm_single_res %>% filter(p_adj < 0.05) %>% pull(term) %>% unique()
  x <- sub("^ligand1(.+)$", "\\1_linker", x)
  sub("^ligand2(.+)$", "linker_\\1", x)
}
sig_ligands <- sig_ligands[!(sig_ligands %in% c("linker_TNF","linker_IL2","linker_IL6","linker_IL27","linker_IL4","linker_IL21","linker_IFNG","linker_IL7"))]

# make heatmap matrix
mat <- codes %>%
  dplyr::select(-any_of(bad_rep_comps)) %>%
  mutate(interaction = gsub("_rep[12]", "", interaction)) %>%
  group_by(interaction) %>%
  summarise(across(where(is.numeric), ~mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  # scale before filtering for single ligands
  mutate(across(where(is.numeric), ~as.numeric(scale(.x)))) %>%
  filter(interaction %in% sig_ligands) %>%
  mutate(interaction = gsub("^linker_|_linker","",interaction)) %>%
  column_to_rownames("interaction") %>% as.matrix() %>% t()

# make heatmap color scale
col_fun <- colorRamp2(seq(-6, 6, length.out = 11), rev(brewer.pal(11, "RdBu")))

# make bar plot
bar_mat <- lm_scored %>%
  filter(p_adj_interaction < 0.05 | p_adj_ligand1 < 0.05 | p_adj_ligand2 < 0.05) %>%
  filter(interaction_class != "buffering") %>%
  #filter(interaction_class %in% class_levels) %>%
  mutate(interaction_class = factor(interaction_class, levels = class_levels)) %>%
  count(component, interaction_class, name = "n") %>%
  complete(component, interaction_class, fill = list(n = 0)) %>%
  pivot_wider(names_from = interaction_class, values_from = n, names_expand = TRUE, values_fill = 0) %>%
  column_to_rownames("component") %>% as.matrix()

common <- intersect(rownames(mat), rownames(bar_mat))
mat <- mat[common, , drop = FALSE]
bar_mat <- bar_mat[common, class_levels, drop = FALSE]

ht <- Heatmap(mat, name = "scaled sPCA loadings",
              col = col_fun,
              cluster_rows = TRUE, cluster_columns = TRUE,
              row_names_side = "left",
              border_gp = gpar(col = "black"),
              heatmap_legend_param = list(title = "scaled\nsPCA\nloadings",
                                          at = seq(-6, 6, by = 3), color_bar = "continuous"))

bars <- rowAnnotation(
  num_interactions = anno_barplot(bar_mat, gp = gpar(fill = interaction_palette[class_levels]),
                       axis_param = list(side = "bottom")),
  width = unit(45, "mm")
)

bar_legend <- Legend(title = "interaction class", at = class_levels,
                     legend_gp = gpar(fill = interaction_palette[class_levels]))

# Assign the draw result to an object
ht_obj <- draw(ht + bars, annotation_legend_list = list(bar_legend),
      heatmap_legend_side = "right", annotation_legend_side = "right")

# Extract the components in their clustered order
clustered_rows <- tibble(component = rownames(mat)[row_order(ht_obj)])
```

## Consistent Ligands + Interactions Bar

`consistent_ligands` is the repo-wide reproducibility filter used elsewhere in this repo (e.g.
`analysis/qc_barcode_counts/02_glm_coefficient_concordance_viz.Rmd`): ligand-pair combos whose
interaction LFCs correlate across biological replicates.

```{r}
repCor <- readr::read_csv(file.path(imports_dir, "SIG13/analysis_outs/replicate_corr/replicate_correlation_interactionLfc_0.2filter.csv"))

consistent_ligands <- repCor %>%
  filter(pearson_corr > 0.25 & num_shared_genes > 1) %>%
  pull(interaction)

length(consistent_ligands)
```

```{r, fig.height=12, fig.width=10}
# scale and clean codes, restricted to reproducible interactions
mat <- codes %>%
  dplyr::select(-any_of(bad_rep_comps))  %>%
  dplyr::mutate(interaction = gsub("_rep[12]","",interaction)) %>%
  filter(interaction %in% consistent_ligands) %>%
  group_by(interaction) %>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) %>%
  dplyr::mutate(across(where(is.numeric), ~ as.numeric(scale(.x, center=F)))) %>%
  column_to_rownames("interaction") %>%
  as.matrix()

# extract ligand1 from rownames (before transposing!)
interaction_names <- rownames(mat)
ligand1_vec <- sapply(strsplit(interaction_names, "_"), `[`, 1)

# build column annotation (applies to columns *after* transpose)
col_anno <- HeatmapAnnotation(
  ligand1 = ligand1_vec,
  col = list(ligand1 = setNames(brewer.pal(length(unique(ligand1_vec)), "Set3"), unique(ligand1_vec)))
)

# transpose
mat <- t(mat)

col_fun <- colorRamp2(
  seq(-4, 4, length.out = 11),
  rev(brewer.pal(n = 11, name = "RdBu"))
)

# make bar plot of synergistic interaction class counts, restricted to reproducible interactions
bar_mat_reproducible <- lm_scored %>%
  filter(interaction %in% consistent_ligands) %>%
  filter(p_adj_interaction < 0.1 | p_adj_ligand1 < 0.1 | p_adj_ligand2 < 0.1) %>%
  filter(interaction_class != "buffering") %>%
  mutate(interaction_class = factor(interaction_class, levels = class_levels)) %>%
  count(component, interaction_class, name = "n") %>%
  complete(component, interaction_class, fill = list(n = 0)) %>%
  pivot_wider(names_from = interaction_class, values_from = n, names_expand = TRUE, values_fill = 0) %>%
  column_to_rownames("component") %>% as.matrix()

common <- intersect(rownames(mat), rownames(bar_mat_reproducible))
mat <- mat[common, , drop = FALSE]
bar_mat_reproducible <- bar_mat_reproducible[common, class_levels, drop = FALSE]

bars_reproducible <- rowAnnotation(
  num_interactions = anno_barplot(bar_mat_reproducible, gp = gpar(fill = interaction_palette[class_levels]),
                       axis_param = list(side = "bottom")),
  width = unit(45, "mm")
)

# Plot
hm_full_reproducible <- Heatmap(
  mat,
  name = "scaled sPCA loadings",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  row_names_side = "left",
  show_column_names = FALSE,
  top_annotation = col_anno,
  border_gp = gpar(col = "black"),
  column_title = "sPCA components across reproducible interactions\nmean across replicates",
  heatmap_legend_param = list(
    title = "scaled sPCA\nloadings",
    at = seq(-4, 4, by = 2),
    color_bar = "continuous"
  )
)
ht_obj_reproducible <- draw(hm_full_reproducible + bars_reproducible, annotation_legend_list = list(bar_legend),
     heatmap_legend_side = "right", annotation_legend_side = "right")
```

## Combined Single + Consistent Ligand Heatmap

Combines the single-ligand heatmap (left, with column labels), the consistent
(reproducible) interaction heatmap (middle, no column labels since there are
many combos), and the interaction-effect bar plot (right), restricted to the
components shared across all three panels.

```{r, fig.height=12, fig.width=20}
# single-ligand matrix: components (rows) x single ligands (columns, labeled)
mat_single <- codes %>%
  dplyr::select(-any_of(bad_rep_comps)) %>%
  mutate(interaction = gsub("_rep[12]", "", interaction)) %>%
  group_by(interaction) %>%
  summarise(across(where(is.numeric), ~mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  mutate(across(where(is.numeric), ~as.numeric(scale(.x)))) %>%
  filter(interaction %in% sig_ligands) %>%
  mutate(interaction = gsub("^linker_|_linker","",interaction)) %>%
  column_to_rownames("interaction") %>% as.matrix() %>% t()

# consistent-ligand matrix: components (rows) x reproducible interactions (columns, unlabeled)
mat_consistent <- codes %>%
  dplyr::select(-any_of(bad_rep_comps)) %>%
  dplyr::mutate(interaction = gsub("_rep[12]","",interaction)) %>%
  filter(interaction %in% consistent_ligands) %>%
  group_by(interaction) %>%
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))

mat_consistent <- mat_consistent %>%
  dplyr::mutate(across(where(is.numeric), ~ as.numeric(scale(.x, center = FALSE)))) %>%
  column_to_rownames("interaction") %>%
  as.matrix() %>% t()

# restrict all three panels to components shared across the single-ligand
# heatmap, the consistent-ligand heatmap, and the reproducible interaction-effect bar plot
common_combined <- Reduce(intersect, list(rownames(mat_single), rownames(mat_consistent), rownames(bar_mat_reproducible)))

mat_single_c <- mat_single[common_combined, , drop = FALSE]
mat_consistent_c <- mat_consistent[common_combined, , drop = FALSE]
bar_mat_combined <- bar_mat_reproducible[common_combined, class_levels, drop = FALSE]

col_fun_single <- colorRamp2(seq(-4, 4, length.out = 11), rev(brewer.pal(11, "RdBu")))
col_fun_consistent <- colorRamp2(seq(-4, 4, length.out = 11), rev(brewer.pal(11, "RdBu")))

ht_consistent_combined <- Heatmap(
  mat_consistent_c,
  name = "consistent interaction\nscaled sPCA loadings",
  col = col_fun_consistent,
  cluster_rows = TRUE, 
  cluster_columns = TRUE,
  show_row_names = TRUE,
  row_names_side = "left",
  show_column_names = FALSE,
  border_gp = gpar(col = "black"),
  column_title = "consistent interactions",
  heatmap_legend_param = list(title = "consistent\nscaled sPCA\nloadings", at = seq(-4, 4, by = 2), color_bar = "continuous")
)

ht_single_combined <- Heatmap(
  mat_single_c,
  name = "single ligand\nscaled sPCA loadings",
  col = col_fun_single,
  cluster_rows = FALSE, # follow row order from ht_consistent_combined's clustering
  cluster_columns = TRUE,
  show_row_names = FALSE,
  show_column_names = TRUE,
  border_gp = gpar(col = "black"),
  column_title = "single ligands",
  heatmap_legend_param = list(title = "single ligand\nscaled sPCA\nloadings", at = seq(-4, 4, by = 2), color_bar = "continuous")
)

bars_combined <- rowAnnotation(
  num_interactions = anno_barplot(bar_mat_combined, gp = gpar(fill = interaction_palette[class_levels]),
                       axis_param = list(side = "bottom")),
  width = unit(45, "mm")
)

draw(ht_consistent_combined + ht_single_combined + bars_combined,
     annotation_legend_list = list(bar_legend),
     heatmap_legend_side = "right", annotation_legend_side = "right")

# extract the components in their clustered order
clustered_rows_reproducible <- tibble(component = rownames(mat_consistent)[row_order(ht_consistent_combined)])
```

## Export

```{r}
pdf(file.path(outs_dir, "plots/spca_alpha1.0_singleLigand_interactions_hm.pdf"), width = 9, height = 12)
draw(ht + bars, annotation_legend_list = list(bar_legend),
     heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()
```

```{r}
pdf(file.path(outs_dir, "plots/spca_alpha1.0_full_hm.pdf"), width = 10, height = 8)
draw(hm_full)
dev.off()
```

```{r}
pdf(file.path(outs_dir, "plots/spca_alpha1.0_full_hm_reproducible.pdf"), width = 10, height = 12)
draw(hm_full_reproducible + bars_reproducible, annotation_legend_list = list(bar_legend),
     heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()
```

```{r}
pdf(file.path(outs_dir, "plots/spca_alpha1.0_single_consistent_combined_hm.pdf"), width = 20, height = 12)
draw(ht_consistent_combined + ht_single_combined + bars_combined,
     annotation_legend_list = list(bar_legend),
     heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()
```

```{r}
write_csv(clustered_rows, file.path(outs_dir, "spca/spca_alpha1.0_full_hm_row_order.csv"))
write_csv(clustered_rows_reproducible, file.path(outs_dir, "spca/spca_alpha1.0_full_hm_reproducible_row_order.csv"))
```

# TNF Interaction Heatmaps

## Define Synergy Comps

```{r}
synergy_pos_comps <- lm_scored %>%
  filter(interaction_class == "synergy positive") %>%
  filter(`freq_synergy positive` > 0.05) %>%
  filter(p_adj_interaction < 0.001) %>%
  group_by(interaction) %>%
  slice_min(order_by = p_adj_interaction, n = 1)
```

## Visualize

```{r, fig.width=12, fig.height=3.5}
# Define the specific order for the categories
cat_levels <- c("IL2_TNF", "IL4_TNF", "IL7_TNF", "IL6_TNF", "IL21_TNF", 
                "IL27_TNF", "IFNG_TNF", "TGFB1_TNF", "TNF_TNF", "single")

# subset and apply factor order
pattern <- "^(IL2|IL4|IL6|IL7|IL21|IL27|TGFB1|IFNG|TNF)_(TNF.*|linker)|^linker_TNF.*"

# define comps
comps <- synergy_pos_comps %>%
  filter(grepl("^(IL2|IL4|IL6|IL7|IL21|IL27|TGFB1|IFNG|TNF)_(TNF.*)", interaction)) %>%
  filter(!grepl("TNF_TNF.*", interaction)) %>%
  pull(component) %>%
  unique()

# Prepare data
mat_prep <- codesMean %>%
  filter(!grepl("TNFSF9|TNFSF14", interaction)) %>%
  filter(grepl(pattern, interaction)) %>%
  dplyr::select(interaction, all_of(comps)) %>%
  column_to_rownames("interaction") %>%
  scale() %>%
  as.data.frame()

# Create category vector ensuring it is a FACTOR with the specific levels
categories <- factor(case_when(
  grepl("linker", rownames(mat_prep)) ~ "single",
  TRUE ~ sub("^([^_]+)_TNF.*", "\\1_TNF", rownames(mat_prep))
), levels = cat_levels)

# Define Tab10 colors manually
tab10_colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", 
                  "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf")

# Map colors to categories
cat_colors <- setNames(tab10_colors[1:length(levels(categories))], levels(categories))

# Create annotation
ha <- HeatmapAnnotation(
  Category = categories,
  col = list(Category = cat_colors)
)

# Transpose
mat <- t(as.matrix(mat_prep))

# Color function
col_fun <- colorRamp2(seq(-3.5, 3.5, length.out = 11), rev(brewer.pal(n = 11, name = "RdBu")))

# Plot with cluster_column_slices = FALSE to respect factor order
ht_tnf <- Heatmap(
  mat,
  name = "scaled sPCA score",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  top_annotation = ha,
  column_split = categories,
  cluster_column_slices = FALSE, # This enforces the factor level order
  column_title = "TNF family sPCA synergy components",
  heatmap_legend_param = list(title = "scaled sPCA\nscore",
                              at = seq(-4,4,2),
                              border="black")
)
ht_tnf
```

```{r}
pdf(file.path(outs_dir, "plots/spca_tnf_family_interaction_programs.pdf"), width = 12, height = 3.5)
draw(ht_tnf)
dev.off()
```

# Visualize Gene Programs

```{r}
# import python interpreters
library(reticulate)
reticulate::use_condaenv("R-deseq2", conda = "auto", required = TRUE)
library(anndata)

rna <- read_h5ad(file.path(imports_dir, "SIG13/scanpy_outs/SIG13_doublets_DSB7_zscore_degs0.1cutoff_pb.h5ad"))
obs <- rna$obs %>%
  as_tibble(rownames = "cell_bc")

# make expression tibble
expTib <- rna$X %>%
  as.matrix() %>%
  as_tibble(rownames = "cell_bc") %>%
  left_join(obs) %>%
  rename(interaction = ligand_call_DSB7)
```

## Helper Functions

```{r}
plot_component_heatmap_simple <- function(pattern, comp, n_genes, features, expTib,
                                          average_replicates = FALSE,
                                          scale_expression = TRUE,
                                          vmin = -2, vmax = 2) {
  # 1. Get Top Genes
  compGenes <- features %>%
    slice_max(order_by = !!sym(comp), n = n_genes) %>%
    pull(gene)

  # 2. Filter and Prepare Data
  exp_sub <- expTib %>%
    filter(interaction %in% pattern) %>%
    select(interaction, any_of(compGenes)) %>%
    mutate(interaction = factor(interaction, levels = pattern)) %>%
    arrange(interaction)

  if (average_replicates) {
    exp_sub <- exp_sub %>%
      group_by(interaction) %>%
      summarise(across(where(is.numeric), mean))
  }

  # 3. Create Matrix and Scale
  mat <- as.matrix(select(exp_sub, -interaction))
  if (scale_expression) mat <- scale(mat)

  # 4. Annotations and Colors
  # Uses Dark2 for interaction bars
  inter_cols <- setNames(RColorBrewer::brewer.pal(8, "Dark2")[1:length(pattern)], pattern)
  row_ha <- rowAnnotation(interaction = exp_sub$interaction, col = list(interaction = inter_cols))
  #col_fun <- colorRamp2(seq(vmin, vmax, length.out = 9), rev(brewer.pal(9,"RdBu")))
  col_fun <- colorRamp2(seq(vmin, vmax, length.out = 100), magma(100))

  # 5. Plot
  Heatmap(mat,
    name = "Expression",
    col = col_fun,
    left_annotation = row_ha,
    cluster_rows = FALSE,
    cluster_columns = TRUE,
    show_row_names = average_replicates,
    column_title = comp
  )
}
```

## Visualization

```{r}
# renders a heatmap once and saves the same object to both the notebook and a PDF,
# instead of calling plot_component_heatmap_simple twice with duplicated arguments
save_component_heatmap <- function(file, width, height, ...) {
  ht <- plot_component_heatmap_simple(...)
  print(ht)
  pdf(file, width = width, height = height)
  draw(ht)
  dev.off()
}
```

```{r, fig.height=4, fig.width=3}
save_component_heatmap(
  file = file.path(outs_dir, "plots/spca_gene_hm/comp_50_IL4_TNF.pdf"), width = 3, height = 4,
  pattern = c("IL4_TNF","IL4_linker","TNF_linker","linker_linker"),
  comp = "comp_50", n_genes = 20, features = features, expTib = expTib,
  average_replicates = F, scale_expression = F, vmin = 0, vmax = 6
)
```

```{r, fig.height=4, fig.width=3}
save_component_heatmap(
  file = file.path(outs_dir, "plots/spca_gene_hm/comp_31_IL2_IL4_TNFSF18.pdf"), width = 3, height = 4,
  pattern = c("IL2_TNFSF18","IL4_TNFSF18","IL2_IL4","IL2_linker","IL4_linker","linker_TNFSF18","linker_linker"),
  comp = "comp_31", n_genes = 20, features = features, expTib = expTib,
  average_replicates = F, scale_expression = F, vmin = 0, vmax = 4
)
```

```{r, fig.height=4, fig.width=3}
save_component_heatmap(
  file = file.path(outs_dir, "plots/spca_gene_hm/comp_26_IFNG_IL27_IFNB1.pdf"), width = 3, height = 4,
  pattern = c("IL27_IFNB1","IL27_IFNG","IL27_linker","linker_IFNG","linker_linker"),
  comp = "comp_26", n_genes = 20, features = features, expTib = expTib,
  average_replicates = F, scale_expression = F, vmin = 0, vmax = 8
)
```

```{r, fig.height=4, fig.width=3}
# NOTE: the on-screen plot and the exported PDF intentionally differ here (n_genes = 20 vs 15),
# matching the pre-existing behavior of this chunk before cleanup
plot_component_heatmap_simple(pattern = c("IFNG_TNF","IL27_TNF","IFNG_IL27","IFNG_linker","linker_IL27","linker_TNF","linker_linker"),
                       comp = "comp_16",
                       n_genes = 20,
                       features = features,
                       expTib = expTib,
                       average_replicates = F,
                       scale_expression = F,
                       vmin=0, vmax=3)

pdf(file.path(outs_dir, "plots/spca_gene_hm/comp_16_IFNG_IL27_TNF.pdf"), width = 3, height = 4)
plot_component_heatmap_simple(pattern = c("IFNG_TNF","IL27_TNF","IFNG_IL27","IFNG_linker","linker_IL27","linker_TNF","linker_linker"),
                       comp = "comp_16",
                       n_genes = 15,
                       features = features,
                       expTib = expTib,
                       average_replicates = F,
                       scale_expression = F,
                       vmin=0, vmax=3)
dev.off()
```

```{r, fig.height=4, fig.width=3}
save_component_heatmap(
  file = file.path(outs_dir, "plots/spca_gene_hm/comp_4_IL6_IL2.pdf"), width = 3, height = 4,
  pattern = c("IL6_IL2","IL6_linker","linker_IL2","linker_linker"),
  comp = "comp_4", n_genes = 20, features = features, expTib = expTib,
  average_replicates = F, scale_expression = F, vmin = 0, vmax = 4
)
```

```{r, fig.height=4, fig.width=3.5}
save_component_heatmap(
  file = file.path(outs_dir, "plots/spca_gene_hm/comp_31_IL21_CCL19_21A.pdf"), width = 3, height = 4.5,
  pattern = c("IL2_linker","IL21_CCL19","IL21_CCL21A","IL21_linker","linker_CCL19","linker_CCL21A","linker_linker"),
  comp = "comp_31", n_genes = 20, features = features, expTib = expTib,
  average_replicates = F, scale_expression = F, vmin = 0, vmax = 4
)
```













