---
title: "coefficient_concordance_viz"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
---

```{r setup}
library(tidyverse)
library(patchwork)
library(viridisLite)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Repo-relative paths: inputs are read from imports_stable/, outputs are written to analysis_outs/
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find the repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "02_combinatorial_screen_signalseq_SIG13")
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
for (d in c("qc_barcode_counts", "plots/qc_barcode_counts")) dir.create(file.path(outs_dir, d), recursive = TRUE, showWarnings = FALSE)
theme_set(theme_Publication())
```

# Import

Full-dataset production GLM outputs (`05_interaction_scoring/glmGamPoi_interaction_slurm.r`), and
the top30/bottom30/random30 count-subset equivalents from `01_glmGamPoi_interaction_countSubset_slurm.r`.

```{r}
full_singles <- read_csv(file.path(imports_dir, "SIG13/analysis_outs/glmGamPoi/glmGamPoi_singles_lfc_0.1filter.csv"))
full_inter   <- read_csv(file.path(imports_dir, "SIG13/analysis_outs/glmGamPoi/glmGamPoi_interaction_lfc_0.1filter.csv"))

load_subset <- function(count_subset) {
  dir <- file.path(imports_dir, "SIG13/analysis_outs/glmGamPoi", paste0("glmGamPoi_interaction_", count_subset, "count"))
  list(
    singles = read_csv(file.path(dir, "glmGamPoi_singles_lfc_0.1filter.csv")),
    inter   = read_csv(file.path(dir, "glmGamPoi_interaction_lfc_0.1filter.csv"))
  )
}

top30    <- load_subset("top30")
bottom30 <- load_subset("bottom30")
random30 <- load_subset("random30")
```

# Define Consistent Ligands

`consistent_ligands` is the repo-wide reproducibility filter: ligand-pair combos whose interaction
LFCs correlate across biological replicates.

```{r}
repCor <- read_csv(file.path(imports_dir, "SIG13/analysis_outs/replicate_corr/replicate_correlation_interactionLfc_0.2filter.csv"))

consistent_ligands <- repCor %>%
  filter(pearson_corr > 0.25 & num_shared_genes > 1) %>%
  pull(interaction)

length(consistent_ligands)
```

# Calculate Concordance

For each ligand combo, correlate the per-gene log-fold-change coefficients between a count subset (top30,
bottom30, or random30) and the full dataset - separately for the single-ligand terms (`ligand1`,
`ligand2`) and the interaction term (`ligand1:ligand2`).

```{r}
label_single_coef <- function(df) {
  df %>%
    mutate(coef_type = case_when(
      str_ends(single_ligand, "_round1") ~ "ligand1",
      str_ends(single_ligand, "_round2") ~ "ligand2",
      TRUE ~ NA_character_
    ))
}

# correlate one coefficient type's lfc between a subset and the full dataset, per combo
compute_concordance <- function(subset_df, full_df, join_cols, coef_label) {
  subset_df %>%
    select(all_of(join_cols), lfc_subset = lfc, padj_subset = adj_pval) %>%
    inner_join(
      full_df %>% select(all_of(join_cols), lfc_full = lfc, padj_full = adj_pval),
      by = join_cols
    ) %>%
    drop_na(lfc_subset, lfc_full) %>%
    # filter for consistent ligands
    filter(interaction %in% consistent_ligands) %>%
    group_by(interaction) %>%
    # filter for significant effects
    filter((padj_subset < 0.1)|(padj_full < 0.1)) %>%
    summarise(
      pearson_corr = cor(lfc_subset, lfc_full, method = "pearson"),
      n_genes      = n(),
      .groups      = "drop"
    ) %>%
    mutate(coef_type = coef_label)
}

concordance_for_subset <- function(subset_list, full_singles, full_inter, count_subset) {
  subset_singles <- label_single_coef(subset_list$singles)
  full_singles_l <- label_single_coef(full_singles)

  lig1 <- compute_concordance(
    filter(subset_singles, coef_type == "ligand1"),
    filter(full_singles_l, coef_type == "ligand1"),
    join_cols = c("interaction", "name"), coef_label = "single"
  )
  lig2 <- compute_concordance(
    filter(subset_singles, coef_type == "ligand2"),
    filter(full_singles_l, coef_type == "ligand2"),
    join_cols = c("interaction", "name"), coef_label = "single"
  )
  inter <- compute_concordance(
    subset_list$inter, full_inter,
    join_cols = c("interaction", "name"), coef_label = "interaction"
  )

  bind_rows(lig1, lig2, inter) %>%
    mutate(count_subset = count_subset)
}

concordance_all <- bind_rows(
  concordance_for_subset(top30, full_singles, full_inter, "top30"),
  concordance_for_subset(bottom30, full_singles, full_inter, "bottom30"),
  concordance_for_subset(random30, full_singles, full_inter, "random30")
)

concordance_all

write_csv(concordance_all, file.path(outs_dir, "qc_barcode_counts/correlation_diff_barcode_cutoffs.csv"))
```

# Plots

```{r, fig.width=10, fig.height=4}
concordance_all %>%
  ggplot(aes(x = count_subset, y = pearson_corr, color = count_subset)) +
  geom_boxplot(position = position_dodge(0.8), width = 0.5) +
  #geom_point(position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.8), alpha = 0.3, size = 0.8) +
  ylim(0, 1) +
  ylab("Pearson r\n(vs. full dataset)") +
  theme(aspect.ratio = 1) +
  facet_wrap(~coef_type) +
  scale_color_brewer(palette = "Dark2") +
  ggtitle(sprintf(
    "Reproducible combos only (n=%d)",
    length(unique(intersect(concordance_all$interaction, consistent_ligands)))
  ))

ggsave(file.path(outs_dir, "plots/qc_barcode_counts/correlation_diff_barcode_cutoffs.pdf"))
```

# Test Significance of Difference

Use Tukey HSD test to examine significance of differences in pearson correlation.

```{r}
concordance_all %>%
  group_by(coef_type, count_subset) %>%
  summarise(
    median_corr = median(pearson_corr, na.rm = TRUE),
    mean_corr   = mean(pearson_corr, na.rm = TRUE),
    n_combos    = n(),
    .groups     = "drop"
  ) %>%
  arrange(coef_type, count_subset)
```

```{r}
# significance for interactions
model <- aov(pearson_corr ~ count_subset, data = concordance_all %>% filter(coef_type == "interaction"))
tukey_results <- TukeyHSD(model)
print(tukey_results)
```

```{r}
# significance for singles
model <- aov(pearson_corr ~ count_subset, data = concordance_all %>% filter(coef_type == "single"))
tukey_results <- TukeyHSD(model)
print(tukey_results)
```
