---
title: "SIG13 ligand validation"
author: "Ian Zumpano"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

# Goal: verify that cytokine signaling is consistent with other publicly avaliable datasets

```{r message=FALSE, warning=FALSE}
# install.packages("BiocManager")
# BiocManager::install(c("fgsea", "msigdbr", "clusterProfiler", "org.Mm.eg.db"))

library(tidyverse)
library(msigdbr)
library(dplyr)
library(tibble)
library(org.Mm.eg.db)
library(fgsea)
```

```{r}
# Repo-relative paths: inputs from imports_stable/, outputs to analysis_outs/
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "02_combinatorial_screen_signalseq_SIG13")
dir.create(file.path(outs_dir, "plots"), showWarnings = FALSE, recursive = TRUE)
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
theme_set(theme_Publication())
```

# Load Data

```{r}
# data <- read.csv(file.path(imports_dir, "SIG13/analysis_outs/glmGamPoi/glmGamPoi_single_term/glmGamPoi_singleTerm_lfc_0.1filter.csv"))
```


## To start, filter for single ligand conditions (linker)

```{r}
# linkers <- data %>% filter(grepl("linker", condition))
# 
# # Get new column with just ligand (eg IL12_linker or linker_IL12 -> IL12)
# 
# linkers$ligand <- sub("^linker_|_linker$", "", linkers$condition)
# 
# # Create new variable that is -log10(padj) x lfc then rank significant genes based on this
# 
# linkers <- linkers %>% mutate(lfc_x_padj = (lfc * -log10(adj_pval)))
```


```{r}
# write_csv(linkers, file.path(outs_dir, "ligand_signature_validation_linkers.csv"))
```


```{r}
linkers <- read_csv(file.path(imports_dir, "SIG13/analysis_outs_zumpano/ligand_signature_validation_linkers.csv"))
```


# Run GSEA on Hallmark gene sets. 

Run the different conditions for the same ligands separate (eg \_linker or linker\_ as individual searches), validate results between paired conditions. Note: for TGFb - use only TGFb_linker for analysis, but run both.  

For each condition, input to GSEA is a ranked list of all genes ranked by lfc.

```{r}
# Define helper function to get hallmark 

run_gsea_for_linkers <- function(linkers, pathways) {
  ligands <- linkers$ligand %>% unique()
  results_all <- data.frame()  # container for all results
  
  for (lig in ligands) {
    ligand_data <- linkers %>% filter(ligand == lig)
    treatments <- ligand_data$condition %>% unique()
    
    for (treatment in treatments) {
      gsea_data <- ligand_data %>% filter(condition == treatment) %>% 
        arrange(desc(lfc_x_padj))

      # Skip if no genes pass filtering
      if (nrow(gsea_data) == 0) next

      gsea_ranks <- gsea_data$lfc
      names(gsea_ranks) <- gsea_data$name

      score_type <- if (all(gsea_ranks > 0)) "pos" else "std"

      fgsea_res <- fgsea(
        pathways = pathways,
        stats = gsea_ranks,
        minSize = 10,
        maxSize = 500,
        nPermSimple = 10000,
        scoreType = score_type
      )

      # Append ligand and condition to results
      fgsea_res <- fgsea_res %>% mutate(ligand = lig, condition = treatment) %>%
        arrange(padj) %>% relocate(ligand, condition, .before = everything())

      # Add to the master results table
      results_all <- bind_rows(results_all, fgsea_res)
    }
  }
  
  return(results_all)
}
```

```{r}
msigdbr_mouse_hallmark <- msigdbr(species = "Mus musculus", category = "H")
hallmark_pathways <- split(msigdbr_mouse_hallmark$gene_symbol, msigdbr_mouse_hallmark$gs_name)

# hallmark_results <- run_gsea_for_linkers(linkers, pathways = hallmark_pathways)
# sig_hallmark_results <- hallmark_results %>% filter(padj < 0.1)
```


```{r}
# # Write GSEA Results
# write_csv(hallmark_results, file.path(outs_dir, "hallmark_gsea_results.csv"))
# write_csv(sig_hallmark_results, file.path(outs_dir, "significant_hallmark_gsea_results.csv"))
```


```{r}
hallmark_results <- read_csv(file.path(imports_dir, "SIG13/analysis_outs_zumpano/hallmark_gsea_results.csv"))
sig_hallmark_results <- read_csv(file.path(imports_dir, "SIG13/analysis_outs_zumpano/significant_hallmark_gsea_results.csv"))
```


## Create a function to measure similarity between significant hits in conditions of same ligands

Goal is to verify that there is consistency between \_linker and linker\_ conditions.

```{r}
library(dplyr)
library(purrr)

jaccard_from_lists <- function(df) {
  
  # Get unique ligands
  ligands <- unique(df$ligand)
  
  results <- map_dfr(ligands, function(lig) {
    # Subset for this ligand
    sub_df <- df %>% filter(ligand == lig)
    
    # Split by condition and pull pathways
    cond_sets <- sub_df %>%
      group_by(condition) %>%
      summarise(pathways = list(unique(pathway)), .groups = "drop") %>%
      pull(pathways)
    
    # Only compute if exactly two conditions
    if (length(cond_sets) != 2) {
      tibble(ligand = lig, jaccard = NA_real_)
    } else {
      # Compute Jaccard
      inter_len <- length(intersect(cond_sets[[1]], cond_sets[[2]]))
      union_len <- length(union(cond_sets[[1]], cond_sets[[2]]))
      tibble(ligand = lig, jaccard = inter_len / union_len)
    }
  })
  
  return(results)
}

# Usage
jaccard_from_lists(sig_hallmark_results) %>% arrange(desc(jaccard))
```

There is some variation but for the most part looks like there is consistency between \_linker and linker\_ conditions


# Plot hallmark gene sets

For plotting, use \_linker conditions only.

```{r}
data_to_plot <- sig_hallmark_results %>% filter(grepl("_linker", condition))
```

## Dotplot of all Hallmark pathways

```{r}
library(ggplot2)

ggplot(data_to_plot, aes(x = condition, y = pathway, size = -log10(padj), color = NES)) +
  geom_point() +
  scale_size_continuous(name = "-log10 padj") +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, name = "NES") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  ) +
  labs(x = "Combination", y = "Pathway", title = "GSEA Dotplot")
```

## Filter for hallmark pathways that correspond to ligands in the data. 

```{r}
cytokine_strings <- c("INTERFERON", "IL2", "TNF", "_IL", "STAT", "INFLAMMATORY", "MYC", "TGF")
filtered_dotplot_data <- data_to_plot %>%
      filter(str_detect(pathway, paste(cytokine_strings, collapse = "|")))
filtered_dotplot_data$pathway %>% unique()
```


```{r}
library(ggplot2)

ggplot(filtered_dotplot_data, aes(x = condition, y = pathway, size = -log10(padj), color = NES)) +
  geom_point() +
  scale_size_continuous(name = "-log10 padj") +
  scale_color_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, name = "NES") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  ) +
  labs(x = "Combination", y = "Pathway", title = "GSEA Results by Ligand")
```

# GSEA Enrichment Plots

Make gsea enrichment plots for each ligand x pathway combination - eg TNF_linker and hallmark_TNF.

```{r}
pairs_to_plot <- tibble(
  treatment = c("IFNG_linker", "IL6_linker", "IL2_linker", "TNF_linker", "TGFB1_linker", "IL27_linker"),
  pathway     = c("HALLMARK_INTERFERON_GAMMA_RESPONSE", 
                  "HALLMARK_IL6_JAK_STAT3_SIGNALING", 
                  "HALLMARK_IL2_STAT5_SIGNALING", 
                  "HALLMARK_TNFA_SIGNALING_VIA_NFKB", 
                  "HALLMARK_TGF_BETA_SIGNALING",
                  "HALLMARK_INTERFERON_GAMMA_RESPONSE"))
```


```{r}
library(dplyr)
library(purrr)
library(fgsea)

enrichment_plots <- pairs_to_plot %>%
  pmap(function(treatment, pathway) {
    # subset ligand-specific data
    ligand_data <- linkers %>% filter(condition == treatment)
    
    # rank genes
    gsea_data <- ligand_data %>% arrange(desc(lfc_x_padj))
    gsea_ranks_to_plot <- gsea_data$lfc
    names(gsea_ranks_to_plot) <- gsea_data$name
    
    # get pathway-specific results
    hallmark_results_to_plot <- hallmark_results %>%
      filter(condition == treatment, pathway == !!pathway)
    
    # make enrichment plot
    plotEnrichment(hallmark_pathways[[pathway]], gsea_ranks_to_plot) +
      labs(
        title = paste0(treatment, " + ", pathway),
        subtitle = paste0(
          "NES = ", round(hallmark_results_to_plot$NES, 2),
          ", padj = ", signif(hallmark_results_to_plot$padj, 2)
        )
      )
  })
```

```{r, fig.height=8, fig.width=10}
library(patchwork)

# Arrange plots in 2x3 grid
wrap_plots(enrichment_plots, ncol = 3, nrow = 2)
ggsave(file.path(outs_dir, "plots/gsea_res.pdf"))
```
