---
title: "DEG Analysis SIG18"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup}
library(tidyverse)
library(ggplot2)
library(patchwork)
library(DESeq2)
library(IHW)
library(RColorBrewer)
library(pheatmap)
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Repo-relative paths: inputs from imports_stable/, outputs to analysis_outs/<module>/
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "05_in_vitro_differentiation_RNAseq_SIG18")
dir.create(file.path(outs_dir, "deg"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(outs_dir, "plots"), recursive = TRUE, showWarnings = FALSE)
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
theme_set(theme_Publication())
```

This Rmd ports the unified condition-level DEG + interaction-scoring workflow
developed in SIG26 to the SIG18 dataset.

SIG18 differs from SIG26 in its reference state: `ligand1` is the Th-polarizing
condition with baseline **Th0** (activated but unpolarized), while `ligand2` is
TNF with baseline **none**. The fully-untreated reference is therefore
`Th0_none`. We encode this by treating `Th0` as the baseline for `ligand1`, so
the derived `condition` variable behaves exactly like SIG26's: `Th0_none` ->
`none`, a single polarizing ligand -> its own name (e.g. `Th1`), `Th0_TNF` ->
`TNF`, and a polarized + TNF well -> the combination (e.g. `Th1-TNF`).

```{r}
# Reference (baseline) value of each ligand. A sample is "untreated" for a
# ligand when that ligand equals its baseline.
baseline1 <- "Th0"    # ligand1 (Th polarization) baseline
baseline2 <- "none"   # ligand2 (TNF) baseline
```

### [Import Data]{.underline}

```{r}
counts <- read.csv(file.path(imports_dir, "SIG18/processing_outs/count_matrix_umiDeDup_SIG18.csv"), row.names = "X")
colnames(counts) <- gsub("\\.", "-", colnames(counts))

metadata <- read_csv(file.path(imports_dir, "SIG18/processing_outs/processed_metadata_SIG18.csv"))

featureNames <- read_csv(file.path(imports_dir, "SIG18/processing_outs/featureNames_SIG18.csv")) %>%
  dplyr::select(-category)
```

Mouse 4 and mouse 9 are dropped: they were both on the very edge of the plate
and clustered together anomalously across analyses.

```{r}
metadata <- metadata %>%
  filter(!replicate %in% c("mouse4", "mouse9")) %>%
  filter(sample_ID != "TGFb_none_mouse3_F_E06_col7_SIG18")
```

```{r}
# Derive a single `condition` factor from the two ligand columns:
#   - both ligands at baseline -> "none"  (Th0_none, the untreated reference)
#   - one ligand active        -> that ligand's name (single-ligand condition)
#   - both ligands active       -> "ligand1-ligand2" (combination)
make_condition <- function(l1, l2) {
  parts <- c(if (l1 != baseline1) l1, if (l2 != baseline2) l2)
  if (length(parts) == 0) "none" else paste(parts, collapse = "-")
}

metaDF <- metadata %>%
  mutate(
    condition = map2_chr(ligand1, ligand2, make_condition),
    replicate = factor(replicate),
    sex       = factor(sex),
    ligand1   = factor(ligand1, levels = c(baseline1, sort(setdiff(unique(ligand1), baseline1)))),
    ligand2   = factor(ligand2, levels = c(baseline2, sort(setdiff(unique(ligand2), baseline2)))),
    condition = factor(condition, levels = c("none", sort(setdiff(unique(condition), "none"))))
  ) %>%
  mutate(condition = factor(condition, c("none","TNF",
                                         "Th1","Th1-TNF",
                                         "Th2","Th2-TNF",
                                         "Th17","Th17-TNF",
                                         "iTreg","iTreg-TNF",
                                         "TGFb","TGFb-TNF"))) %>%
  as.data.frame()
rownames(metaDF) <- metaDF$sample_ID

counts <- counts[, rownames(metaDF)]
```

### [Helper Functions]{.underline}

```{r}
# Build DDS object with counts filtering
build_dds <- function(counts_mat, meta, design, min_count = 10, min_samples = 3) {
  dds <- DESeqDataSetFromMatrix(counts_mat, colData = meta, design = design)
  dds <- dds[rowSums(counts(dds) >= min_count) >= min_samples, ]
  DESeq(dds, parallel = TRUE)
}

# Extract DEG results for a vector of coefficient names and combine into one tibble
extract_deg_results <- function(dds, terms, label_col = "term", clean_fn = identity) {
  map_dfr(terms, function(term) {
    results(dds, name = term,
            independentFiltering = FALSE, filterFun = ihw, parallel = TRUE) %>%
      as_tibble(rownames = "ensembl_ID") %>%
      left_join(featureNames, by = "ensembl_ID") %>%
      mutate(!!label_col := clean_fn(term)) %>%
      arrange(padj)
  })
}

# Look up the exact resultsNames() term for a condition value.
# Handles DESeq2 encoding hyphens as either "-" or "." in model matrix names.
cond_term <- function(res_names, cond) {
  pat  <- paste0("^condition_", gsub("-", "[-.]", cond, fixed = FALSE), "_vs_none$")
  hits <- res_names[grepl(pat, res_names)]
  if (length(hits) != 1) stop("Expected 1 match for '", cond, "', found: ", length(hits))
  hits
}
```

### [QC Plots]{.underline}

```{r, message=FALSE}
# Condition-level model on all samples; reused for both QC and all DEG analyses below
dds_cond <- build_dds(counts, metaDF, ~ condition + replicate)
vsd      <- vst(dds_cond, blind = TRUE)
```

#### Sample Distance Heatmap

```{r, fig.height=10, fig.width=11}
sampleDists <- dist(t(assay(vsd)))
distMatrix  <- as.matrix(sampleDists)
rownames(distMatrix) <- colnames(dds_cond)
colnames(distMatrix) <- colnames(dds_cond)

ann_df <- as.data.frame(colData(vsd)[, c("ligand1", "ligand2", "sex")])
mk_pal <- function(n) colorRampPalette(brewer.pal(8, "Dark2"))(n)
ann_colors <- list(
  ligand1 = setNames(mk_pal(length(unique(ann_df$ligand1))), sort(unique(ann_df$ligand1))),
  ligand2 = setNames(mk_pal(length(unique(ann_df$ligand2))), sort(unique(ann_df$ligand2))),
  sex     = setNames(mk_pal(length(unique(ann_df$sex))),     sort(unique(ann_df$sex)))
)

pheatmap(distMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = viridisLite::viridis(100),
         annotation_row = ann_df,
         annotation_colors = ann_colors,
         show_rownames = FALSE,
         show_colnames = FALSE)
```

#### Per-Replicate Distance Heatmaps

```{r, fig.height=8, fig.width=9}
for (rep in levels(metaDF$replicate)) {
  keep    <- colnames(vsd)[colData(vsd)$replicate == rep]
  vsd_rep <- vsd[, keep]

  dists_rep  <- dist(t(assay(vsd_rep)))
  mat_rep    <- as.matrix(dists_rep)
  rownames(mat_rep) <- colnames(vsd_rep)
  colnames(mat_rep) <- colnames(vsd_rep)

  ann_rep <- as.data.frame(colData(vsd_rep)[, c("ligand1", "ligand2")])
  ann_colors_rep <- list(
    ligand1 = setNames(mk_pal(length(unique(ann_rep$ligand1))), sort(unique(ann_rep$ligand1))),
    ligand2 = setNames(mk_pal(length(unique(ann_rep$ligand2))), sort(unique(ann_rep$ligand2)))
  )

  pheatmap(mat_rep,
           clustering_distance_rows = dists_rep,
           clustering_distance_cols = dists_rep,
           col = viridisLite::viridis(100),
           annotation_row = ann_rep,
           annotation_colors = ann_colors_rep,
           show_rownames = FALSE,
           show_colnames = FALSE,
           main = paste("Replicate:", rep))
}
```

#### PCA

```{r, fig.height=10, fig.width=10}
p1 <- plotPCA(vsd, intgroup = "ligand1",   ntop = 2000) + theme(aspect.ratio = 1)
p2 <- plotPCA(vsd, intgroup = "ligand2",   ntop = 2000) + theme(aspect.ratio = 1)
p3 <- plotPCA(vsd, intgroup = "condition",   ntop = 2000) + theme(aspect.ratio = 1) +
  scale_color_brewer(palette = "Paired")
p4 <- plotPCA(vsd, intgroup = "replicate", ntop = 2000) + theme(aspect.ratio = 1)
p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
ggsave(file.path(outs_dir, "plots/SIG18_PCA.pdf"))
```

### [Analysis 1: Condition Effects]{.underline}

Effect of every condition vs. none (Th0_none), using the condition-level model. Captures both single-ligand and combination conditions in a single unified analysis.

```{r}
cond_terms <- grep("^condition_", resultsNames(dds_cond), value = TRUE)

res_conditions <- extract_deg_results(
  dds_cond, cond_terms,
  label_col = "condition",
  # DESeq2 may encode hyphens as dots in term names; restore hyphens for readability
  clean_fn  = function(x) gsub("\\.", "-", gsub("^condition_|_vs_none$", "", x))
)
```

```{r}
write_csv(res_conditions, file.path(outs_dir, "deg/res_conditions_SIG18.csv"))
```

```{r}
res_conditions %>%
  filter(padj < 0.01) %>%
  group_by(condition) %>%
  summarise(n_degs = n()) %>%
  ggplot(aes(x = condition, y = n_degs)) +
    geom_col() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### [Analysis 2: Interaction Scoring]{.underline}

For each gene in each combination condition, quantifies whether the combined response is greater than (synergy), less than (buffering), or equal to (additive) the sum of the individual responses.

**Definitions** (following SIG18 scoring):

All LFC values are pulled directly from DESeq2 model.

- `lfc_lig1`: logFC of ligand1 alone vs. none (Th polarization vs. Th0)
- `lfc_lig2`: logFC of ligand2 alone vs. none (TNF vs. none, on Th0 background)
- `lfc_interaction`: `lfc_combo − lfc_lig1 − lfc_lig2` (deviation from additivity)
- `lfc_total`: `lfc_lig1 + lfc_lig2 + lfc_interaction` = `lfc_combo`
- `interaction_score`: `lfc_interaction / lfc_total` — fraction of total effect attributable to non-additivity (set to 0 when `padj_interaction > 0.1`)
- `padj_interaction`: adjusted p-value from the synergy contrast `lfc_combo − lfc_lig1 − lfc_lig2`

**Classification**:
- `synergy positive`: interaction significant, total effect positive, score > 0
- `synergy negative`: interaction significant, total effect negative, score > 0
- `buffering`: interaction significant, score < 0
- `none`: interaction not significant

Note: ligand1/ligand2 assignment is arbitrary; scores and classes are symmetric with respect to swapping the two ligands.

```{r}
# Define combo conditions and respective ligands (both ligands active,
# i.e. a polarized Th condition combined with TNF)
combo_map <- metaDF %>%
  filter(as.character(ligand1) != baseline1, as.character(ligand2) != baseline2) %>%
  dplyr::select(condition, ligand1, ligand2) %>%
  mutate(across(everything(), as.character)) %>%
  distinct()

combo_map
```

#### Synergy contrast p-values

Here we are  calculating interaction effects using a contrast design.
e.g. Th1-TNF_vs_none - (Th1_vs_none + TNF_vs_none)

```{r}
rn_cond <- resultsNames(dds_cond)

res_synergy_pvals <- map_dfr(seq_len(nrow(combo_map)), function(i) {
  row <- combo_map[i, ]
  results(dds_cond,
          contrast = list(
            c(cond_term(rn_cond, row$condition)),
            c(cond_term(rn_cond, row$ligand1), cond_term(rn_cond, row$ligand2))
          ),
          independentFiltering = FALSE, filterFun = ihw, parallel = TRUE) %>%
    as_tibble(rownames = "ensembl_ID") %>%
    dplyr::select(ensembl_ID, lfc_interaction = log2FoldChange, padj_interaction = padj) %>%
    mutate(condition = row$condition)
})
```

#### Compute interaction scores

```{r}
lfc_lookup <- res_conditions %>%
  dplyr::select(ensembl_ID, gene, condition, log2FoldChange) %>%
  dplyr::rename(lfc = log2FoldChange)

interaction_raw <- map_dfr(seq_len(nrow(combo_map)), function(i) {
  row <- combo_map[i, ]
  lfc_lookup %>%
    filter(condition == row$condition) %>%
    dplyr::select(ensembl_ID, gene, lfc_combo = lfc) %>%
    left_join(
      lfc_lookup %>% filter(condition == row$ligand1) %>%
        dplyr::select(ensembl_ID, lfc_lig1 = lfc),
      by = "ensembl_ID"
    ) %>%
    left_join(
      lfc_lookup %>% filter(condition == row$ligand2) %>%
        dplyr::select(ensembl_ID, lfc_lig2 = lfc),
      by = "ensembl_ID"
    ) %>%
    left_join(
      res_synergy_pvals %>% filter(condition == row$condition) %>%
        dplyr::select(ensembl_ID, lfc_interaction, padj_interaction),
      by = "ensembl_ID"
    ) %>%
    mutate(condition = row$condition, ligand1 = row$ligand1, ligand2 = row$ligand2)
})
```

#### Classify interaction effects

```{r}
interaction_scored <- interaction_raw %>%
  mutate(
    lfc_total       = lfc_combo,
    interaction_score = case_when(
      is.na(padj_interaction) | padj_interaction > 0.1 ~ 0,
      TRUE ~ lfc_interaction / lfc_total
    ),
    interaction_class = case_when(
      padj_interaction <= 0.1 & lfc_total > 0 & interaction_score > 0 ~ "synergy positive",
      padj_interaction <= 0.1 & lfc_total < 0 & interaction_score > 0 ~ "synergy negative",
      padj_interaction <= 0.1 & interaction_score < 0                  ~ "buffering",
      TRUE ~ "none"
    )
  )
```

```{r}
interaction_scored %>%
  filter(interaction_class != "none") %>%
  group_by(condition, interaction_class) %>%
  summarise(n = n(), .groups = "drop") %>%
  arrange(condition, interaction_class)
```

```{r}
write_csv(interaction_scored, file.path(outs_dir, "deg/res_interaction_scored_SIG18.csv"))
```
