---
title: "SIG16 ATAC: DESeq2 QC (merged replicates, IDR 0.05 peak atlas)"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
---

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(DESeq2)
  library(pheatmap)
  library(RColorBrewer)
  library(ggrepel)
})
knitr::opts_chunk$set(echo = TRUE)

# Repo-relative paths (inputs: imports_stable/, outputs: analysis_outs/)
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "04_tnf_interaction_ATACseq_SIG16")
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
theme_set(theme_Publication())

set.seed(1)
options(ggrepel.max.overlaps = Inf)

idr <- 0.05
job.name <- paste0("idr-", idr, "/")

IN_DIR   <- file.path(imports_dir, "SIG16/analysis_outs/deseq2-merged-reps", job.name)
OUT_DIR  <- file.path(outs_dir, "deseq2-merged-reps", job.name)
PLOT_DIR <- file.path(outs_dir, "plots_new", "deseq2-merged-reps", job.name)
dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(PLOT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(outs_dir, "plots"), recursive = TRUE, showWarnings = FALSE)
```

# Goal

Rebuild the DESeq2 object for the SIG16 bulk ATAC screen (naive CD4 T cells stimulated
with `ligand1` in {none, IL4, IL6, TGFb} crossed with `ligand2` in {none, TNF}, 3 mice)
over the IDR 0.05 consensus peak atlas, and reproduce the QC panels from the external
collaborator's `ATAC-analysis/SIG16_ATAC/1-deseq.R`.

The design is `~ mouse + ligand1 + ligand2 + ligand1:ligand2`, so the model carries a
per-mouse blocking term, the two single-ligand main effects, and the interaction term
that the downstream synergy/antagonism calls are built on.

# Import

`counts_mat.rds` is the `Rsubread::featureCounts` output over the peak atlas; `annotation.tsv`
is the ChIPseeker annotation of the same atlas (peak -> nearest gene / distance to TSS).

```{r}
counts_mat <- readRDS(file.path(IN_DIR, "counts_mat.rds"))
count.matrix <- counts_mat$counts

annot <- read.csv(file.path(imports_dir, "SIG16/analysis_outs/chipseeker", job.name,
                            "all.peakatlas/annotation.tsv"), sep = "\t")

# Condition / ligand colors come from the RColorBrewer "Paired" palette: each ligand1
# gets one light/dark pair, light = ligand2 none, dark = ligand2 TNF. Written back to
# `plots/palette.rds`, which is what 02_/03_/04_ read.
paired <- brewer.pal(12, "Paired")
names(paired) <- c("lt.blue", "blue", "lt.green", "green", "lt.red", "red",
                   "lt.orange", "orange", "lt.purple", "purple", "lt.brown", "brown")

pal <- list(
  condition = c(
    none_none = unname(paired["lt.purple"]),
    none_TNF  = unname(paired["purple"]),
    IL4_none  = unname(paired["lt.blue"]),
    IL4_TNF   = unname(paired["blue"]),
    IL6_none  = unname(paired["lt.green"]),
    IL6_TNF   = unname(paired["green"]),
    TGFb_none = unname(paired["lt.red"]),
    TGFb_TNF  = unname(paired["red"])
  ),
  ligand1 = c(
    none = unname(paired["purple"]),
    IL4  = unname(paired["blue"]),
    IL6  = unname(paired["green"]),
    TGFb = unname(paired["red"])
  ),
  ligand2 = c(
    none = unname(paired["lt.orange"]),
    TNF  = unname(paired["orange"])
  ),
  rep = c(`1` = unname(paired["blue"]),
          `2` = unname(paired["green"]),
          `3` = unname(paired["red"]))
)
saveRDS(pal, file.path(outs_dir, "plots/palette.rds"))

dim(count.matrix)
colSums(count.matrix)
```

# Sample metadata

Sample names encode `{ligand1}_{ligand2}_{rep}`; replicate index is also the mouse of origin.

```{r}
condition <- factor(str_split_i(colnames(count.matrix), "_[1-3]", i = 1),
                    levels = c("none_none", "none_TNF", "IL4_none", "IL4_TNF",
                               "IL6_none", "IL6_TNF", "TGFb_none", "TGFb_TNF"))
ligand1 <- factor(str_split_i(colnames(count.matrix), "_", i = 1),
                  levels = c("none", "IL4", "IL6", "TGFb"))
ligand2 <- factor(str_split_i(colnames(count.matrix), "_", i = 2),
                  levels = c("none", "TNF"))
rep   <- factor(str_split_i(colnames(count.matrix), "_", i = 3))
mouse <- factor(rep(c(1, 2, 3), 8), levels = c(1, 2, 3))

metadata <- data.frame(condition = condition, ligand1 = ligand1, ligand2 = ligand2,
                       rep = rep, mouse = mouse)
rownames(metadata) <- colnames(count.matrix)
metadata
```

# FRiP

Fraction of reads in peaks, taken from the `featureCounts` assignment summary.

```{r, fig.width=8, fig.height=6}
FRiP_scores_atlas <- apply(counts_mat$stat[, !colnames(counts_mat$stat) %in% c("Status")],
                           2, function(sample) sample[1] / sum(sample))
FRiP_scores_atlas.df <- data.frame(sample = colnames(count.matrix),
                                   FRiP = FRiP_scores_atlas,
                                   condition = condition)

frip.p <- ggplot(FRiP_scores_atlas.df) +
  geom_bar(aes(x = sample, y = FRiP, fill = condition), stat = "identity") +
  scale_fill_manual(values = pal$condition) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
frip.p
ggsave(file.path(outs_dir, "plots_new", paste0("bar-FRiP-idr-", idr, ".pdf")),
       frip.p, width = 8, height = 6)
```

# Fit DESeq2

Set `refit <- TRUE` to re-run `DESeq()` (a few minutes over ~63k peaks); otherwise the
stored fit is loaded.

```{r}
refit <- FALSE

if (refit) {
  dds <- DESeqDataSetFromMatrix(count.matrix, colData = metadata,
                                design = ~ mouse + ligand1 + ligand2 + ligand1:ligand2)
  dds <- DESeq(dds)

  saveRDS(dds, file.path(OUT_DIR, "deseq2_dataset.rds"))
  saveRDS(metadata, file.path(OUT_DIR, "metadata.rds"))
  write.csv(metadata, file.path(OUT_DIR, "metadata.csv"), quote = FALSE)

  # VST matrix with the ChIPseeker annotation columns (distance to TSS / nearest gene)
  # pasted on, keyed by peak ID. Row order and rownames follow the DESeq2 object.
  normalized_counts <- vst(dds, blind = FALSE) %>%
    assay() %>%
    as.data.frame() %>%
    rownames_to_column("peak") %>%
    left_join(annot %>% dplyr::select(peak = name, distanceToTSS, geneSymbol),
              by = "peak") %>%
    column_to_rownames("peak")
  saveRDS(normalized_counts, file.path(OUT_DIR, "normalized_counts.rds"))
} else {
  # stable copy of the stored fit
  dds <- readRDS(file.path(IN_DIR, "deseq2_dataset.rds"))
  metadata <- readRDS(file.path(IN_DIR, "metadata.rds"))
  normalized_counts <- readRDS(file.path(IN_DIR, "normalized_counts.rds"))
}

resultsNames(dds)
```

Peaks that are zero in every sample (should be none, the atlas is built from called peaks):

```{r}
sum(rowSums(normalized_counts[, rownames(metadata)] == 0) == nrow(metadata))
```

# Dispersion

```{r, fig.width=6, fig.height=6}
plotDispEsts(dds)

pdf(file.path(PLOT_DIR, "dispersion.pdf"))
plotDispEsts(dds)
invisible(dev.off())
```

# PCA

VST on the top 20% most variable peaks, `blind = FALSE` so the design is respected.

```{r, fig.width=4, fig.height=4}
vsd <- vst(dds, blind = FALSE)
vsd.mat <- assay(vsd)

pcaData <- plotPCA(vsd, intgroup = "condition", returnData = TRUE,
                   ntop = round(0.2 * nrow(vsd)))
percentVar <- round(100 * attr(pcaData, "percentVar"))
pcaData$sample <- pcaData$name

pca.p <- ggplot(pcaData, aes(PC1, PC2, color = condition, label = sample)) +
  geom_point(size = 3) +
  scale_color_manual(values = pal$condition) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme(aspect.ratio = 1)
pca.p
ggsave(file.path(PLOT_DIR, "pca-name.pdf"), pca.p)
```

# Sample-sample correlation

```{r, fig.width=12, fig.height=12}
vsd.cor <- cor(vsd.mat)
heat.colors <- colorRampPalette(brewer.pal(6, "YlOrRd"))(100)

cor.p <- pheatmap::pheatmap(vsd.cor,
                            annotation = metadata[, c("ligand1", "ligand2", "mouse")],
                            color = heat.colors,
                            cluster_cols = TRUE, cluster_rows = TRUE,
                            border_color = "white")
save_pheatmap_pdf(cor.p, file.path(PLOT_DIR, "heatmap-sample_correlation.pdf"),
                  width = 14, height = 14)
```
