---
title: "SIG16 ATAC: TF motif enrichment in differentially accessible regions"
author: "EYW"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    html_preview: false
---

```{r setup}
suppressPackageStartupMessages({
  library(tidyverse)
  library(patchwork)
})
knitr::opts_chunk$set(echo = TRUE)

# Repo-relative paths (inputs: imports_stable/, outputs: analysis_outs/)
repo_dir <- (function(d = normalizePath(getwd())) {
  while (!dir.exists(file.path(d, "imports_stable"))) {
    if (dirname(d) == d) stop("Could not find repo root (folder containing imports_stable/)")
    d <- dirname(d)
  }
  d
})()
imports_dir <- file.path(repo_dir, "imports_stable")
outs_dir <- file.path(repo_dir, "analysis_outs", "04_tnf_interaction_ATACseq_SIG16")
source(file.path(repo_dir, "functions/r_custom/plotting_fxns.R"))
theme_set(theme_Publication())

set.seed(1)

idr <- 0.05
job.name <- paste0("idr-", idr)

DESEQ_DIR <- file.path(imports_dir, "SIG16/analysis_outs/deseq2-merged-reps", job.name)
OUT_DIR   <- file.path(outs_dir,
                       "motif_enrichment-hypergeometric-merged-reps", job.name)
PLOT_DIR  <- file.path(outs_dir, "plots",
                       "motif_enrichment-hypergeometric-merged-reps")

dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(PLOT_DIR, recursive = TRUE, showWarnings = FALSE)

padj.threshold <- 0.05
lfc.threshold  <- log2(1.5)
```

# Goal

Ask which TF motifs are over-represented in each set of differentially accessible regions,
reproducing `ATAC-analysis/SIG16_ATAC/3-motif-enrichment.R`.

The test is a one-sided hypergeometric test per motif: given the number of atlas peaks
containing a FIMO hit for that motif, how surprising is the number of hits inside the region
set of interest? Fold enrichment is the hit rate in the region set over the hit rate across
the whole atlas, and p-values are BH-adjusted across motifs within each region set.

Seventeen region sets are tested:

- **8 single-ligand sets** — gained / lost peaks for each of TGFb, IL4, IL6, TNF.
- **9 interaction sets** — the three SIG13 interaction classes (`synergy positive`,
  `synergy negative`, `buffering`) for each of the three `ligand1 x TNF` pairs, taken from
  `03_interaction_classification.Rmd`.

Splitting the interaction peaks by class rather than by the sign of the interaction coefficient
is the point of this notebook: `synergy positive` and `synergy negative` peaks both have a
same-signed interaction term but move in opposite directions overall, so the motifs driving them
should differ. `buffering` peaks are the ones where the second ligand blunts the first.

> Note: the collaborator's script tested TGFb, IL4 and TNF singles plus the TGFb x TNF and
> IL4 x TNF interactions, split only on the sign of the interaction coefficient. IL6 is included
> here as well, since the corresponding DESeq2 and interaction-classification results exist.

# Import

`motif_mtx.rds` is a peaks x motifs binary matrix of FIMO hits over the same atlas;
`motif_info.rds` maps each motif to its TF and structural family.

```{r}
motif.mtx.peak.final <- readRDS(file.path(imports_dir, "SIG16/analysis_outs/fimo", job.name, "motif_mtx.rds"))
colnames(motif.mtx.peak.final) <- stringr::str_split_i(colnames(motif.mtx.peak.final), "_[A-Z]", 1)

motif.info <- readRDS(file.path(imports_dir, "SIG16/analysis_outs/motif_info.rds"))
pal <- readRDS(file.path(imports_dir, "SIG16/analysis_outs/plots/pal_family.rds"))

dim(motif.mtx.peak.final)
```

```{r}
read_single <- function(term) {
  readRDS(file.path(DESEQ_DIR, paste0("name-", term), "unshrunken-res.table.rds"))
}
# interaction-classification.csv carries the SIG13 schema: `name` + `interaction_class`
read_itxn <- function(l1, l2 = "TNF") {
  df <- read.csv(file.path(DESEQ_DIR, paste0("name-ligand1", l1, ".ligand2", l2),
                           "interaction-classification.csv"))
  rownames(df) <- df$name
  df
}

singles <- list(TGFb = read_single("ligand1_TGFb_vs_none"),
                IL4  = read_single("ligand1_IL4_vs_none"),
                IL6  = read_single("ligand1_IL6_vs_none"),
                TNF  = read_single("ligand2_TNF_vs_none"))

itxns <- list(TGFb.TNF = read_itxn("TGFb"),
              IL4.TNF  = read_itxn("IL4"),
              IL6.TNF  = read_itxn("IL6"))
```

# Define region sets

```{r}
split_by_lfc <- function(df, lfc.col, padj.col) {
  keep <- !is.na(df[[padj.col]]) & df[[padj.col]] < padj.threshold
  list(pos = df[keep & df[[lfc.col]] >  lfc.threshold, ],
       neg = df[keep & df[[lfc.col]] < -lfc.threshold, ])
}

itxn.classes <- c("synergy positive", "synergy negative", "buffering")
set_name <- function(prefix, cl) paste0(prefix, ".", gsub(" ", "_", cl))

regions <- list()
for (nm in names(singles)) {
  s <- split_by_lfc(singles[[nm]], "log2FoldChange", "padj")
  regions[[paste0(nm, ".gained")]] <- s$pos
  regions[[paste0(nm, ".lost")]]   <- s$neg
}
for (nm in names(itxns)) {
  for (cl in itxn.classes) {
    peaks <- itxns[[nm]][!is.na(itxns[[nm]]$interaction_class) &
                           itxns[[nm]]$interaction_class == cl, ]
    rownames(peaks) <- peaks$name   # the enrichment test matches peaks on rownames
    regions[[set_name(nm, cl)]] <- peaks
  }
}

tibble(region = names(regions), n_peaks = map_int(regions, nrow))
```

# Hypergeometric enrichment

Several interaction classes are legitimately tiny (the TGFb x TNF and IL6 x TNF interaction
terms yield only a handful of significant peaks), so empty sets are skipped rather than fed to
`phyper` with a zero denominator.

```{r}
# One-sided hypergeometric test per motif: of the `all_peaks` in the atlas,
# `matches_in_all_peaks` carry the motif; ask whether the `DARs` in this region set
# carry it more often than that background rate.
hypergeom.res.list <- compact(imap(regions, function(input.regions, region.name) {
  if (nrow(input.regions) == 0) return(NULL)

  peak.pos <- match(rownames(input.regions), rownames(motif.mtx.peak.final))
  n.all <- nrow(motif.mtx.peak.final)
  n.dar <- length(peak.pos)

  res <- imap_dfr(motif.mtx.peak.final, function(hits, motif) {
    hits.all  <- sum(hits)
    hits.dar  <- sum(hits[peak.pos])
    tibble(motif = motif,
           pval = phyper(hits.dar - 1, hits.all, n.all - hits.all, n.dar,
                         lower.tail = FALSE),
           fold_enrichment = (hits.dar / n.dar) / (hits.all / n.all),
           matches_in_DARs = hits.dar, DARs = n.dar,
           matches_in_all_peaks = hits.all, all_peaks = n.all)
  }) %>%
    arrange(pval) %>%
    mutate(p.adjust = p.adjust(pval, method = "BH"), .after = pval) %>%
    relocate(motif, .after = all_peaks) %>%
    as.data.frame()
  rownames(res) <- res$motif

  res$family <- motif.info[match(rownames(res), motif.info$TF), "family"]
  write.csv(res, file.path(OUT_DIR, paste0(region.name, ".csv")), quote = FALSE)
  res
}))
```

```{r}
padj.cutoff <- 0.05
fe.cutoff   <- 1.5
k <- 10

map_dfr(hypergeom.res.list, ~ tibble(
  n_motifs_tested = nrow(.x),
  n_significant = sum(.x$p.adjust < padj.cutoff & .x$fold_enrichment > fe.cutoff, na.rm = TRUE)
), .id = "region") %>%
  left_join(tibble(region = names(regions), n_peaks = map_int(regions, nrow)), by = "region") %>%
  relocate(n_peaks, .after = region)
```

# Top-motif bar plots

Top `r k` motifs by fold enrichment, among motifs passing `p.adjust < 0.05` and
fold enrichment > 1.5, coloured by TF structural family.

```{r}
top_motifs <- function(res, k = 10) {
  sig.tf <- res[(res$p.adjust < padj.cutoff) & (res$fold_enrichment > fe.cutoff), ]
  sig.tf <- sig.tf %>% arrange(desc(fold_enrichment))
  sig.tf <- sig.tf[seq_len(min(k, nrow(sig.tf))), ]
  sig.tf[!is.na(sig.tf$pval), ]
}

plot_bar <- function(sig.tf, region.name) {
  ggplot(sig.tf) +
    geom_col(aes(x = fold_enrichment, y = reorder(motif, fold_enrichment), fill = family)) +
    theme(axis.text.y = element_text(size = 12), axis.title.y = element_blank()) +
    labs(title = region.name, x = "fold enrichment") +
    scale_fill_manual(values = pal$family) +
    theme(aspect.ratio = 2)
}
```

```{r, results='asis', fig.width=7, fig.height=4, message=FALSE, warning=FALSE}
bar.plots <- list()
for (region.name in names(hypergeom.res.list)) {
  sig.tf <- top_motifs(hypergeom.res.list[[region.name]], k)
  cat("\n## ", region.name, " (", nrow(sig.tf), " motifs shown)\n", sep = "")
  if (nrow(sig.tf) == 0) { cat("No motifs pass the enrichment cutoffs.\n"); next }
  bar.plots[[region.name]] <- plot_bar(sig.tf, region.name)
}
```

All panels are assembled into a single multi-panel PDF with `patchwork` rather than one file per
region set; the fill legend is collected once for the whole figure.

```{r, fig.height=10, fig.width=20}
n.col <- 4
n.row <- ceiling(length(bar.plots) / n.col)
wrap_plots(bar.plots, ncol = n.col) +
  plot_layout() &
  theme(legend.position = "right")

ggsave(file.path(PLOT_DIR, paste0("barplots-top", k, "-motifs-padj-", padj.cutoff, ".pdf")))
```

# Bi-directional bar plots

Two opposing region sets on one axis; bar length is fold enrichment in either direction. For the
single ligands the pair is gained (right) vs lost (left); for the interactions it is
`synergy positive` (right) vs `synergy negative` (left). The `buffering` sets have no natural
opposite and appear only in the uni-directional plots above.

```{r}
bidirectional_pairs <- tibble::tribble(
  ~label,     ~pos,                        ~neg,                        ~pos.lab,           ~neg.lab,
  "TGFb",     "TGFb.gained",               "TGFb.lost",                 "gained",           "lost",
  "IL4",      "IL4.gained",                "IL4.lost",                  "gained",           "lost",
  "IL6",      "IL6.gained",                "IL6.lost",                  "gained",           "lost",
  "TNF",      "TNF.gained",                "TNF.lost",                  "gained",           "lost",
  "TGFb.TNF", "TGFb.TNF.synergy_positive", "TGFb.TNF.synergy_negative", "synergy positive", "synergy negative",
  "IL4.TNF",  "IL4.TNF.synergy_positive",  "IL4.TNF.synergy_negative",  "synergy positive", "synergy negative",
  "IL6.TNF",  "IL6.TNF.synergy_positive",  "IL6.TNF.synergy_negative",  "synergy positive", "synergy negative")

plot_bidirectional <- function(pair) {
  sig.tf.pos <- top_motifs(hypergeom.res.list[[pair$pos]], k)
  sig.tf.neg <- top_motifs(hypergeom.res.list[[pair$neg]], k)
  if (!is.null(sig.tf.neg)) sig.tf.neg$fold_enrichment <- sig.tf.neg$fold_enrichment * -1

  sig.tf <- rbind(sig.tf.pos, sig.tf.neg)
  if (is.null(sig.tf) || nrow(sig.tf) == 0) return(NULL)

  sig.tf <- sig.tf %>% arrange(fold_enrichment)
  sig.tf$rank <- factor(seq_len(nrow(sig.tf)), levels = seq_len(nrow(sig.tf)))

  annotations <- data.frame(
    xpos = c(-Inf, Inf),
    ypos = c(-Inf, -Inf),
    annotateText = c(pair$neg.lab, pair$pos.lab),
    hjustvar = c(0, 1.1),
    vjustvar = c(-0.5, -0.5))

  bar.p <- ggplot(sig.tf) +
    geom_col(aes(x = fold_enrichment, y = rank, fill = family)) +
    theme(axis.text.y = element_text(size = 12), axis.title.y = element_blank()) +
    #scale_x_continuous(breaks = seq(-10, 10, 1), labels = abs(seq(-10, 10, 1))) +
    # extra room at the bottom so the direction labels clear the last bar
    scale_y_discrete(labels = sig.tf$motif, expand = expansion(add = c(1.2, 0.6))) +
    scale_fill_manual(values = pal$family) +
    labs(title = pair$label, x = "fold enrichment") +
    geom_text(data = annotations,
              aes(x = xpos, y = ypos, hjust = hjustvar, vjust = vjustvar,
                  label = annotateText)) +
    theme(aspect.ratio = 2)
  bar.p
}
```

```{r, results='asis', fig.width=7, fig.height=6, message=FALSE, warning=FALSE}
bidir.plots <- list()
for (i in seq_len(nrow(bidirectional_pairs))) {
  pair <- as.list(bidirectional_pairs[i, ])
  cat("\n\n## ", pair$label, "\n\n", sep = "")
  if (!all(c(pair$pos, pair$neg) %in% names(hypergeom.res.list))) {
    cat("One or both region sets are empty; no plot.\n\n"); next
  }
  p <- plot_bidirectional(pair)
  if (is.null(p)) {
    cat("No motifs pass the enrichment cutoffs.\n\n")
  } else {
    bidir.plots[[pair$label]] <- p
    print(p)
  }
}
```

```{r}
n.col <- 4
n.row <- ceiling(length(bidir.plots) / n.col)
bidir.panel <- wrap_plots(bidir.plots, ncol = n.col) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

ggsave(file.path(PLOT_DIR,
                 paste0("barplots-bidirectional-top", k, "-motifs-padj-", padj.cutoff, ".pdf")),
       bidir.panel, width = 5 * n.col, height = 6 * n.row, limitsize = FALSE)
```
